### 2015-03-02 draft.tex
up to date from abc_sgld

\documentclass[]{article}
\usepackage{proceed2e}

% Set the typeface to Times Roman
\usepackage{times}

% For figures
\usepackage{graphicx} % more modern
%\usepackage{epsfig} % less modern
%\usepackage{subcaption} 
%usepackage[numers]{natbib}
% For citations
%\usepackage[numbers]{natbib}

% For algorithms
%\usepackage{algorithm}
%\usepackage{algorithmic}

% As of 2011, we use the hyperref package to produce hyperlinks in the
% resulting PDF.  If this breaks your system, please commend out the
% following usepackage line and replace \usepackage{icml2014} with
% \usepackage[nohyperref]{icml2014} above.
%\usepackage{hyperref}
% Packages hyperref and algorithmic misbehave sometimes.  We can fix
% this with the following command.
%\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Employ the following version of the ``usepackage'' statement for
% submitting the draft version of the paper for review.  This will set
% the note in the first column to ``Under review.  Do not distribute.''
\usepackage{bm}
\usepackage{amsmath}
\usepackage[numbers]{natbib}
%\usepackage{array}
%\usepackage{nips_style} 
%\usepackage{sidecap}
%\usepackage[export]{adjustbox}
%\include{cdefs}

% For algorithms
\usepackage{algorithm}
\usepackage{algorithmic}

% As of 2011, we use the hyperref package to produce hyperlinks in the
% resulting PDF.  If this breaks your system, please commend out the
% following usepackage line and replace \usepackage{icml2014} with
% \usepackage[nohyperref]{icml2014} above.
%\usepackage{hyperref}
% Packages hyperref and algorithmic misbehave sometimes.  We can fix
% this with the following command.
%\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Leave date blank
\date{}

\pagestyle{myheadings}
%\usepackage{sidecap}
%\usepackage[export]{adjustbox}

%\newcommand{\theHalgorithm}{\arabic{algorithm}}

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}

\input{cdefs}

\title{Hamiltonian ABC}

\author{} % LEAVE BLANK FOR ORIGINAL SUBMISSION.
          % UAI  reviewing is double-blind.

% \author{
% Edward Meeds \\
% Informatics Institute \\
% University of Amsterdam \\
% \texttt{tmeeds@gmail.com} \\
% \and
% \author{
% Robert Leenders \\
% Informatics Institute \\
% University of Amsterdam \\
% \texttt{robert.leenders@student.uva.nl} \\
% \and
% Max Welling \\
% Informatics Institute\\
% University of Amsterdam \\
%  \texttt{welling.max@gmail.com}
% }
%\nipsfinalcopy
\begin{document} 
	\vskip -0.3in

\maketitle

% 
% 
% \begin{abstract} 
% \input{abstract}
% \end{abstract} 
\begin{abstract} 
  Approximate Bayesian computation (ABC) is a powerful and elegant framework for performing inference in simulation-based models.  However, due to the difficulty in scaling likelihood estimates, ABC remains useful for relatively low-dimensional problems. We introduce Hamiltonian ABC (HABC), a set of likelihood-free algorithms that apply recent advances in scaling Bayesian learning using Hamiltonian Monte Carlo (HMC) and stochastic gradients.     We find that a small number forward simulations can effectively approximate the ABC gradient, allowing Hamiltonian dynamics to efficiently traverse parameter spaces.  We also describe a new simple yet general approach of incorporating random seeds into the state of the Markov chain, further reducing the random walk behavior of HABC.  We demonstrate HABC on several typical ABC problems, and show that HABC performs comparably to regular Bayesian inference on a high-dimensional problem from machine learning.
  %
  %
  % Models in simulation-based science domains are considered likelihood-free (LF) since they are not defined by a probabilistic model, but by a simulator that maps parameters to pseudo-observations.  To perform parameter inference in the models, a powerful set of statistical procedures called Approximate Bayesian Computation (ABC) can be applied.  Although there is a wide variety of ABC algorithms, they are dominated by SMC/PMC and MCMC.  These have been successful, but their inefficiency of producing samples from the approximate posterior has limited ABC to simulations with relatively few parameters.  Hamiltonian Monte Carlo (HMC)  provides a framework for extending ABC to high-dimensions, but requires computing the gradient of the log-likelihood.  With Hamiltonian ABC, we show how recent advances of stochastic gradient HMC algorithms applied to the big data domain can be also used for efficient sampling for LF models.  We find that a small number forward simulations can effectively approximate the true ABC gradient, allowing Hamiltonian dynamics to efficiently traverse parameter spaces.   We demonstrate HABC on several typical ABC problems, and show that HABC performs comparably to regular Bayesian inference on a high-dimensional problem from machine learning.
%
%
%   Bayesian inference in simulation-based science relies
%   Approximate Bayesian Computation (ABC) is a set of inference procedures
%   for so-called {\em likelihood-free} models
%
% In this paper we apply recent techniques from Bayesian inference that use gradient information to sample efficiently from the true posterior distribution to the {\em likelihood-free} or {\em approximate Bayesian computation} (ABC) setting.  To do this for ABC we adopt a {\em gradient-free} stochastic approximation algorithm by Spall~\cite{spall1999}.  Together these algorithms provide both optimization and inference for likelihood-free models as the algorithm ABC-SGLD transitions from optimization to sampling.  We demonstrate ABC-SGLD on problems where the true gradient information is known and on challenging ABC simulators.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{INTRODUCTION} \label{introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In simulation-based science, models are defined by a simulator and its parameters.  These are called {\em likelihood-free} models because, in contrast to probabilistic models, their likelihoods are either intractable to compute or must be approximated by simulations.  To perform inference in likelihood-free models, a broad class of algorithms called Approximate Bayesian Computation \cite{beaumont2002approximate,marjoram2003markov,sisson2007sequential,sisson:2010,marin:2012,fan:2013} are employed.

At the core of every ABC algorithm is simulation.  To evaluate the quality of a parameter vector $\thetav$, a simulation is run using $\thetav$ as inputs and producing outputs $\x$.  If the pseudo-data $\x$ is ``close'' to observations $\y$, then $\thetav$ is kept as a sample from the approximate posterior.  Parameters $thetav$ are then adjusted, depending upon the algorithm, to obtain the next sample.

In ABC, there is a fundamental trade-off between the computation required to obtain independent samples and the approximation to the true posterior.  If the parameter measuring closeness is too small, then samplers ``mix'' poorly; on the other hand, if it is too large, then the approximation is poor.  As the dimension of the parameters grows, the problem worsens, just as it does for general Bayesian inference with probabilistic models, but it is more acute for ABC due to its simulation requirement.  There is therefore a deep interest in improving the efficiency of ABC samplers (in terms of computation per independent sample).  In this paper we address this issue directly by using Hamiltonian dynamics to approximately sample from likelihood-free models with high-dimensional parameters.

Hamiltonian Monte Carlo (HMC) \cite{duane1987hybrid, neal2011mcmc} is perhaps the only Bayesian inference algorithm that scales to high-dimensional parameter spaces.  The core computation of HMC is the gradient of the log-likelihood.  Two problems arise if we consider HMC for ABC: one, how can the gradients be computed for high-dimensional likelihood-free models, and two, given a stochastic approximation to the gradient, can a valid HMC algorithm be derived?

 To answer the latter, we turn to recent developments in scaling Bayesian inference using HMC and stochastic gradients \cite{welling2011bayesian,chen2014stochastic,ding2014bayesian}.  We call these {\em stochastic gradient Hamiltonian dynamics} (SGHD) algorithms. SGHD are computationally efficient for two reasons.  First, they avoid computing the gradient of the log-likelihood over the entire data set, instead approximating it using small batches of data, i.e. computing stochastic gradients.  Second, they can maintain reasonable approximations to the Hamiltonian dynamics and therefore avoid a Metropolis-Hastings correction step involving the full data set.  Different strategies are employed to do this: small step-sizes combined with  Langevin dynamics \cite{welling2011bayesian}, using friction to prevent accumulation of errors in the Hamiltonian \cite{chen2014stochastic}, and using a thermostat to control the temperature of the Hamiltonian \cite{ding2014bayesian}.  Each of these strategies can be used by HABC.

%Recent advances in Bayesian inference within the {\em big data} domain have combined the efficiency of stochastic gradient algorithms used for optimization with Hamiltonian dynamics;  these include Stochastic Gradient Langevin Dynamics (SGLD) \cite{welling2011bayesian}, Stochastic Gradient HMC \cite{chen2014stochastic}, and Stochastic Gradient Thermostats \cite{ding2014bayesian}.
%By using small batches of the full data set these methods are able to quickly compute a stochastic approximation to the true gradient, which with a small enough step-size and with an appropriate amount of injected noise, can produce samples from the true posterior.  The Hamiltonian dynamics avoid random walk behavior by providing momentum to the sampler allowing it to reach far away regions of parameter space.  Critically for the big data domain, the small step-sizes prevent the trajectories of the sampler from accumulating too much error and therefore they can avoid a Metropolis-Hastings correction step involving the full data set.

In HABC, we use forward simulations to approximate the likelihood-free gradient. The key difference between SGHD methods and HABC is that the stochasticity of the gradient does not come from approximating the full data gradient with a batch gradient, but by the stochasticity of the simulator.  It is therefore not the expense of the simulator (though this could very well be the case for many interesting simulation-based models -- see Section~\ref{sec:conclusion}) that requires an approximation to the gradient, but the likelihood-free nature of the problem.

There are several difficulties in estimating gradients of likelihood-free models that we address with HABC.  The first is do to the form of the ABC log-likelihood.  As we show in Section~\ref{sec:abc}, using a conditional model for $\pi( \x | \thetav )$ provides an estimate of the ABC likelihood that is less sensitive to $\eps$ and therefore is more conducive to stochastic gradient computations.  The second difficulty is that for high-dimensional parameter spaces, computing the gradients naively (i.e. by finite differences \cite{kiefer1952stochastic}) can squash the gains brought by the Hamiltonian dynamics.  Fortunately, we can use existing stochastic approximation algorithms \cite{spall1992multivariate,spall2000adaptive} that can be used to compute unbiased estimators of the gradient with a small number of forward simulations that is {\em independent} of the parameter dimension.  The {\em stochastic perturbation stochastic approximation} (SPSA) \cite{spall1992multivariate} is described in Section~\ref{sec:habc}

%Approximate Bayesian Computation (ABC) is a set of inference procedures for so-called {\em likelihood-free} models \cite{abccitations}.  In contrast to typical Bayesian modeling, where the parameters of a probabilistic model of data are the focus, the primary interest in ABC are the parameters of a simulation model.  The simulator can generate pseudo or auxiliary data by running the simulator with a set of parameter values, but it cannot compute a likelihood directly.  Inference involves running simulations, comparing pseudo data with observations, and using a distance between the two as an approximate likelihood.  Markov chain Monte Carlo (MCMC) or Sequential Monte Carlo (SMC) are the most common inference algorithms.  It can be shown that if the statistics used in the likelihood function are sufficient, then these algorithms sample correctly from an approximation to the true posterior.

%The approximation to the true posterior can be made arbitrarily small, in principle, by controlling the kernel width parameter $\epsilon$.  As $\epsilon \rightarrow 0$, however, simulated data must be closer and closer to the observations (or receive zero likelihood).  The acceptance rate of a Markov chain for small $\epsilon$ can therefore be low.  This inefficiency of ABC sampling is exacerbated by the random walk proposals.  As the dimensionality of the parameter space grows,  $\epsilon$ must grow to prevent low acceptance rates.  This has perhaps prevented application of ABC to high-dimensional simulators.  Our paper addresses this directly: can we use the sampling efficiency of Hamiltonian dynamics (i.e. high acceptance rates) in high-dimensional simulators?

%In this paper we adopt the developments in scaling Bayesian inference to ABC.  To do this we use forward simulations to approximate the likelihood-free gradient.  The key difference with SG methods mentioned above is that the stochasticity of the gradient does not come from approximating the full data gradient with a batch gradient, but by the stochasticity of the simulator.  It is therefore not the expense of the simulator (though this could very well be the case for many interesting simulation-based models, and we address this in Future Work) that requires an approximation to the gradient, but the likelihood-free nature of the problem.  For high-dimensional parameter spaces, computing the gradients naively (i.e. by finite differences) can squash the gains brought by the Hamiltonian dynamics; to address this further approximations to the gradient can be used \cite{spall}.

%The approximate gradients used by HABC can be noisy and difficult to manage is care is not taken.  To address this we show how a local Gaussian approximation of the simulator output can improve the stability of the gradient estimates over a direct estimate using the ABC kernel function.

A further innovation of this paper is the use of common random numbers (CRN) to improve the efficiency of the Hamiltonian dynamics.  The idea behind CRNs is to use the same set of random seeds for estimating a gradient by FD or SPSA, i.e. when simulating $\pi(\x| \theta + d\theta)$ and $\pi(\x| \theta - d\theta)$ use the same random seeds.  This was applied successfully to SPSA \cite{kleinman1999simulation} (and is analogous to using the same mini-batch in stochastic gradient methods).  We extend and simplify this approach by including the random seeds $\omega$ into the state of the Markov chain;  by keeping the random seeds fixed for several consecutive steps, the second order gradient stochasticity is greatly reduced.  We show that this doing this produces is a valid MCMC procedure.  This approach is not exclusive to HABC; we apply successfully to ABC-MCMC as well.

We briefly review ABC in Section~\ref{sec:abc}.  In Section~\ref{sec:scaling} we review three approaches to stochastic gradient inference using Hamiltonian dynamics: SGLD, SGHMC, and SGNHT.  We then introduce Hamiltonian ABC in Section~\ref{sec:habc}, where  we will show how to improve the stability of the gradient estimates by using CRNs and local density estimators of the simulator.  Extensions to high-dimensional parameter spaces are also discussed.  In Section~\ref{sec:demo} we show how HABC behaves on a simple one-dimensional problem, then in Section~\ref{sec:experiments} we compare HABC with ABC-MCMC for two problems: a low-dimensional model of chaotic population dynamics and a high-dimensional problem.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{APPROXIMATE BAYESIAN COMPUTATION}\label{sec:abc}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Consider the Bayesian inference task of either drawing samples from or learning an approximate model of the following (usually intractable) posterior distribution:
\begin{equation}
  \pi(\thetav | \y_1, \ldots, \y_N ) \propto \pi(\thetav) \pi( \y_1, \ldots, \y_N  | \thetav )
\end{equation}
where $\pi(\thetav)$ is a prior distribution over parameters $\thetav \in {\rm I\!R}^{D}$ and $\pi( \y_1, \ldots, \y_N  | \thetav )$ is the likelihood of $N$ data observations, where $\y_i \in {\rm I\!R}^J$.  In ABC, the vector of $J$ observations are typically informative statistics of the raw observations.  It can be shown that if the statistics used in the likelihood function are sufficient, then these algorithms sample correctly from an approximation to the true posterior ([KEEP?]).  
  The simulator is treated as generator of random pseudo-observations, i.e. $\x \simsim \pi( \x | \thetav )$ is a draw from the simulator.  Discrepancies between the simulator outputs $\x$ and the observations $\y$ are scaled by a closeness parameter $\eps$ and treated as likelihoods.  This is the equivalent to putting an $\eps$-kernel around the observations, and using a Monte Carlo estimate of the likelihood using $S$ draws of $\x$: 
\begin{equation}
  \pi_{\epsvec}( \y | \thetav ) =  \int \pi_{\epsvec}(\y | \x ) \pi( \x | \thetav ) d\x 
                           \approx  \frac{1}{S} \sum_{s=1}^S \pi_{\epsvec}(\y | \x^{(s)} ) \label{eq:abc_mc_approx}
\end{equation}

In ABC Markov chain Monte Carlo (MCMC) \cite{marjoram2003markov,Wilkinson2013,sisson:2010} the Metropolis-Hastings (MH) proposal distribution is composed of the product of the proposal for the parameters $\thetav$ and the proposal for the simulator outputs:
\begin{equation}
  q( \thetavp, \x^{(1)'}, \ldots, \x^{(S)'} | \thetav ) =  q( \thetavp | \thetav ) \prod_s \pi( \x^{(s)'} | \thetavp)
\end{equation}
Using this form of the proposal distribution, and using the Monte Carlo approximation eq~\ref{eq:abc_mc_approx}, we arrive at the following Metropolis-Hastings accept-reject probability,
%
\begin{equation}
\alpha = \min \lp 1, \frac{\pi\lp\thetavp\rp \sum_{s=1}^S \pi_{\epsvec}(\y | \xps )  q( \thetav | \thetavp )}{\pi\lp\thetav\rp \sum_{s=1}^S \pi_{\epsvec}(\y | \xs ) q( \thetavp | \thetav )} \rp \label{eq:abc_mh_acceptance_with_s}
\end{equation}
%
If the simulations are part of the Markov chain, the algorithm corresponds to the pseudo-marginal (PM) sampler \cite{andrieu2009pseudo}, otherwise it is a marginal sampler \cite{marjoram2003markov,sisson:2010}.   For this paper we will be interested in the PM sampler because this is equivalent to having the random states that generated the simulation outputs in the state of the Markov chain, which we will use within a valid ABC sampling algorithm in Section~\ref{sec:habc}.
%
%   If they are discarded, the
% Two subtly different versions of ABC-MCMC are the pseudo-marginal (PM) \cite{delmoral2008,andrieu2009pseudo} and marginal sampler \cite{marjoram2003markov,sisson:2010}.  In PM, the simulations are part of the state of the Markov chain whereas in the marginal sampler they are not
% When only the numerator is re-estimated at every iteration (and the denominator is carried over from the previous iteration), then this algorithm corresponds to pseudo-marginal (PM) sampling \cite{delmoral2008,andrieu2009pseudo}. PM sampling is asymptotically correct (taking for granted the approximation introduced by the kernel $\pi_{\epsvec}$) but can display very poor mixing properties. By resampling the denominator as well, we improve mixing at the cost of introducing a further approximation. This sampler is known as the marginal sampler \cite{marjoram2003markov,sisson:2010}.
% [TODO: END TAKEN FROM POPE]

An alternative approach to computing the ABC likelihood is to estimate the parameters of a conditional model  $\pi(\x|\thetav)$, e.g. kernel density estimate \cite{TurnerGenLik2014} or a Gaussian model \cite{wood2010statistical}.  While either approach should be adequate and both have their own limits and advantages, for this paper we will use a Gaussian model.  In ABC, using a conditional Gaussian model for  $\pi(\x|\thetav)$ is called a {\em synthetic likelihood} (SL) model \cite{wood2010statistical}.  For a SL log-likelihood model, we compute estimators of the first and second moments of $\pi(\x|\thetav)$.  The advantage is that for a Gaussian $\epsilon$-kernel, we can convolve the two densities   
\begin{eqnarray}
  \pi_{\epsvec}( \y | \thetav ) & = & \int \mathcal{N}( \y | \x, \eps^2 ) \mathcal{N}( \x | 
  \mu_{\thetav}, \sigma^2_{\thetav} ) d\x \\
                          & = & \mathcal{N}( \y | \mu_{\thetav}, \sigma^2_{\thetav} + \eps^2 )
\end{eqnarray}

Of particular concern to this paper is the behavior of the log-likelihoods for different values of $\epsilon$.  In the $\eps$-kernel case, the log-likelihood is very sensitive to small values of $\eps$:
\begin{eqnarray}
  \log \pi_{\epsvec}( \y | \thetav ) & = & \log \sum_s \mathcal{N}( \y | \x^{(s)}, \eps^2 ) \\
                          & = & \log \mathcal{N}( \y | \x^{(s)}, \eps^2 ) + \log\lp 1 + H \rp \\
                          & \approx & -\log \eps -\frac{1}{2 \eps^2} ( \y - \x^{(m)} )^2 
\end{eqnarray}
where $m$ is the simulation that is closest to $\y$, $H$ is a sum over terms close to $0$. We can see that the log-likelihood can be set arbitrarily small by decreasing $\eps$.  On the other hand, by using a model of the simulation at $\thetav$
\begin{eqnarray}
 \log \pi_{\epsvec}( \y | \thetav ) & \approx & -\frac{1}{2} \log (\sigma^2_{\thetav} + \eps^2 ) -\frac{( \y - \mu_{\thetav} )^2 }{2 (\sigma^2_{\thetav} + \eps^2 )} 
\end{eqnarray}
 For the SL model, $\eps$ acts as a smoothing term and can be set to small values with little change to the log-likelihood, as long as the SL estimators are fit appropriately.  This insensitivity to $\eps$ will be used in Section~\ref{sec:habc} for estimating gradients of the ABC likelihood.  Before describing HABC in full detail however, we now explain how scaling Hamiltonian dynamics in Bayesian learning can be accomplished using stochastic gradients from batched data.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{SCALING BAYESIAN INFERENCE USING HAMILTONIAN DYNAMICS} \label{sec:scaling}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Scaling Bayesian inference algorithms to massive datasets is necessary for their existing relevance in the so-called {\em big data} era.  We now review the role stochastic gradient methods combined with Hamiltonian dynamics have played in recent advances in scaling Bayesian inference.   Most importantly, these methods have combined the ability of HMC to explore high-dimensional parameter spaces with the computational efficiency of using stochastic gradients based on small batches of the full dataset.  After an overview of HMC, we will briefly describe stochastic gradient Hamiltonian dynamics (SGHDs), starting with using  Langevin dynamics \cite{welling2011bayesian}, then HMC with friction \cite{chen2014stochastic}, and finally HMC with thermostats \cite{ding2014bayesian}.  We will then make the connection between SGHDs and HABC in Section~\ref{sec:habc}.

\subsection{Hamiltonian Monte Carlo}\label{sec:hmc}

Hamiltonian dynamics are often necessary to adequately explore the target distribution of high-dimensional parameter spaces.  By proposing parameters that are far from the current location and yet have high acceptance probability, Hamiltonian Monte Carlo \cite{duane1987hybrid, neal2011mcmc}  can efficiently avoid random walk behavior that can render proposals in high-dimensions painfully slow to mix.

HMC simulates the trajectory of a particle along a frictionless surface, using random initial momentum $\rhov$ and position $\thetav$.  The Hamiltonian function computes the energy of the system and the dynamics govern how the momentum and position change over time.  The continuous Hamiltonian dynamics can be simulated by discretizing time into small steps $\eta$.  If $\eta$ is small, the value of $\thetav$ at the end of a simulation can be used as proposals within the Metropolis-Hastings algorithm.  Hamiltonian dynamics should propose $\thetav$ that are always accepted, but errors due to discretization may require a  Metropolis-Hastings correction.  It is this correction step that SGHD algorithms want to avoid as it requires computing the log-likelihood over the full data set.

More formally, the Hamiltonian $H\lp\thetav, \rhov \rp = U(\thetav) + K(\rhov)$ is a function of the current potential energy $U(\thetav)$ and kinetic energy $K(\rhov) = \rhov^T M^{-1}\rhov/2$ ($M$ is a diagonal matrix of masses which for presentation are set to $1$).  The potential energy is defined by negative log joint density of the data and prior:
\begin{equation}
  U(\thetav) = - \log \pi(\thetav) - \sum_{i=1}^N \log \pi(\yi | \thetav )
\end{equation}
The Hamiltonian dynamics follow 
\begin{equation}
  d\thetav = \rhov dt ~~~~~~~~~~~~ d\rhov = -\nabla U(\thetav) dt
\end{equation}
in simulation $dt = \eta$. 
% Actual simulation of the Hamiltonian requires taking discrete steps, and therefore introduces some error.  The error can be corrected using a MH step, as shown in Algorithm 1 (showing leap-frog algorithm).  By only taking a single leap-frog step, HMC reduces to Langevin dynamics which can be represented as a gradient descent algorithm with injected noise.  By using small step-sizes the acceptance rate approaches 1, so that no MH correction steps are necessary and the algorithm can be run to generate a stream of (possibly highly correlated) samples from the posterior.

% Two problems must be addressed.  The first is loglikelihood computation.  For correctness, Bayesian inference algorithms require frequent computation of the (deterministic) loglikelihood over the entire observation set.  Contrast this to optimization, where stochastic gradient techniques are necessary for learning parameters.  The second problem is that massive parameter sets is synonymous with massive datasets.  As datasets grow, so too do our models.  This is particularly true in the area of deep learning, where parameter vectors start in the hundred thousands and for some models are in the billions. Hamiltonian Monte Carlo addresses this for Bayesian inference: by using gradients of the log-likelihood (log-prior) it is possible to avoid random walk behavior and more efficiently explore the posterior.

% Recent advances in scaling Bayesian inference to big data has shown that both batches and gradients can be used to correctly sample from the posterior distribution.  These use Hamiltonian dynamics based on stochastic gradients (where the stochasticity comes from approximating the true gradient with one based on a batch of data) with appropriate step-sizes (and variations, see below).  We now describe these algorithms and then show how we can adopt them for ABC.

\subsection{Stochastic Gradient Hamiltonian Dynamics}\label{se:sghd}
If the log-likelihood over the full data set is replaced with a batch estimate, as is done for the following {\em stochastic gradient Hamiltonian dynamics} (SGHDs) algorithms, then the error in simulating the Hamiltonian dynamics comes not only from the discretization, but from the variance of the stochastic gradient.  As long as this error is controlled, either by using small steps $\eta$ (SGLD), or adding friction terms $A$ (SGHMC), or using a thermostat $\xi$ (SGNHT), the expensive MH correction step can be avoided and values of $\thetav$ from the Hamiltonian dynamics can be used as samples from the posterior.

SGHDs replace the full potential energy and its gradient with a batch approximation:
\begin{eqnarray}
  \hat{U}(\thetav)       & = & - \log \pi(\thetav) - \frac{N}{n} \sum_{i=h_1}^{h_n} \log \pi(\yi | \thetav ) \\
  \nabla\hat{U}(\thetav) & = & - \nabla \log \pi(\thetav) - \frac{N}{n} \sum_{i=h_1}^{h_n} \nabla \log \pi(\yi | \thetav ) 
\end{eqnarray} 
where $n$ is the batch size, and $h_i$ are indices chosen randomly without replacement from $[1,N]$ (i.e. it defined a random batch).  
%The main motivation for these is to avoid the expensive MH correction step while maximizing the benefit of Hamiltonian dynamics.  For example, we want to take large steps but without introducing errors that cause the dynamics to fail.

{\bf Stochastic gradient Langevin dynamics} \cite{welling2011bayesian} performs one full leap-frog step of HMC.   Starting with a half step for the momentum, the update for $\thetav$ is 
\begin{eqnarray}
  \rhov_t & \sim & \mathcal{N}(0,\eye_p) \\
  \rhov_{t+\frac{1}{2}} & = & \rhov_t - \eta \nabla \hat{U}(\thetav_t) /2 \\
  \thetav_{t+1} & = & \thetav_t + \eta \rhov_{t+\frac{1}{2}}
\end{eqnarray}
Tt is not necessary to include $\rhov$ in the updates since there is only one step:
\begin{eqnarray}
  \thetav_{t+1} & = & \thetav_t + \eta \mathcal{N}(0,\eye_p) - \eta^2 \nabla \hat{U}(\thetav_t) /2 
\end{eqnarray}
One of the potential drawbacks of SGLD is that the momentum term is {\em refreshed} for every update of the parameters, and since this means the parameter update only uses the current gradient approximation, it limits the benefits of using Hamiltonian dynamics.  On the other hand, this also prevents SGLD from accumulating errors in the Hamiltonian dynamics.

{\bf Stochastic Gradient HMC} (SGHMC) \cite{chen2014stochastic} avoids $\rhov$ refreshment altogether.  By applying HMC directly using the stochastic approximation $\hat{U}$ and $\nabla \hat{U}$, which the authors call {\em naive SGHMC}, the variance of the gradient will introduce errors that left unaddressed will result in sampling from the incorrect target distribution.  Under the assumption that $\nabla \hat{U}(\thetav) = \nabla U(\thetav) + \mathcal{N}\lp \zerov, \varv_{\thetav}\rp$, where $\varv_{\thetav}$ is the covariance of the gradient approximation, and  updates $\rhov_{t+1} = \rhov_{t} + \Delta \rhov_{t}$ and $\thetav_{t+1} = \thetav_t + \eta \rhov_{t+1}$, the change in momenta $\Delta \rhov$ from one full step is
 \begin{equation}
   - \eta \lp \nabla U(\thetav) + \mathcal{N}\lp \zerov, \varv_{\thetav}\rp\rp  =  - \eta \nabla U(\thetav) + \mathcal{N}\lp \zerov, \eta^2 \varv_{\thetav}\rp  
 \end{equation}
By adding a friction term $\Bv$ to $\Delta \rhov$ proportional to $\varv_{\thetav}$, the correction step can be avoided
 \begin{eqnarray}
   \Delta \rhov & = & - \eta \Bv \rhov_t - \eta \nabla U(\thetav_t) + \mathcal{N}\lp \zerov, 2 \eta \Bv\rp 
 \end{eqnarray}
 where  $\Bv=\frac{1}{2}\eta \varv_{\thetav_t}$.   In practice, since we can only estimate $\Bv$ by some $\hat{\Bv}$ and can only compute $\hat{U}$, a user defined friction term $\Cv$ is used (with $\Cv - \hat{\Bv}$ is semi-positive definite).  Thus the updates used for $\Delta \rhov$ for SGHMC:
 \begin{equation}
   - \eta \Cv \rhov_t - \eta \nabla \hat{U}(\thetav_t) + \mathcal{N}\lp \zerov, 2 \eta (\Cv-\hat{\Bv}) \rp 
 \end{equation}
 In our experiments we compute an online estimate $\hat{\varv}$ and set $\Cv = c \eye_p + \hat{\varv}$.  
 %As mentioned above, the main benefit of using SGHMC over SGLD is the ability of avoiding refreshing $\rhov$ while not accumulating Hamiltonian errors.

 {\bf Stochastic Gradient thermostats} (SGT) \cite{ding2014bayesian} addresses the difficulty of estimating $\hat{\Bv}$ by introducing a scalar variable $\xi$ who's addition to the Hamiltonian dynamics maintains the temperature of the system constant, i.e. it acts as a (Nose-Hoovier) thermostat \cite{leimkuhler:2009}.  The update equations remain simple: initialize $\xi = \Cv$ (or $c$), then for $t=1\ldots$
 \begin{eqnarray}
   \rhov_{t+1} & = & \rho_t - \eta \xi_t \rhov - \eta \nabla \hat{U}(\thetav_t) + \mathcal{N}\lp \zerov, 2 \eta_t \Cv \rp \\
   \thetav_{t+1} & = & \thetav_t + \eta \rhov_{t+1} \\
   \xi_{t+1} & = & \xi_t + \eta \lp  \rhov_{t+1}^T \rhov_{t+1} / p - 1 \rp
 \end{eqnarray}
 In summary, the hyperparameters required for these algorithms are $\eta$ and $\Cv$ (for SGHMC and SGNHT only), and in practice, some way of estimating $\hat{\varv}$ for SGHMC.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{HAMILTONIAN ABC}\label{sec:habc}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The general approach of applying Hamiltonian dynamics to ABC requires choosing one of the SGHD algorithms and then plugging in the ABC gradient approximation $\nabla \hat{U}(\thetav)$.  With this in mind we leave the details of the Hamiltonian updates to previous work \cite{welling2011bayesian,chen2014stochastic,ding2014bayesian} and focus on the details of how stochastic gradients are computed in the likelihood-free setting.

\subsection{Deterministic Representations of Simulations}
Implicit in each simulation run $\x \simsim \pi(\x | \thetav)$ is a sequence if internally generated random numbers that are used to produce random draws from $\pi(\x | \thetav)$.  These random numbers are important to HABC because we wish to control the stochasticity of the simulator when computing its gradient.  Furthermore, we will control the random numbers over multiple time steps.  Instead of keeping track of random numbers, we can equivalently keep a vector of $S$ random seeds $\omegav$.  This allows HABC to treat the simulation function $\pi(\x|\thetav)$ as a blackbox, outside of which we can control the random number generator (RNG), and represent $\x^{(s)}$ as the output of a deterministic function; i.e.  $\x^{(s)} = f( \thetav, \omega_s)$ instead of $\x^{(s)} \simsim \pi(\x | \thetav)$.   We include $\omegav$ as part of the state of our Markov chain.

\subsection{Kernel-$\eps$ versus Synthetic-likelihood -based Gradients}
In Section~\ref{sec:abc} we showed that the synthetic-likelihood representation of $\lleps(\thetav)$ is less sensitive to small choices of $\epsvec$.  This is particularly important to HABC as our gradient approximations are proportional to differences in $\lleps(\thetav)$; if the variance of the stochastic gradients is too high, then we must choose a very small step-size $\eta$, eliminating the usefulness of HMC for ABC.  Under the deterministic representation of $\x^{(s)}$, we can write the loglikelihoods
\begin{eqnarray}
 \lleps(\thetav) & \propto & \log \sum_s \mathcal{N}( \y | f( \thetav, \omega_s), \eps^2 )\\
                 & \approx & -\log \eps -\frac{1}{2 \eps^2} ( \y - f( \thetav, \omega_m) )^2 
\end{eqnarray}
In the second line we have again assumes $\eps$ is very small and $m$ is the index of the random seed producing the closest simulation to $\y$.  For a finite difference approximation, $\partial \lleps(\thetav)/\partial\thetav$ is
\begin{equation}
 \frac{1}{4\dtheta \eps^2} \lp ( \y - f( \thetav-\dtheta, \omega_m^-) )^2 -( \y - f( \thetav + \dtheta, \omega_m^+) )^2 \rp
\end{equation}

On the other hand, the synthetic-likelihood is stable; using deterministic representation, 
\begin{equation}
\mu_{\thetav} = \frac{1}{S} \sum_s f( \thetav, \omega_s ) ~~~~~~~~~~~~~ \sigma^s_{\thetav} = \frac{1}{S-1} \sum_s ( \mu_{\thetav} - f( \thetav, \omega_s ) )^2
\end{equation}
the gradients (for a 1-dim problem) use $\eps$ as a smoothness prior in $\partial \lleps(\thetav)/\partial\thetav$:
\begin{equation}
  -\frac{1}{2}\log\lp\frac{\sigma_{\theta^+}^2 + \eps^2}{\sigma_{\theta^-}^2 + \eps^2}\rp -\frac{( \y - \mu_{\thetav^+} )^2 }{2 (\sigma^2_{\thetav^+} + \eps^2 )} +\frac{( \y - \mu_{\thetav^-} )^2 }{2 (\sigma^2_{\thetav^-} + \eps^2 )} 
\end{equation}

\begin{figure}[t]
\vskip 0.2in
\begin{center}
\includegraphics[width=0.95\columnwidth]{./images/exp_varg_figure.pdf}
\caption{\small{Variance of gradient estimation using kernel-$\eps$ and SL for different values of $S\in\{5,50\}$ and fixed $\eps = 0.37$ (the same used in the other results).  For kernel-$\eps$, $\varv(\thetav_{\text{MAP}}) = 6.7 \pm 131.5$ and $\varv(\thetav_{\text{MAP}}) = 6.9 \pm 18.9$ (for $S=5$ and $S=50$, respectively).  For SL,$\varv(\thetav_{\text{MAP}}) = 6.35 \pm 42.4$ and $\varv(\thetav_{\text{MAP}}) = 0.62 \pm 5.2$ (for $S=5$ and $S=50$, respectively).  The ``true'' ABC gradient at $\thetav_{\text{MAP}}$ is $0.26$.  The ``true'' ABC loglikelihood is based on a Gaussian approximation (smoothed by $\eps$), with mean and variance set by moment matching with the true Gamma distribution (the sum of $N$ exponentials). Not only does kernel-$\eps$ keep a high variance for even large $S$, it seems to have a bias in its gradient estimate.  [look into this?] }}
\label{fig:exp-varg}
\end{center}
\vskip -0.2in
\end{figure}

\subsection{From Finite Differences to Simultaneous Perturbations}
\input{algo_fdsa_abc}
\input{algo_spsa_abc}
Algorithm~\ref{algo:fdsa} shows the {\em finite difference stochastic approximation} (FDSA) \cite{kiefer1952stochastic} to $\nabla U(\thetav)$ as a function of random seeds $\omegav$.  Note we have deliberately shown the deterministic simulations ($f$) outside of $\lleps$ to emphasize its dependence on $\x$.  The number of simulations required for FDSA is $2 S D$, which may be acceptable for some small ABC problems.  Our goal is to scale ABC to high-dimensions and for that we need an alternative stochastic approximation of $\nabla U(\thetav)$.

In the gradient-free setting, Spall \cite{spall1992multivariate, spall2000adaptive} provides a stochastic approximate to the true gradient using only $2$ forward simulations for any dimension $D$ (though the approximation can be improved by averaging $R$ estimates).  Spall's {\em simultaneous perturbation stochastic approximation} (SPSA) algorithm works as follows. Let $L$ be the gradient-free function we wish to optimize.  Each approximation randomly generates a {\em perturbation mask} (our name) $\Deltav$ of dimension $D=|\thetav|$ where entry $\deltad \sim \text{Bernouilli}(1/2)$ (actually, values $\pm 1$, not $0/1$).  Then $L$ is evaluated at $\thetav+\dtheta \Deltav$ and $\thetav-\dtheta \Deltav$, giving the gradient approximation $\gradvest(\thetav) \approx \partial L(\thetav) / \partial \thetav$:
\begin{equation}
  \gradvest(\thetav) = \frac{L\lp\thetav+\dtheta \Deltav\rp - L\lp\thetav-\dtheta \Deltav\rp}{2 \dtheta} \begin{bmatrix} 
                                   1/\Delta_1 \\
                                   1/\Delta_2 \\
                                   \vdots \\
                                   1/\Delta_D \\
                                \end{bmatrix}
\end{equation}
If we let $\gradvest_r(\thetav)$ be the estimate using perturbation mask $\Deltav_r$, the estimate $\gradvest(\thetav)$ can be improved by averaging $\gradvest(\thetav) = 1/R \sum_r \gradvest_r(\thetav)$.
Algorithm~\ref{algo:spsa} shows SPSA to estimate $\nabla U(\thetav)$.  The number of simulations required for SPSA is $2 S R$, where $R \geq 1$.

Variations of SPSA include {\em one-sided} SPSA \cite{spall2000adaptive} (we use what Spall calls 2SPSA) and an algorithm for estimating the Hessian based on the same principal as SPSA \cite{spall2005monte}.   The one-sided version is attractive computationally, but for HABC, the updates for $\thetav$ require simulating two-sides anyway, so using 2SPSA makes more sense.  SPSA has also been used within a procedure for maximum-likelihood estimation for hidden Markov models using ABC \cite{Ehrlich2013}.

\begin{figure}[t]
\vskip 0.2in
\begin{center}
\includegraphics[width=0.95\columnwidth]{./images/exp_crn_figure.pdf}
\caption{\small{A view of a simulator in terms of common random numbers.  The horizontal line represents $\y$ and red shading $\pm 2\eps$.  The shaded curved region represents $2\sigma$ of $\pi(\x|\thetav)$.  The dashed lines are $f(\thetav, \omega_s)$ for several values of $\omega$.  The blue circles are potential random samples from $\pi(\x|\thetav)$.  For a fixed value $\omega_s$, the simulator produces deterministic outputs that change smoothly, even though the simulator itself is quite noisy.}}
\label{fig:exp-crns}
\end{center}
\vskip -0.2in
\end{figure}

\subsection{Sticky Random Numbers}
An obvious use of common random numbers, as previously demonstrated for SPSA \cite{kleinman1999simulation}, is within the gradient approximation so that the noise from the simulator has less of an influence that $\dtheta$.  In addition to this use, we have found that using the same random seeds over multiple time-steps improves the performance of HABC (TBD).  This is very similar to using dependent random streams to drive MCMC \cite{Murray2012,Neal2012}, the main difference we believe is that we are using the Hamiltonian dynamics to drive proposal in $\thetav$ and using $\omegav$ to give persistent simulations, reducing the variance from the unknown simulator noise function.

Our updates for $\omegav$ are simple.  At each time step $t$, each seed is refreshed to a new random value with probability $\gamma$, i.e. $\omega_s \leftarrow \omega_s$ with probability $\gamma$, and $\omega_s \leftarrow \mathcal{U}(1,W)$, where $W=10^8$ in our algorithms.  By observing the simulation at different $\thetav$ but using same seeds, we can eliminate much of the simulator noise from the Markov chain.  By treating $\omegav$ as part of the state of the Markov chain, we can smoothly move along noiseless outputs for several iterations and then propose new seeds.  Though proposing new seeds produces a valid Metropolis-Hastings algorithm (keeping $\thetav$ fixed and changing $\omegav$), we found that using {\em sticky} random seeds works well.  These seeds can be precomputed before the chain is run and can be seen as apart from the state of the Markov chain.  [CONFUSING]

%[We may want anti-thetic instead of CRNs, ie overrelax versus underrelax, etc.]

\begin{figure}[t]
\vskip 0.2in
\begin{center}
\includegraphics[width=0.45\columnwidth]{./images/exp-SL-MCMC-posterior-hist-omega-rate-0p01-chain0.pdf}
\includegraphics[width=0.45\columnwidth]{./images/exp-SG-HMC-posterior-hist-omega-rate-0p01-chain0.pdf}
\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Langevin-posterior-hist-omega-rate-0p01-chain0.pdf}
\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior-hist-omega-rate-0p01-chain0.pdf}
%\includegraphics[width=0.45\columnwidth]{./images/exp-ABC-MCMC-posterior_hist.pdf}
%\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
%\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Langevin-posterior_hist.pdf}
%\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
\caption{\small{Using sticky numbers at rate $1\%$}}
\label{fig:exp-posteriors}
\end{center}
\vskip -0.2in
\end{figure}

\begin{figure}[t]
\vskip 0.2in
\begin{center}
\includegraphics[width=0.45\columnwidth]{./images/exp-SL-MCMC-posterior-hist-omega-rate-100p0-chain0.pdf}
\includegraphics[width=0.45\columnwidth]{./images/exp-SG-HMC-posterior-hist-omega-rate-100p0-chain0.pdf}
\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Langevin-posterior-hist-omega-rate-100p0-chain0.pdf}
\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior-hist-omega-rate-100p0-chain0.pdf}
%\includegraphics[width=0.45\columnwidth]{./images/exp-ABC-MCMC-posterior_hist.pdf}
%\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
%\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Langevin-posterior_hist.pdf}
%\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
\caption{\small{No sticky numbers}}
\label{fig:exp-posteriors}
\end{center}
\vskip -0.2in
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Demonstration}\label{sec:demo}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We use a simple $D=1$ problem to demonstrate HABC.  Let $y= \frac{1}{N} \sum_{i} e_i$, where $e_i \sim \text{Exp}(1/\thetastar)$; $\thetastar = 0.15$, $N=20$, and $y=7.74$ in our concrete example.  Assuming $\pi(\theta ) = \text{Gamma}(\alpha, \beta)$, the true posterior is a gamma distribution with shape $\alpha+N$ and rate $\beta + N y$.  Our simulator therefore generates the average of $N$ exponential random variates with rate $\lambda = 1/\theta$.  Data $x \simsim \pi(x|\theta)$ are shown in Figure~\ref{fig:exp-crns}.  We have explicitly shown the smoothness of the simulator by generating data along trajectories of fixed seeds $\omega_s$; i.e. for several $\omega_s$ we vary $\theta$ (dashed lines are function $f(\theta, \omega_s)$) and randomly reveal data (blue circles).  The horizontal line with shading indicates $y \pm 2 \eps$, where $\eps = 0.37$ is used throughout the demonstration.

\begin{figure*}[t]
%\vskip 0.2in
\setlength{\linewidth}{\textwidth}
\setlength{\hsize}{\textwidth}
\begin{center}
\includegraphics[width=0.95\columnwidth]{./images/exp-SL-MCMC-theta-timeseries-omega-rate-0p01-chain0.pdf}
\includegraphics[width=0.95\columnwidth]{./images/exp-SG-HMC-theta-timeseries-omega-rate-0p01-chain0.pdf}
\includegraphics[width=0.95\columnwidth]{./images/exp-SG-Langevin-theta-timeseries-omega-rate-0p01-chain0.pdf}
\includegraphics[width=0.95\columnwidth]{./images/exp-SG-Thermostats-theta-timeseries-omega-rate-0p01-chain0.pdf}
\caption{\small{Using sticky numbers at rate $1\%$}}
\label{fig:exp-theta-traces}
\end{center}
\vskip -0.2in
\end{figure*}

\begin{figure*}[t]
%\vskip 0.2in
\setlength{\linewidth}{\textwidth}
\setlength{\hsize}{\textwidth}
\begin{center}
\includegraphics[width=0.95\columnwidth]{./images/exp-SL-MCMC-theta-timeseries-omega-rate-100p0-chain0.pdf}
\includegraphics[width=0.95\columnwidth]{./images/exp-SG-HMC-theta-timeseries-omega-rate-100p0-chain0.pdf}
\includegraphics[width=0.95\columnwidth]{./images/exp-SG-Langevin-theta-timeseries-omega-rate-100p0-chain0.pdf}
\includegraphics[width=0.95\columnwidth]{./images/exp-SG-Thermostats-theta-timeseries-omega-rate-100p0-chain0.pdf}
\caption{\small{No sticky numbers}}
\label{fig:exp-theta-traces}
\end{center}
\vskip -0.2in
\end{figure*}

\subsection{Bias and Variance of $\nabla \hat{U}(\thetav)$}
At $\theta_{\text{MAP}}$ we estimated the gradient $\nabla \hat{U}(\theta)$ using finite differences for both the kernel-$epsilon$ and synthetic-likelihood representation of $\pi_{eps}(y | x )$.  For $S=5$ and $S=50$, $10K$ repetitions of computing the gradients was performed; there histograms are shown in Figure~\ref{fig:exp-varg}.  At $S=5$ both methods reveal a small bias (values X and Y for kernel-$\eps$ and SL, respectively).  At $S=50$, the bias for kernel-$\eps$ is much smaller, but its variance remains high; for SL, the bias remains unchanged but the variance has decreased significantly (values X pm S and Y pm T). The bias in the SL representation comes from the Gaussian approximation $\pi(y|\theta)$ which we can compute analytically using the CLT as $\text{N}( 1/\theta, 1/N \theta^2 + \eps^2)$. 
(We believe this bias is due to the normal approximation assuming a symmetric mode as the bias goes to zero at the mean of the true posterior).  Note that this bias is directional signal only appears nears the mode.  For the $S=5$ setting in our experiment both approaches contain similar bias. [REWRITE END]

\subsection{Posterior Inference using HABC}
In Figures~\ref{fig:exp-theta-traces} we show the posterior trace plots for ABC-MCMC, SGLD, SGHMC, and SGNHT versions of HABC using SL gradient estimates ($S=5$).  The behavior of the trajectories is quite different, though they all produce the same correct posterior estimates (Figure~\ref{fig:exp-posteriors}).  In Table~\ref{tab:exp-posteriors-nosticky} we report the {\em total variational distance} between the true posterior and the ABC posteriors after different numbers of samples (repeated 5 times).

\begin{table}[h]
\caption{Posterior inference for Exponential problem}
\label{tab:exp-prob-no-sticky}
\begin{center}
\begin{tabular}{l||c|c|c|c|c}
{\bf Algo}  & 100 & 1000 & 10K & 50K \\
\hline \\
SL-MCMC &  $0.26$ & $0.099$ & $0.047$ & $0.045$ \\
SGLD    &  $0.42$ & $0.22$ & $0.14$ & $0.045$ \\
SGHMC   &  $0.26$ & $0.099$ & $0.047$ & $0.045$ \\
SGNHT   &  $0.26$ & $0.099$ & $0.047$ & $0.045$ \\
\end{tabular}
\end{center}
\end{table}

\begin{table}[h]
\caption{Posterior inference for Exponential problem}
\label{tab:exp-prob-no-sticky}
\begin{center}
\begin{tabular}{l||c|c|c|c|c}
{\bf Algo}  & 100 & 1000 & 10K & 50K \\
\hline \\
SL-MCMC &  $0.23$ & $0.12$ & $0.053$ & $0.044$ \\
SGLD    &  $0.26$ & $0.099$ & $0.047$ & $0.045$ \\
SGHMC   &  $0.26$ & $0.099$ & $0.047$ & $0.045$ \\
SGNHT   &  $0.26$ & $0.099$ & $0.047$ & $0.045$ \\
\end{tabular}
\end{center}
\end{table}

% DIR: exp-eta-0p005-dtheta-0p01-C10-S5-omega-0p01
% SL-MCMC omega_rate 100%
% 
% times 10,   100,  1000,  5000, 10000, 20000, 30000, 40000, 50000
% mean 0.6052864 ,  0.26126289,  0.09895101,  0.05295502,  0.04721529,0.04513989,  0.04622511,  0.04491134,  0.04539522
% std 0.09348374,  0.046628  ,  0.01530586,  0.00861296,  0.00604954, 0.00239858,  0.00327312,  0.00303499,  0.00242383
% 
% omega_Rate 1%
% mean 0.63282705,  0.23141512,  0.11508726,  0.07034678,  0.05338616, 0.05246243,  0.0538948 ,  0.0507813 ,  0.04435618
% std 0.11183616,  0.05139663,  0.05390608,  0.03376141,  0.01785321, 0.0125916 ,  0.00662542,  0.01255384,  0.00603819

% SGLD omega_rate 100%
% array([ 0.66893503,  0.41631787,  0.22451057,  0.15184296,  0.14112572, 0.15204372,  0.14857288,  0.15116372,  0.15076312])
% std(0)array([ 0.08087815,  0.11567097,  0.04555416,  0.03197215,  0.01855117, 0.01417793,  0.01341749,  0.01312386,  0.00855595])
% SGLD omega_rate 1%
% mean 0.75547928,  0.35519016,  0.17736532,  0.09150572,  0.06159863,0.0553024 ,  0.06191501,  0.05637656,  0.05472899
% std 0.11641948,  0.09003194,  0.08414296,  0.03832745,  0.01937652, 0.02211211,  0.01263801,  0.00796597,  0.00836038
% SGHMC omega 100%
% array([ 0.59033585,  0.25581098,  0.24731947,  0.22616621,  0.22842994,
%         0.22780039,  0.22743227,  0.2256994 ,  0.22623605])
%
% In [25]: tvd.std(0)
% Out[25]:
% array([ 0.1870709 ,  0.05443133,  0.04280439,  0.01782881,  0.01695731,
%         0.01044903,  0.00970157,  0.0081758 ,  0.00759141])
% omega 1%
% array([ 0.51193673,  0.36801527,  0.16197407,  0.09853979,  0.0929902 ,
%         0.0976414 ,  0.09743771,  0.09113688,  0.08687801])
%
% In [28]: tvd.std(0)
% Out[28]:
% array([ 0.17030828,  0.10124135,  0.0653025 ,  0.0180961 ,  0.01414473,
%         0.01492984,  0.0169915 ,  0.01351012,  0.009596  ])
% 
% SGTHEMO
% omega 100%
% In [33]: tvd.mean(0)
% Out[33]:
% array([ 0.61253343,  0.2419202 ,  0.15161436,  0.16610309,  0.18284641,
%         0.18875997,  0.19055003,  0.1928232 ,  0.19340706])
%
% In [34]: tvd.std(0)
% Out[34]:
% array([ 0.17654396,  0.04334123,  0.04052037,  0.01846459,  0.0212887 ,
%         0.009026  ,  0.00640309,  0.00575303,  0.00415439])
%

% omega 1%
% In [30]: tvd.std(0)
% Out[30]:
% array([ 0.17343684,  0.10098802,  0.06547166,  0.02232569,  0.01339252,
%         0.01902682,  0.0181153 ,  0.01225494,  0.00760486])
%
% In [31]: tvd.mean(0)
% Out[31]:
% array([ 0.47453793,  0.33112199,  0.14890592,  0.07600464,  0.06685565,
%         0.06371391,  0.06363715,  0.05882387,  0.05590603])

\subsection{Sticky Random Numbers}
Finally we compare posterior estimates using sticky CRNs.  Using a flip rate of $1\%$, i.e. each $\omega_s$ has $1\%$ chance of changing randomly at each time-step.  In Figure~\ref{exp-traces-x-sticky} we show the traces plots of $<x^{(s)>}$ for the different methods.  While the Hamiltonian dynamics provide persistence in $\thetav$-space, the sticky random numbers do the same for the ABC statistics.  The Table~\ref{tab:exp-tvd-sticky} we report the same {\em tvd} error using sticky numbers.  [DISCUSS difference with no-sticky]. The potential benefit of using persistent random seeds is shown for all methods, including ABC-MCMC.

%\subsection{Visualizing Noise from Gradient versus Injected Noise}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Experiments}\label{sec:experiments}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We present experimental results comparing HABC with standard ABC-MCMC for two challenging simulators.  The first is low-dimensional ($D=6$), but exhibits chaotic behavior.  For the second problem we apply the techniques of HABC to a Bayesian image classifier and autoencoder.  The benefit of this problem is we can compare directly with the SGHD algorithms using the analytic gradients.  This problem is has very high-dimension ($D>15000$), yet we are able to successfully apply SPSA-based gradients to HABC.

\subsection{Blowfly}\label{sec:bf}
For these experiments, a simulator of adult sheep blowfly populations \cite{wood2010statistical} is used with statistics set to those from \cite{Meeds2014GpsUai}.  The observational vector $\y$ is a time-series of a fly population counted daily. The population dynamics are modeled using a stochastic differential equation\footnote{Equation~1 in Section 1.2.3 of the supplementary information in \cite{wood2010statistical}.}
\begin{equation}
N_{t+1} = P N_{t-\tau} \exp(-N_{t-\tau}/N_0) e_t + N_t \exp(-\delta \epsilon_t) \nonumber
\end{equation}
where $e_t \sim  \mathcal{G}( 1/{\sigma_p^2},1/{\sigma_p^2})$ and $\epsilon_t 
 \sim  \mathcal{G}( 1/{\sigma_d^2},1/{\sigma_d^2})$  
are sources of noise, and $\tau$ is an integer.  In total, there are $D=6$ parameters $\theta = \{ \log P, \log \delta, \log N_0, \log \sigma_d, \log \sigma_p, \tau\}$.  As \cite{Meeds2014GpsUai} we place broad log-normal priors over $\theta_{1\ldots 5}$ and a Poisson prior over $\tau$.  This is considered a challenging problem because slight changes to some parameter settings can produce degenerate $\x$, while others settings can be very noisy due to the chaotic nature of the equations.  The statistics from \cite{Meeds2014GpsUai} are used ($J=10$): the log average of $4$ quantiles of $N/1000$, the average of $4$ quantiles of the first-order differences in $N/1000$, and the number of maximal population peaks under two different thresholds.

%\subsubsection{Posterior Inference using HABC}
We compare difference HABC algorithms with ABC-MCMC for the blowfly population problem.  We use $\epsvec = \{ 1/2,1/2,1/2,1/2, 1/4,1/4,1/4,1/4,3/4,3/4 \}$ (slightly different $\epsvec$ from \cite{Meeds2014GpsUai}) and $S=10$ for all experiments.  We use SPSA with $R=1$ using SL log-likelihoods for all HABC gradient estimates.  Note that this requires $2S$ forward simulations per time-step, the same as marginal ABC and twice pseudo-marginal ABC.

Figure~\ref{fig:bf-posterior-histograms} show the posterior distributions for $\log P, \log \delta$ and the posterior predictive distributions of [FILL IN].  Traces of $\thetav$ are shown for SGHMC and SGNHT in Figure~\ref{}.  Finally, we compare the convergence to $\y$ using the online posterior predictive distribution in Figure~\ref{}.  By using stick random numbers, the convergence is [FILL IN].

\subsection{Bayesian Autoencoders}\label{sec:auto}
We perform Bayesian inference on an autoencoder neural network using MNIST images as observations.  We apply two inference algorithms: SGNHT and HABC.  For SGNHT, we use the known gradient and for HABC we apply 2SPSA.  After 1000 iterations we collect parameter vectors every 100 iterations for a total of 100 sets. (Do we use SGHMC to estimate Bhat in first 1000 iterations?).  Use same batches at each iteration for SGNHT and HABC?

results: convergence of training error (batches), convergence in test error using parameters collected (record nbr of forward or backward passes required), show mean and variance of filters (how?), compare with using the last parameter sample only (stochastic optimization).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{DISCUSSION} \label{sec:discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \begin{itemize}
%   \item
%   \item
% \end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{CONCLUSION} \label{sec:conclusion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{itemize}
  \item New set of ABC algorithms for efficient exploration of posterior distribution.
  \item Key: opens the door to high-dimensional ABC inference.
  \item [Test for more accurate gradients] For ABC, the gradients we compute are stochastic because we must approximate $\pi(\x | \thetav)$ with forward simulations.  We can make the approximation arbitrarily good by running more simulations, but this is often unnecessary for sampling, because we can make use of the uncertainty in our current approximation to decide if our sample is correct or not (cite UAI2014?). 
  \item Unbiased gradients with low noise.  SL bias might be intolerable; use KDE to get stable unbiased gradients?
\end{itemize}

A further complication, which we do not address on this paper, is the forward simulation might be, and for many interesting problems is, expensive.  We would therefore prefer to resort to some sort of surrogate modeling \cite{Meeds2014GpsUai,wilkinson:2014, michael} to avoid simulations whenever possible. We address this scenario in Section~\ref{future}.  Shown for relatively fast simulators.  Extensions to surrogates with GPS (Meeds2014GpsUai, rasmussen:2003, wilkinson:2014) can use derivative of GP for likelihood as GP of gradient.  These could be matched by occasional gradient estimates (Osborne).

%\subsubsection*{References}
%{
\clearpage
\bibliographystyle{icml2014}
\bibliography{abcsgld}
%}

\end{document}

### 2015-03-02 scripts_for_figures/exp_problem_tables.py
exp posterior table

@@ -1,0 +1,66 @@
+import numpy as np
+import pylab as pp
+import scipy as sp
+
+results_dir = "../../abc_sgld/code/exponential_problem/images/"
+
+algonames = ["SL-ABC", "SGLD", "SGHMC", "SGNHT"]
+
+times = np.loadtxt( results_dir+"exp-SL-MCMC-tvd-times-omega-rate-0p01.txt" )
+times = times.astype(int)
+sticky_files = []
+sticky_files.append( results_dir+"exp-SL-MCMC-tvd-omega-rate-0p01.txt" )
+sticky_files.append( results_dir+"exp-SG-Langevin-tvd-omega-rate-0p5.txt" )
+sticky_files.append( results_dir+"exp-SG-HMC-tvd-omega-rate-0p01.txt" )
+sticky_files.append( results_dir+"exp-SG-Thermostats-tvd-omega-rate-0p01.txt" )
+
+
+non_sticky_files = []
+non_sticky_files.append( results_dir+"exp-SL-MCMC-tvd-omega-rate-100p0.txt" )
+non_sticky_files.append( results_dir+"exp-SG-Langevin-tvd-omega-rate-100p0.txt" )
+non_sticky_files.append( results_dir+"exp-SG-HMC-tvd-omega-rate-100p0.txt" )
+non_sticky_files.append( results_dir+"exp-SG-Thermostats-tvd-omega-rate-100p0.txt" )
+
+print times
+time_ids = [2,4,8]
+
+for algo, fn in zip( algonames, sticky_files ):
+  tdv = np.loadtxt( fn )
+  mn = tdv.mean(0)
+  std = tdv.std(0)
+  for idx in time_ids:
+    print "%s at %08d has %3.4f tdv"%(algo,times[idx],mn[idx])
+  #print mn
+
+for algo, fn in zip( algonames, non_sticky_files ):
+  tdv = np.loadtxt( fn )
+  mn = tdv.mean(0)
+  std = tdv.std(0)
+  for idx in time_ids:
+    print "%s at %08d has %3.4f tdv"%(algo,times[idx],mn[idx])
+
+algonames = ["SL-ABC", "SGLD", "SGNHT", "SGNHT*"]
+files2use = []
+files2use.append( results_dir+"exp-SL-MCMC-tvd-omega-rate-100p0.txt" )
+files2use.append( results_dir+"exp-SG-Langevin-tvd-omega-rate-100p0.txt" )
[…]

### 2015-03-02 draft.tex
exp posterior table
minor exp edits

@@
 
-\begin{figure}[t]
-\vskip 0.2in
-\begin{center}
-\includegraphics[width=0.95\columnwidth]{./images/exp_varg_figure.pdf}
-\caption{\small{Variance of gradient estimation using kernel-$\eps$ and SL for different values of $S\in\{5,50\}$ and fixed $\eps = 0.37$ (the same used in the other results).
-For kernel-$\eps$, $\varv(\thetav_{\text{MAP}}) = 6.7 \pm 131.5$ and $\varv(\thetav_{\text{MAP}}) = 6.9 \pm 18.9$ (for $S=5$ and $S=50$, respectively).
-For SL,$\varv(\thetav_{\text{MAP}}) = 6.35 \pm 42.4$ and $\varv(\thetav_{\text{MAP}}) = 0.62 \pm 5.2$ (for $S=5$ and $S=50$, respectively).
-The ``true'' ABC gradient at $\thetav_{\text{MAP}}$ is $0.26$.
-The ``true'' ABC loglikelihood is based on a Gaussian approximation (smoothed by $\eps$), with mean and variance set by moment matching with the true Gamma distribution (the sum of $N$ exponentials).
-Not only does kernel-$\eps$ keep a high variance for even large $S$, it seems to have a bias in its gradient estimate.
-[look into this?] }}
-\label{fig:exp-varg}
-\end{center}
-\vskip -0.2in
-\end{figure}
-
 \subsection{From Finite Differences to Simultaneous Perturbations}
@@
 
+\begin{figure}[t]
+\vskip 0.2in
+\begin{center}
+\includegraphics[width=0.95\columnwidth]{./images/exp_varg_figure.pdf}
+\caption{\small{Variance of gradient estimation using kernel-$\eps$ and SL for different values of $S\in\{5,50\}$ and fixed $\eps = 0.37$ (the same used in the other results).
+For kernel-$\eps$, $\varv(\thetav_{\text{MAP}}) = 6.7 \pm 131.5$ and $\varv(\thetav_{\text{MAP}}) = 6.9 \pm 18.9$ (for $S=5$ and $S=50$, respectively).
+For SL,$\varv(\thetav_{\text{MAP}}) = 6.35 \pm 42.4$ and $\varv(\thetav_{\text{MAP}}) = 0.62 \pm 5.2$ (for $S=5$ and $S=50$, respectively).
+The ``true'' ABC gradient at $\thetav_{\text{MAP}}$ is $0.26$.
+The ``true'' ABC loglikelihood is based on a Gaussian approximation (smoothed by $\eps$), with mean and variance set by moment matching with the true Gamma distribution (the sum of $N$ exponentials).
+Not only does kernel-$\eps$ keep a high variance for even large $S$, it seems to have a bias in its gradient estimate.
+[look into this?] }}
+\label{fig:exp-varg}
+\end{center}
+\vskip -0.2in
+\end{figure}
+
 \subsection{Sticky Random Numbers}
@@
 %[We may want anti-thetic instead of CRNs, ie overrelax versus underrelax, etc.]
-
 \begin{figure}[t]
@@
 \begin{center}
-\includegraphics[width=0.45\columnwidth]{./images/exp-SL-MCMC-posterior-hist-omega-rate-0p01-chain0.pdf}
-\includegraphics[width=0.45\columnwidth]{./images/exp-SG-HMC-posterior-hist-omega-rate-0p01-chain0.pdf}
-\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Langevin-posterior-hist-omega-rate-0p01-chain0.pdf}
-\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior-hist-omega-rate-0p01-chain0.pdf}
-%\includegraphics[width=0.45\columnwidth]{./images/exp-ABC-MCMC-posterior_hist.pdf}
-%\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
-%\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Langevin-posterior_hist.pdf}
-%\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
-\caption{\small{Using sticky numbers at rate $1\%$}}
-\label{fig:exp-posteriors}
-\end{center}
-\vskip -0.2in
-\end{figure}
+\includegraphics[width=0.32\columnwidth]{./images/exp-SL-MCMC-posterior-hist-omega-rate-100p0-chain0.pdf}
+%\includegraphics[width=0.45\columnwidth]{./images/exp-SG-HMC-posterior-hist-omega-rate-100p0-chain0.pdf}
+\includegraphics[width=0.32\columnwidth]{./images/exp-SG-Langevin-posterior-hist-omega-rate-100p0-chain0.pdf}
+\includegraphics[width=0.32\columnwidth]{./images/exp-SG-Thermostats-posterior-hist-omega-rate-100p0-chain0.pdf}
 
-\begin{figure}[t]
-\vskip 0.2in
-\begin{center}
-\includegraphics[width=0.45\columnwidth]{./images/exp-SL-MCMC-posterior-hist-omega-rate-100p0-chain0.pdf}
-\includegraphics[width=0.45\columnwidth]{./images/exp-SG-HMC-posterior-hist-omega-rate-100p0-chain0.pdf}
-\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Langevin-posterior-hist-omega-rate-100p0-chain0.pdf}
-\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior-hist-omega-rate-100p0-chain0.pdf}
+\includegraphics[width=0.32\columnwidth]{./images/exp-SL-MCMC-posterior-hist-omega-rate-0p01-chain0.pdf}
+%\includegraphics[width=0.45\columnwidth]{./images/exp-SG-HMC-posterior-hist-omega-rate-0p01-chain0.pdf}
+\includegraphics[width=0.32\columnwidth]{./images/exp-SG-Langevin-posterior-hist-omega-rate-0p01-chain0.pdf}
+\includegraphics[width=0.32\columnwidth]{./images/exp-SG-Thermostats-posterior-hist-omega-rate-0p01-chain0.pdf}
+
 %\includegraphics[width=0.45\columnwidth]{./images/exp-ABC-MCMC-posterior_hist.pdf}
@@
 \end{figure} 
+%
+% \begin{figure}[t]
+% \vskip 0.2in
+% \begin{center}
+% \includegraphics[width=0.45\columnwidth]{./images/exp-SL-MCMC-posterior-hist-omega-rate-0p01-chain0.pdf}
+% \includegraphics[width=0.45\columnwidth]{./images/exp-SG-HMC-posterior-hist-omega-rate-0p01-chain0.pdf}
+% \includegraphics[width=0.45\columnwidth]{./images/exp-SG-Langevin-posterior-hist-omega-rate-0p01-chain0.pdf}
+% \includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior-hist-omega-rate-0p01-chain0.pdf}
+% %\includegraphics[width=0.45\columnwidth]{./images/exp-ABC-MCMC-posterior_hist.pdf}
+% %\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
+% %\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Langevin-posterior_hist.pdf}
+% %\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
+% \caption{\small{Using sticky numbers at rate $1\%$}}
+% \label{fig:exp-posteriors}
+% \end{center}
+% \vskip -0.2in
+% \end{figure}
 
@@
 \begin{center}
+\includegraphics[width=0.95\columnwidth]{./images/exp-SL-MCMC-theta-timeseries-omega-rate-100p0-chain0.pdf}
 \includegraphics[width=0.95\columnwidth]{./images/exp-SL-MCMC-theta-timeseries-omega-rate-0p01-chain0.pdf}
-\includegraphics[width=0.95\columnwidth]{./images/exp-SG-HMC-theta-timeseries-omega-rate-0p01-chain0.pdf}
+%\includegraphics[width=0.95\columnwidth]{./images/exp-SG-HMC-theta-timeseries-omega-rate-100p0-chain0.pdf}
+\includegraphics[width=0.95\columnwidth]{./images/exp-SG-Langevin-theta-timeseries-omega-rate-100p0-chain0.pdf}
 \includegraphics[width=0.95\columnwidth]{./images/exp-SG-Langevin-theta-timeseries-omega-rate-0p01-chain0.pdf}
+\includegraphics[width=0.95\columnwidth]{./images/exp-SG-Thermostats-theta-timeseries-omega-rate-100p0-chain0.pdf}
 \includegraphics[width=0.95\columnwidth]{./images/exp-SG-Thermostats-theta-timeseries-omega-rate-0p01-chain0.pdf}
-\caption{\small{Using sticky numbers at rate $1\%$}}
+\caption{\small{No sticky numbers  Explanation thermostats: with no sticky, need to lower eta, so it shows smoother trajectory.
+With sticky, the gradients are more consistent, allowing for a larger eta and therfore more noise injection.}}
 \label{fig:exp-theta-traces}
@@
 
-\begin{figure*}[t]
-%\vskip 0.2in
-\setlength{\linewidth}{\textwidth}
-\setlength{\hsize}{\textwidth}
-\begin{center}
-\includegraphics[width=0.95\columnwidth]{./images/exp-SL-MCMC-theta-timeseries-omega-rate-100p0-chain0.pdf}
-\includegraphics[width=0.95\columnwidth]{./images/exp-SG-HMC-theta-timeseries-omega-rate-100p0-chain0.pdf}
-\includegraphics[width=0.95\columnwidth]{./images/exp-SG-Langevin-theta-timeseries-omega-rate-100p0-chain0.pdf}
-\includegraphics[width=0.95\columnwidth]{./images/exp-SG-Thermostats-theta-timeseries-omega-rate-100p0-chain0.pdf}
-\caption{\small{No sticky numbers}}
-\label{fig:exp-theta-traces}
-\end{center}
-\vskip -0.2in
-\end{figure*}
+% \begin{figure*}[t]
+% %\vskip 0.2in
+% \setlength{\linewidth}{\textwidth}
+% \setlength{\hsize}{\textwidth}
+% \begin{center}
+% \includegraphics[width=0.95\columnwidth]{./images/exp-SL-MCMC-theta-timeseries-omega-rate-0p01-chain0.pdf}
+% %\includegraphics[width=0.95\columnwidth]{./images/exp-SG-HMC-theta-timeseries-omega-rate-0p01-chain0.pdf}
+% \includegraphics[width=0.95\columnwidth]{./images/exp-SG-Langevin-theta-timeseries-omega-rate-0p01-chain0.pdf}
+% \includegraphics[width=0.95\columnwidth]{./images/exp-SG-Thermostats-theta-timeseries-omega-rate-0p01-chain0.pdf}
+% \caption{\small{Using sticky numbers at rate $1\%$}}
+% \label{fig:exp-theta-traces}
+% \end{center}
+% \vskip -0.2in
+% \end{figure*}
 
@@
 
+% \begin{table}[h]
+% \caption{Posterior inference for Exponential problem}
+% \label{tab:exp-prob-no-sticky}
+% \begin{center}
+% \begin{tabular}{l||c|c|c|c|c}
+% {\bf Algo}  & 100 & 1000 & 10K & 50K \\
+% \hline \\
+% SL-MCMC &  $0.26$ & $0.099$ & $0.047$ & $0.045$ \\
+% SGLD    &  $0.42$ & $0.22$ & $0.14$ & $0.045$ \\
+% SGHMC   &  $0.26$ & $0.099$ & $0.047$ & $0.045$ \\
+% SGNHT   &  $0.26$ & $0.099$ & $0.047$ & $0.045$ \\
+% \end{tabular}
+% \end{center}
+% \end{table}
+
 \begin{table}[h]
 \caption{Posterior inference for Exponential problem}
-\label{tab:exp-prob-no-sticky}
+\label{tab:exp-posterior}
 \begin{center}
-\begin{tabular}{l||c|c|c|c|c}
-{\bf Algo}  & 100 & 1000 & 10K & 50K \\
-\hline \\
-SL-MCMC &  $0.26$ & $0.099$ & $0.047$ & $0.045$ \\
-SGLD    &  $0.42$ & $0.22$ & $0.14$ & $0.045$ \\
-SGHMC   &  $0.26$ & $0.099$ & $0.047$ & $0.045$ \\
-SGNHT   &  $0.26$ & $0.099$ & $0.047$ & $0.045$ \\
+\begin{tabular}{|l||c|c|c|c|c}
+  \hline 
+Algo & $1000$ & $10000$ & $50000$ \\ \hline \hline 
+SL-ABC & $0.099$ & $0.047$ & $0.045$ \\
+SGLD & $0.082$ & $0.049$ & $0.048$ \\
+SGNHT & $0.271$ & $0.162$ & $0.161$ \\
+SGNHT* & $0.144$ & $0.056$ & $0.075$ \\ \hline 
 \end{tabular}
@@
 
-\begin{table}[h]
-\caption{Posterior inference for Exponential problem}
-\label{tab:exp-prob-no-sticky}
-\begin{center}
-\begin{tabular}{l||c|c|c|c|c}
-{\bf Algo}  & 100 & 1000 & 10K & 50K \\
-\hline \\
-SL-MCMC &  $0.23$ & $0.12$ & $0.053$ & $0.044$ \\
-SGLD    &  $0.26$ & $0.099$ & $0.047$ & $0.045$ \\
-SGHMC   &  $0.26$ & $0.099$ & $0.047$ & $0.045$ \\
-SGNHT   &  $0.26$ & $0.099$ & $0.047$ & $0.045$ \\
-\end{tabular}
-\end{center}
-\end{table}
+% \begin{table}[h]
+% \caption{Posterior inference for Exponential problem}
+% \label{tab:exp-prob-no-sticky}
+% \begin{center}
+% \begin{tabular}{l||c|c|c|c|c}
+% {\bf Algo}  & 100 & 1000 & 10K & 50K \\
+% \hline \\
+% SL-MCMC &  $0.23$ & $0.12$ & $0.053$ & $0.044$ \\
+% SGLD    &  $0.26$ & $0.099$ & $0.047$ & $0.045$ \\
+% SGHMC   &  $0.26$ & $0.099$ & $0.047$ & $0.045$ \\
+% SGNHT   &  $0.26$ & $0.099$ & $0.047$ & $0.045$ \\
+% \end{tabular}
+% \end{center}
+% \end{table}
 

### 2015-03-02 draft.tex
edits, rewrite sticky

@@
 If the pseudo-data $\x$ is ``close'' to observations $\y$, then $\thetav$ is kept as a sample from the approximate posterior.
-Parameters $thetav$ are then adjusted, depending upon the algorithm, to obtain the next sample.
+Parameters $\thetav$ are then adjusted, depending upon the algorithm, to obtain the next sample.
 
@@
 In HABC, we use forward simulations to approximate the likelihood-free gradient.
-The key difference between SGHD methods and HABC is that the stochasticity of the gradient does not come from approximating the full data gradient with a batch gradient, but by the stochasticity of the simulator.
+The key difference between SGHD methods and HABC is that the stochasticity of the gradient does not come from approximating the full data gradient with a mini-batch gradient, but by the stochasticity of the simulator.
 It is therefore not the expense of the simulator (though this could very well be the case for many interesting simulation-based models -- see Section~\ref{sec:conclusion}) that requires an approximation to the gradient, but the likelihood-free nature of the problem.
@@
 There are several difficulties in estimating gradients of likelihood-free models that we address with HABC.
-The first is do to the form of the ABC log-likelihood.
+The first is due to the form of the ABC log-likelihood.
 As we show in Section~\ref{sec:abc}, using a conditional model for $\pi( \x | \thetav )$ provides an estimate of the ABC likelihood that is less sensitive to $\eps$ and therefore is more conducive to stochastic gradient computations.
@@
 We extend and simplify this approach by including the random seeds $\omega$ into the state of the Markov chain;  by keeping the random seeds fixed for several consecutive steps, the second order gradient stochasticity is greatly reduced.
-We show that this doing this produces is a valid MCMC procedure.
-This approach is not exclusive to HABC; we apply successfully to ABC-MCMC as well.
+We show that doing this produces a valid MCMC procedure.
+This approach is not exclusive to HABC; our experiments show it also helps random-walk ABC-MCMC.
+
+%apply successfully to ABC-MCMC as well.
 
@@
 In ABC, the vector of $J$ observations are typically informative statistics of the raw observations.
-It can be shown that if the statistics used in the likelihood function are sufficient, then these algorithms sample correctly from an approximation to the true posterior ([KEEP?]).
+It can be shown that if the statistics used in the likelihood function are sufficient, then these algorithms sample correctly from an approximation to the true posterior \cite{abc-sources}.
   The simulator is treated as generator of random pseudo-observations, i.e.
@@
 \begin{equation}
-  q( \thetavp, \x^{(1)'}, \ldots, \x^{(S)'} | \thetav ) =  q( \thetavp | \thetav ) \prod_s \pi( \x^{(s)'} | \thetavp)
+  q( \thetavp, \x^{(1)'}, \ldots, \x^{(S)'} | \thetav ) =  q( \thetavp | \thetav ) \prod_s \pi( \x^{(s)'} | \thetavp) \label{eq:pm-proposal}
 \end{equation}
@@
 An alternative approach to computing the ABC likelihood is to estimate the parameters of a conditional model  $\pi(\x|\thetav)$, e.g.
-kernel density estimate \cite{TurnerGenLik2014} or a Gaussian model \cite{wood2010statistical}.
+using kernel density estimate \cite{TurnerGenLik2014} or a Gaussian model \cite{wood2010statistical}.
 While either approach should be adequate and both have their own limits and advantages, for this paper we will use a Gaussian model.
@@
 For a SL log-likelihood model, we compute estimators of the first and second moments of $\pi(\x|\thetav)$.
-The advantage is that for a Gaussian $\epsilon$-kernel, we can convolve the two densities
+The advantage is that for a Gaussian $\epsvec$-kernel, we can convolve the two densities
+% \begin{eqnarray}
+%   \pi_{\epsvec}( \y | \thetav ) & = & \int \mathcal{N}( \y | \x, \eps^2 ) \mathcal{N}( \x |
+%   \mu_{\thetav}, \sigma^2_{\thetav} ) d\x \\
+%                           & = & \mathcal{N}( \y | \mu_{\thetav}, \sigma^2_{\thetav} + \eps^2 )
+% \end{eqnarray}
 \begin{eqnarray}
-  \pi_{\epsvec}( \y | \thetav ) & = & \int \mathcal{N}( \y | \x, \eps^2 ) \mathcal{N}( \x | 
+  \pi_{\epsvec}( \y | \thetav ) & = & \int \mathcal{N}( \y | \x, \epsvec^2 ) \mathcal{N}( \x | 
   \mu_{\thetav}, \sigma^2_{\thetav} ) d\x \\
-                          & = & \mathcal{N}( \y | \mu_{\thetav}, \sigma^2_{\thetav} + \eps^2 )
+                          & = & \mathcal{N}( \y | \mu_{\thetav}, \sigma^2_{\thetav} + \epsvec^2 )
 \end{eqnarray}
@@
 \begin{eqnarray}
-  \log \pi_{\epsvec}( \y | \thetav ) & = & \log \sum_s \mathcal{N}( \y | \x^{(s)}, \eps^2 ) \\
-                          & = & \log \mathcal{N}( \y | \x^{(s)}, \eps^2 ) + \log\lp 1 + H \rp \\
-                          & \approx & -\log \eps -\frac{1}{2 \eps^2} ( \y - \x^{(m)} )^2 
+  \log \pi_{\epsvec}( \y | \thetav ) & = & \log \sum_s \mathcal{N}( \y | \x^{(s)}, \epsvec^2 ) \\
+                          & = & \log \mathcal{N}( \y | \x^{(s)}, \epsvec^2 ) + \log\lp 1 + H \rp \\
+                          & \approx & -\log \epsvec -\frac{1}{2 \epsvec^2} ( \y - \x^{(m)} )^2 
 \end{eqnarray}
@@
 \begin{eqnarray}
- \log \pi_{\epsvec}( \y | \thetav ) & \approx & -\frac{1}{2} \log (\sigma^2_{\thetav} + \eps^2 ) -\frac{( \y - \mu_{\thetav} )^2 }{2 (\sigma^2_{\thetav} + \eps^2 )} 
+ \log \pi_{\epsvec}( \y | \thetav ) & \approx & -\frac{1}{2} \log (\sigma^2_{\thetav} + \epsvec^2 ) -\frac{( \y - \mu_{\thetav} )^2 }{2 (\sigma^2_{\thetav} + \epsvec^2 )} 
 \end{eqnarray}
- For the SL model, $\eps$ acts as a smoothing term and can be set to small values with little change to the log-likelihood, as long as the SL estimators are fit appropriately.
+ For the SL model, $\epsvec$ acts as a smoothing term and can be set to small values with little change to the log-likelihood, as long as the SL estimators are fit appropriately.
 This insensitivity to $\eps$ will be used in Section~\ref{sec:habc} for estimating gradients of the ABC likelihood.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-Scaling Bayesian inference algorithms to massive datasets is necessary for their existing relevance in the so-called {\em big data} era.
+Scaling Bayesian inference algorithms to massive datasets is necessary for their continuing relevance in the so-called {\em big data} era.
 We now review the role stochastic gradient methods combined with Hamiltonian dynamics have played in recent advances in scaling Bayesian inference.
-Most importantly, these methods have combined the ability of HMC to explore high-dimensional parameter spaces with the computational efficiency of using stochastic gradients based on small batches of the full dataset.
+Most importantly, these methods have combined the ability of HMC to explore high-dimensional parameter spaces with the computational efficiency of using stochastic gradients based on small mini-batches of the full dataset.
 After an overview of HMC, we will briefly describe stochastic gradient Hamiltonian dynamics (SGHDs), starting with using  Langevin dynamics \cite{welling2011bayesian}, then HMC with friction \cite{chen2014stochastic}, and finally HMC with thermostats \cite{ding2014bayesian}.
@@
 More formally, the Hamiltonian $H\lp\thetav, \rhov \rp = U(\thetav) + K(\rhov)$ is a function of the current potential energy $U(\thetav)$ and kinetic energy $K(\rhov) = \rhov^T M^{-1}\rhov/2$ ($M$ is a diagonal matrix of masses which for presentation are set to $1$).
-The potential energy is defined by negative log joint density of the data and prior:
+The potential energy is defined by the negative log joint density of the data and prior:
 \begin{equation}
@@
 \subsection{Stochastic Gradient Hamiltonian Dynamics}\label{se:sghd}
-If the log-likelihood over the full data set is replaced with a batch estimate, as is done for the following {\em stochastic gradient Hamiltonian dynamics} (SGHDs) algorithms, then the error in simulating the Hamiltonian dynamics comes not only from the discretization, but from the variance of the stochastic gradient.
-As long as this error is controlled, either by using small steps $\eta$ (SGLD), or adding friction terms $A$ (SGHMC), or using a thermostat $\xi$ (SGNHT), the expensive MH correction step can be avoided and values of $\thetav$ from the Hamiltonian dynamics can be used as samples from the posterior.
+If the log-likelihood over the full data set is replaced with a mini-batch estimate, as is done for the following {\em stochastic gradient Hamiltonian dynamics} (SGHDs) algorithms, then the error in simulating the Hamiltonian dynamics comes not only from the discretization, but from the variance of the stochastic gradient.
+As long as this error is controlled, either by using small steps $\eta$ (SGLD), or adding friction terms $A$ (SGHMC), or using a thermostat $\xi$ (SGNHT), the expensive MH correction step can be avoided and values of $\thetav$ from the Hamiltonian dynamics can be used as approximate samples from the posterior.
 
-SGHDs replace the full potential energy and its gradient with a batch approximation:
+SGHDs replace the full potential energy and its gradient with a mini-batch approximation:
 \begin{eqnarray}
@@
 \end{eqnarray} 
-where $n$ is the batch size, and $h_i$ are indices chosen randomly without replacement from $[1,N]$ (i.e. it defined a random batch).  
+where $n$ is the mini-batch size, and $h_i$ are indices chosen randomly without replacement from $[1,N]$ (i.e. it defined a random mini-batch).  
 %The main motivation for these is to avoid the expensive MH correction step while maximizing the benefit of Hamiltonian dynamics.
@@
 \end{eqnarray}
-Tt is not necessary to include $\rhov$ in the updates since there is only one step:
+It is not necessary to include $\rhov$ in the updates since there is only one step:
 \begin{eqnarray}
@@
 
- {\bf Stochastic Gradient thermostats} (SGT) \cite{ding2014bayesian} addresses the difficulty of estimating $\hat{\Bv}$ by introducing a scalar variable $\xi$ who's addition to the Hamiltonian dynamics maintains the temperature of the system constant, i.e.
-it acts as a (Nose-Hoovier) thermostat \cite{leimkuhler:2009}.
+ {\bf Stochastic Gradient thermostats} (SGNHT) \cite{ding2014bayesian} addresses the difficulty of estimating $\hat{\Bv}$ by introducing a scalar variable $\xi$ who's addition to the Hamiltonian dynamics maintains the temperature of the system constant, i.e.
+it acts as a (Nose-Hoover) thermostat \cite{leimkuhler2009metropolis}.
 The update equations remain simple: initialize $\xi = \Cv$ (or $c$), then for $t=1\ldots$
  \begin{eqnarray}
-   \rhov_{t+1} & = & \rho_t - \eta \xi_t \rhov - \eta \nabla \hat{U}(\thetav_t) + \mathcal{N}\lp \zerov, 2 \eta_t \Cv \rp \\
+   \rhov_{t+1} & = & \rhov_t - \eta \xi_t \rhov_t - \eta \nabla \hat{U}(\thetav_t) + \mathcal{N}\lp \zerov, 2 \eta_t \Cv \rp \\
    \thetav_{t+1} & = & \thetav_t + \eta \rhov_{t+1} \\
-   \xi_{t+1} & = & \xi_t + \eta \lp  \rhov_{t+1}^T \rhov_{t+1} / p - 1 \rp
+   \xi_{t+1} & = & \xi_t + \eta \lp  \rhov_{t+1}^T \rhov_{t+1} / D - 1 \rp
  \end{eqnarray}
@@
 
-\subsection{Kernel-$\eps$ versus Synthetic-likelihood -based Gradients}
+\subsection{Kernel-$\epsvec$ versus Synthetic-likelihood -based Gradients}
 In Section~\ref{sec:abc} we showed that the synthetic-likelihood representation of $\lleps(\thetav)$ is less sensitive to small choices of $\epsvec$.
 This is particularly important to HABC as our gradient approximations are proportional to differences in $\lleps(\thetav)$; if the variance of the stochastic gradients is too high, then we must choose a very small step-size $\eta$, eliminating the usefulness of HMC for ABC.
-Under the deterministic representation of $\x^{(s)}$, we can write the loglikelihoods
+Under the deterministic representation of $\x^{(s)}$, we can write the loglikelihood as
 \begin{eqnarray}
- \lleps(\thetav) & \propto & \log \sum_s \mathcal{N}( \y | f( \thetav, \omega_s), \eps^2 )\\
-                 & \approx & -\log \eps -\frac{1}{2 \eps^2} ( \y - f( \thetav, \omega_m) )^2 
+ \lleps(\thetav) & \propto & \log \sum_s \mathcal{N}( \y | f( \thetav, \omega_s), \epsvec^2 )\\
+                 & \approx & -\log \epsvec -\frac{1}{2 \epsvec^2} ( \y - f( \thetav, \omega_m) )^2 
 \end{eqnarray}
-In the second line we have again assumes $\eps$ is very small and $m$ is the index of the random seed producing the closest simulation to $\y$.
+In the second line we have assumed $\epsvec$ is very small and $m$ is the index of the random seed producing the closest simulation to $\y$.
 For a finite difference approximation, $\partial \lleps(\thetav)/\partial\thetav$ is
 \begin{equation}
- \frac{1}{4\dtheta \eps^2} \lp ( \y - f( \thetav-\dtheta, \omega_m^-) )^2 -( \y - f( \thetav + \dtheta, \omega_m^+) )^2 \rp
+ \frac{1}{4\dtheta \epsvec^2} \lp ( \y - f( \thetav-\dtheta, \omega_m^-) )^2 -( \y - f( \thetav + \dtheta, \omega_m^+) )^2 \rp
 \end{equation}
 
-On the other hand, the synthetic-likelihood is stable; using deterministic representation, 
+On the other hand, the synthetic-likelihood is stable; using a deterministic representation, 
 \begin{equation}
@@
 \end{equation}
-the gradients (for a 1-dim problem) use $\eps$ as a smoothness prior in $\partial \lleps(\thetav)/\partial\thetav$:
+the gradients (for a 1-dim problem) use $\epsvec$ as a smoothness prior in $\partial \lleps(\thetav)/\partial\thetav$:
 \begin{equation}
-  -\frac{1}{2}\log\lp\frac{\sigma_{\theta^+}^2 + \eps^2}{\sigma_{\theta^-}^2 + \eps^2}\rp -\frac{( \y - \mu_{\thetav^+} )^2 }{2 (\sigma^2_{\thetav^+} + \eps^2 )} +\frac{( \y - \mu_{\thetav^-} )^2 }{2 (\sigma^2_{\thetav^-} + \eps^2 )}
+  -\frac{1}{2}\log\lp\frac{\sigma_{\theta^+}^2 + \epsvec^2}{\sigma_{\theta^-}^2 + \epsvec^2}\rp -\frac{( \y - \mu_{\thetav^+} )^2 }{2 (\sigma^2_{\thetav^+} + \epsvec^2 )} +\frac{( \y - \mu_{\thetav^-} )^2 }{2 (\sigma^2_{\thetav^-} + \epsvec^2 )}
 \end{equation} 
+In Figure~\ref{fig:exp-varg}, as part of our demonstration of HABC, we compare the gradient approximations around the true $\thetav_{\text{MAP}}$ using SL versus kernel-$\epsvec$ Figure~\ref{fig:exp-varg} for a simple problem.
+We find that although, for this particular problem, SL has a small bias due to its Gaussian assumption, it has much smaller variance, an important property for HABC.
 
@@
 Let $L$ be the gradient-free function we wish to optimize.
-Each approximation randomly generates a {\em perturbation mask} (our name) $\Deltav$ of dimension $D=|\thetav|$ where entry $\deltad \sim \text{Bernouilli}(1/2)$ (actually, values $\pm 1$, not $0/1$).
+Each approximation randomly generates a {\em perturbation mask} (our name) $\Deltav$ of dimension $D=|\thetav|$ where entry $\Deltavd \sim 2 \text{Bernouilli}(1/2) - 1$.
 Then $L$ is evaluated at $\thetav+\dtheta \Deltav$ and $\thetav-\dtheta \Deltav$, giving the gradient approximation $\gradvest(\thetav) \approx \partial L(\thetav) / \partial \thetav$:
@@
 
-Variations of SPSA include {\em one-sided} SPSA \cite{spall2000adaptive} (we use what Spall calls 2SPSA) and an algorithm for estimating the Hessian based on the same principal as SPSA \cite{spall2005monte}.
-The one-sided version is attractive computationally, but for HABC, the updates for $\thetav$ require simulating two-sides anyway, so using 2SPSA makes more sense.
+Variations of SPSA include {\em one-sided} SPSA \cite{spall2000adaptive} (we use what Spall calls 2SPSA) and an algorithm for estimating the Hessian based on the same principle as SPSA \cite{spall2005monte}.
+The one-sided version is attractive computationally, but for HABC, the updates for $\thetav$ require simulating two-sides anyway (once at $\thetav$, after an step, and once for the one-sided gradient), so using 2SPSA makes more sense.
 SPSA has also been used within a procedure for maximum-likelihood estimation for hidden Markov models using ABC \cite{Ehrlich2013}.
@@
 Not only does kernel-$\eps$ keep a high variance for even large $S$, it seems to have a bias in its gradient estimate.
-[look into this?] }}
+[THIS IS AT POSTERIOR MEAN, SO NO BIAS FOR SL, BUT FOR KEPS.
+REPLACE with AT MAP -- THIS IS DONE] }}
 \label{fig:exp-varg}
@@
 
-\subsection{Sticky Random Numbers}
-An obvious use of common random numbers, as previously demonstrated for SPSA \cite{kleinman1999simulation}, is within the gradient approximation so that the noise from the simulator has less of an influence that $\dtheta$.
+\subsection{Common and Sticky Random Numbers}
+The usefulness of using common random numbers (CRNS) for SPSA has been previously demonstrated \cite{kleinman1999simulation}.
+In that work, the same random numbers are used to simulate both sides of the optimization function.
+This makes sense intuitively, as we would generally assume that the expected simulation function changes smoothly with $d \thetav$;  by using CRNs, the smoothness can be captured by the gradient more easily (see Figure~\ref{fig:exp-crns}).
+This is analogous to using the same mini-batch for both sides of the gradient computation, if we were to apply SPSA to Bayesian learning.
+
+In addition to using CRNs in simulations for each gradient computation, we have found that using persistent or {\em sticky} random numbers helps the HABC explore the parameter landscape more easily.
+Intuitively, for a gradient-based sampling algorithm, it means a particle can slide along smooth Hamiltonian landscape because    the additive noise is suppressed.
+This is very similar to using dependent random streams to drive MCMC \cite{Murray2012,Neal2012}, the main difference we believe is that we are using the Hamiltonian dynamics to drive proposals for $\thetav$ and using $\omegav$ suppress simulation noise.
+
+  % , is within the gradient approximation so that the noise from the simulator has less of an influence that $\dtheta$.
 In addition to this use, we have found that using the same random seeds over multiple time-steps improves the performance of HABC (TBD).
@@
 
-Our updates for $\omegav$ are simple.
+We propose including persistent random seeds $\omegav$ in the state of our Markov chain.
+As part of a Metropolis-Hastings transition operator, our idea is to randomly propose {\em flipping} each seed $\omega_s$ at time $t$ with some probability $\gamma$.
+This implies that our simulation $\x^(s)$ changes automatically (as it is deterministic).
+We argue that this is a valid transition operator using the similar arguments that can be made for pseudo-marginal ABC-MCMC which using equation~\ref{eq:pm-proposal} proposes $\thetavp$ and $\{ \x^{'(s)}\}$ jointly.
+This is equivalent to flipping all $\omegav$ {\em and} proposing $\thetavp$.
+In ABC-MCMC, one could instead propose $q(\x^{'(s)} | \thetav)$ (fixing $\thetav$) and still sample correctly (this is like rejection sampling $\x | \y, \thetav$ based on $\epsvec$).
+What we propose is slightly different.
+Again keeping $\thetav$ fixed, we propose new random seeds with some probability $\gamma$ from some uniform prior distribution over seeds.
+The Metropolis-Hastings ratio reduces to the ratio of likelihoods for observations $\y$ given the new deterministic simulation values (just like it would for ABC-MCMC).
+By keeping the random seeds fixed, we can sample $\thetav$ using HABC and use $\omegav$ as CRNs within the gradient computation step.
+
+% Our updates for $\omegav$ are simple.
 At each time step $t$, each seed is refreshed to a new random value with probability $\gamma$, i.e.
@@
 \subsection{Bias and Variance of $\nabla \hat{U}(\thetav)$}
-At $\theta_{\text{MAP}}$ we estimated the gradient $\nabla \hat{U}(\theta)$ using finite differences for both the kernel-$epsilon$ and synthetic-likelihood representation of $\pi_{eps}(y | x )$.
+[REWRITE] At $\theta_{\text{MAP}}$ we estimated the gradient $\nabla \hat{U}(\theta)$ using finite differences for both the kernel-$\epsilon$ and synthetic-likelihood representation of $\pi_{\eps}(y | x )$.
 For $S=5$ and $S=50$, $10K$ repetitions of computing the gradients was performed; there histograms are shown in Figure~\ref{fig:exp-varg}.
@@
 A further complication, which we do not address on this paper, is the forward simulation might be, and for many interesting problems is, expensive.
-We would therefore prefer to resort to some sort of surrogate modeling \cite{Meeds2014GpsUai,wilkinson:2014, michael} to avoid simulations whenever possible.
+We would therefore prefer to resort to some sort of surrogate modeling \cite{Meeds2014GpsUai,wilkinson:2014} to avoid simulations whenever possible.
 We address this scenario in Section~\ref{future}.

### 2015-03-03 draft.tex
conclusions

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{DISCUSSION} \label{sec:discussion}
+%\section{DISCUSSION} \label{sec:discussion}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+Hamiltonian ABC proposes a new set of algorithms for Bayesian inference of likelihood-free models.
+HABC builds  upon the connections between Hamilton Monte Carlo with stochastic gradients and well-established gradient approximations based on a minimal number of forward simulations, even for high-dimensional parameter spaces.
+By showing the feasibility of running ABC in high-dimensions, we hope that the door has been opened for exploration of larger simulation-based models.
 
-\begin{itemize}
-  \item New set of ABC algorithms for efficient exploration of posterior distribution.
-  \item Key: opens the door to high-dimensional ABC inference.
-  \item [Test for more accurate gradients] For ABC, the gradients we compute are stochastic because we must approximate $\pi(\x | \thetav)$ with forward simulations.
+Another innovation we introduce is the use of sticky or persistent random seeds to suppress the simulator noise and therefore smooth the simulation landscape over a local region of parameter space.
+By doing this, all our methods, but especially SGHMC and SGNHT, with their longer Hamiltonian trajectories, benefit from this smoothness that changes slowly over time.
+We feel that new classes of ABC algorithms could develop from using persistent random seeds, not solely gradient-based samplers.
+
+There are several unresolved and open questions regarding the application of stochastic gradients to ABC.
+The first issue is the importance of the bias-variance relationship for different ABC likelihood models.
+We found that using gradients based on the synthetic-likelihood greatly reduced their variance, but introduced a small bias, because of its Gaussian assumption.
+
+% \begin{itemize}
+%   \item New set of ABC algorithms for efficient exploration of posterior distribution.
+%   \item Key: opens the door to high-dimensional ABC inference.
+%   \item [Test for more accurate gradients] For ABC, the gradients we compute are stochastic because we must approximate $\pi(\x | \thetav)$ with forward simulations.
 We can make the approximation arbitrarily good by running more simulations, but this is often unnecessary for sampling, because we can make use of the uncertainty in our current approximation to decide if our sample is correct or not (cite UAI2014?).
-  \item Unbiased gradients with low noise.  SL bias might be intolerable; use KDE to get stable unbiased gradients?
-\end{itemize}
+%   \item Unbiased gradients with low noise.  SL bias might be intolerable; use KDE to get stable unbiased gradients?
+% \end{itemize}
 
-A further complication, which we do not address on this paper, is the forward simulation might be, and for many interesting problems is, expensive.
-We would therefore prefer to resort to some sort of surrogate modeling \cite{Meeds2014GpsUai,wilkinson:2014} to avoid simulations whenever possible.
-We address this scenario in Section~\ref{future}.
-Shown for relatively fast simulators.
+Expensive simulators are an important class of models that we do not address in this work.
+However, previous work in Bayesian inference has shown the usefulness of HMC-based proposals based on Gaussian process of log-likelihood surfaces \cite{rasmussen:2003}.
+We could similarly use HABC with recent ABC surrogate models \cite{Meeds2014GpsUai,wilkinson:2014} to avoid simulations whenever possible, yet benefit from Hamiltonian dynamics.
+%We address this scenario in Section~\ref{future}.  
+%Shown for relatively fast simulators.
 Extensions to surrogates with GPS (Meeds2014GpsUai, rasmussen:2003, wilkinson:2014) can use derivative of GP for likelihood as GP of gradient.

### 2015-03-03 draft.tex
variance figure

@@
 \caption{\small{Variance of gradient estimation using kernel-$\eps$ and SL for different values of $S\in\{5,50\}$ and fixed $\eps = 0.37$ (the same used in the other results).  
-For kernel-$\eps$, $\varv(\thetav_{\text{MAP}}) = 6.7 \pm 131.5$ and $\varv(\thetav_{\text{MAP}}) = 6.9 \pm 18.9$ (for $S=5$ and $S=50$, respectively).
-For SL,$\varv(\thetav_{\text{MAP}}) = 6.35 \pm 42.4$ and $\varv(\thetav_{\text{MAP}}) = 0.62 \pm 5.2$ (for $S=5$ and $S=50$, respectively).
-The ``true'' ABC gradient at $\thetav_{\text{MAP}}$ is $0.26$.
-The ``true'' ABC loglikelihood is based on a Gaussian approximation (smoothed by $\eps$), with mean and variance set by moment matching with the true Gamma distribution (the sum of $N$ exponentials).
-Not only does kernel-$\eps$ keep a high variance for even large $S$, it seems to have a bias in its gradient estimate.
-[THIS IS AT POSTERIOR MEAN, SO NO BIAS FOR SL, BUT FOR KEPS.
-REPLACE with AT MAP -- THIS IS DONE] }}
+%
+When $S=5$, the empirical estimates of  $\nabla \hat{U}(\thetav_{\text{MAP}})$ are $-12 \pm 147$ (kernel-$\eps$) and $-9.3 \pm 43$ (SL).
+When $S=50$ they are $-0.80 \pm 19$ (kernel-$\eps$) and $-7.3 \pm 4.9$ (SL).
+Note the large discrepancy in variance.
+Note the limit of $S \rightarrow \infty$,   $\nabla \hat{U}(\thetav_{\text{MAP}}) = -7.8$.
+The bias if SL gradients is due to its Gaussian approximation (smoothed by $\eps$) of $\pi(\x|\thetav)$, which is a heavy-tailed Gamma distribution (the sum of $N$ exponentials).}}
 \label{fig:exp-varg}

### 2015-03-03 draft.tex
up to blowfly done

@@
 
-\subsection{Kernel-$\epsvec$ versus Synthetic-likelihood -based Gradients}
+\subsection{Kernel-$\epsvec$ versus Synthetic-likelihood -based Gradients}\label{sec:habc-grads}
 In Section~\ref{sec:abc} we showed that the synthetic-likelihood representation of $\lleps(\thetav)$ is less sensitive to small choices of $\epsvec$.
@@
 
+Using random seeds (versus, say, a set of random numbers) allows us to treat the simulator as a black-box, setting the random seed of its RNG without knowing the internal mechanisms it uses to generate random numbers.
 We propose including persistent random seeds $\omegav$ in the state of our Markov chain.
@@
 \includegraphics[width=0.32\columnwidth]{./images/exp-SL-MCMC-posterior-hist-omega-rate-100p0-chain0.pdf}
-%\includegraphics[width=0.45\columnwidth]{./images/exp-SG-HMC-posterior-hist-omega-rate-100p0-chain0.pdf}
 \includegraphics[width=0.32\columnwidth]{./images/exp-SG-Langevin-posterior-hist-omega-rate-100p0-chain0.pdf}
-\includegraphics[width=0.32\columnwidth]{./images/exp-SG-Thermostats-posterior-hist-omega-rate-100p0-chain0.pdf}
+\includegraphics[width=0.32\columnwidth]{./images/exponential/exp3-SG-Thermostats-posterior-hist-omega-rate-100p0-chain0.pdf}
 
-\includegraphics[width=0.32\columnwidth]{./images/exp-SL-MCMC-posterior-hist-omega-rate-0p01-chain0.pdf}
-%\includegraphics[width=0.45\columnwidth]{./images/exp-SG-HMC-posterior-hist-omega-rate-0p01-chain0.pdf}
-\includegraphics[width=0.32\columnwidth]{./images/exp-SG-Langevin-posterior-hist-omega-rate-0p01-chain0.pdf}
-\includegraphics[width=0.32\columnwidth]{./images/exp-SG-Thermostats-posterior-hist-omega-rate-0p01-chain0.pdf}
+\includegraphics[width=0.32\columnwidth]{./images/exponential/exp2-SL-MCMC-posterior-hist-omega-rate-0p1-chain1.pdf}
+\includegraphics[width=0.32\columnwidth]{./images/exponential/exp2-SG-Langevin-posterior-hist-omega-rate-0p1-chain0.pdf}
+\includegraphics[width=0.32\columnwidth]{./images/exponential/exp2-SG-Thermostats-posterior-hist-omega-rate-0p1-chain1.pdf}
 
@@
 %\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
-\caption{\small{No sticky numbers}}
+\caption{\small{Posterior distributions for the demonstration problem.
+{\bf Top row:} No persistent seeds.
+{\bf Bottom row:} Persistent seeds with $\gamma=0.1$.
+Histograms of the posterior estimates are overlaid with the true posterior (dashed line).
+All algorithms (except for SG-Thermostats for non-persistent $\omegav$) give roughly the same posterior estimate.
+By adding persistent $\omegav$ SG-Thermostats achieved similar posteriors to the other algorithms.}}
 \label{fig:exp-posteriors}
@@
 
-\begin{figure*}[t]
-%\vskip 0.2in
-\setlength{\linewidth}{\textwidth}
-\setlength{\hsize}{\textwidth}
-\begin{center}
-\includegraphics[width=0.95\columnwidth]{./images/exp-SL-MCMC-theta-timeseries-omega-rate-100p0-chain0.pdf}
-\includegraphics[width=0.95\columnwidth]{./images/exp-SL-MCMC-theta-timeseries-omega-rate-0p01-chain0.pdf}
-%\includegraphics[width=0.95\columnwidth]{./images/exp-SG-HMC-theta-timeseries-omega-rate-100p0-chain0.pdf}
-\includegraphics[width=0.95\columnwidth]{./images/exp-SG-Langevin-theta-timeseries-omega-rate-100p0-chain0.pdf}
-\includegraphics[width=0.95\columnwidth]{./images/exp-SG-Langevin-theta-timeseries-omega-rate-0p01-chain0.pdf}
-\includegraphics[width=0.95\columnwidth]{./images/exp-SG-Thermostats-theta-timeseries-omega-rate-100p0-chain0.pdf}
-\includegraphics[width=0.95\columnwidth]{./images/exp-SG-Thermostats-theta-timeseries-omega-rate-0p01-chain0.pdf}
-\caption{\small{No sticky numbers  Explanation thermostats: with no sticky, need to lower eta, so it shows smoother trajectory.
-With sticky, the gradients are more consistent, allowing for a larger eta and therfore more noise injection.}}
-\label{fig:exp-theta-traces}
-\end{center}
-\vskip -0.2in
-\end{figure*}
-
 % \begin{figure*}[t]
@@
 \subsection{Bias and Variance of $\nabla \hat{U}(\thetav)$}
-[REWRITE] At $\theta_{\text{MAP}}$ we estimated the gradient $\nabla \hat{U}(\theta)$ using finite differences for both the kernel-$\epsilon$ and synthetic-likelihood representation of $\pi_{\eps}(y | x )$.
+To test our assumption that the SL likelihood is better suited for HABC, we ran FDSA at the true $\theta_{\text{MAP}}$.
+Using $S=5$ and $S=50$ and fixing $\eps = 0.37$, we gather $10K$ gradients samples using kernel-$\epsvec$ and SL likelihoods.
+These gradient estimate densities are shown in Figure~\ref{fig:exp-varg}.
+An unbiased estimate of the gradient should be centered at $0$.
+There are two important results.
+First, the SL estimates have a small bias, even at $S=50$.
+This is because it is estimating the true Gamma distribution of $\pi(\x|\thetav)$ with a Gaussian.
+We can analytically estimate this bias as $S \rightarrow \infty$; for this example it is $-7.8$ which is what SL estimates are centered around ($-9.3$ for $S=5$ and $7.3$ for $S=50$).
+The kernel-$\epsvec$ likelihood has very low bias at $S=50$ (but large at $S=5$, $-12$).
+However, the second important result is the variances.
+SL variances decrease quickly with $S$: $\sigma^2 = 43^2 \rightarrow 4.9^2$, whereas kernel-$\epsvec$ starts very high and remains much higher: $\sigma^2 = 147^2 \rightarrow 19^2$.
+It is for this reason that we have chosen to use SL likelihoods for our gradient estimates, despite their small bias.
+As mentioned in Section~\ref{sec:habc-grads} it is possible that other likelihood models, such as KDE, might provide low bias and low variance gradient estimates.
+We leave this for future work.
+%
+%
+% [REWRITE] At $\theta_{\text{MAP}}$ we estimated the gradient $\nabla \hat{U}(\theta)$ using finite differences for both the kernel-$\epsilon$ and synthetic-likelihood representation of $\pi_{\eps}(y | x )$.
 For $S=5$ and $S=50$, $10K$ repetitions of computing the gradients was performed; there histograms are shown in Figure~\ref{fig:exp-varg}.
@@
 The bias in the SL representation comes from the Gaussian approximation $\pi(y|\theta)$ which we can compute analytically using the CLT as $\text{N}( 1/\theta, 1/N \theta^2 + \eps^2)$.
-(We believe this bias is due to the normal approximation assuming a symmetric mode as the bias goes to zero at the mean of the true posterior).
+% (We believe this bias is due to the normal approximation assuming a symmetric mode as the bias goes to zero at the mean of the true posterior).
 Note that this bias is directional signal only appears nears the mode.
@@
 
+\begin{figure*}[t]
+%\vskip 0.2in
+\setlength{\linewidth}{\textwidth}
+\setlength{\hsize}{\textwidth}
+\begin{center}
+\includegraphics[width=0.95\columnwidth]{./images/exp-SL-MCMC-theta-timeseries-omega-rate-100p0-chain0.pdf}
+\includegraphics[width=0.95\columnwidth]{./images/exponential/exp2-SL-MCMC-theta-timeseries-omega-rate-0p1-chain3.pdf}
+%\includegraphics[width=0.95\columnwidth]{./images/exp-SG-HMC-theta-timeseries-omega-rate-100p0-chain0.pdf}
+\includegraphics[width=0.95\columnwidth]{./images/exp-SG-Langevin-theta-timeseries-omega-rate-100p0-chain0.pdf}
+\includegraphics[width=0.95\columnwidth]{./images/exponential/exp2-SG-Langevin-theta-timeseries-omega-rate-0p1-chain3.pdf}
+\includegraphics[width=0.95\columnwidth]{./images/exponential/exp3-SG-Thermostats-theta-timeseries-omega-rate-100p0-chain0.pdf}
+\includegraphics[width=0.95\columnwidth]{./images/exponential/exp2-SG-Thermostats-theta-timeseries-omega-rate-0p1-chain3.pdf}
+\caption{\small{Last $1000$ $\thetav$ trajectories for the demonstration problem.
+{\bf Left column:} Non-persistent random seeds.
+{\bf Right column:} Persistent random seeds with $\gamma = 0.1$.
+Each algorithm's parameters were optimized to minimize the total variational distance.
+With persistent seeds, each algorithm's random walk behavior is suppressed.
+Without persistent seeds, SG-Thermostats chooses a small step-size $\eta$, resulting in an under-dispersed estimate of the posterior; when the seeds are persistent, the gradients are more consistent, allowing for a larger $\eta$ and therefore more injected noise.
+The resulting posteriors are shown in Figure~\ref{fig:exp-posteriors}.
+%For SG-Thermostats, persistent $\omegav$ allows for larger $\eta$.Sticky  Explanation thermostats: with no %sticky, need to lower eta, so it shows smoother trajectory.
+With sticky, the gradients are more consistent, %allowing for a larger eta and therefore more noise injection.
+}}
+\label{fig:exp-theta-traces}
+\end{center}
+\vskip -0.2in
+\end{figure*}
+
 \subsection{Posterior Inference using HABC}
-In Figures~\ref{fig:exp-theta-traces} we show the posterior trace plots for ABC-MCMC, SGLD, SGHMC, and SGNHT versions of HABC using SL gradient estimates ($S=5$).
+We ran chains of length $50K$ for SL-MCMC, SGLD, SGHMC, and SGNHT versions of HABC using SL gradient estimates ($S=5$).
+A pseudo-marginal version of SL-MCMC was used.
+We note that SGHMC gave results nearly identical to SGNHT, so are not shown do to space limitations.
+In one set of experiments, common random seeds were used for gradient computations only, and did not persist over time steps; these experiments are called {\em non-persistent}.
+In another set of runs, we resampled $\omega_s$ at each time step with probability $\gamma = 0.1$; these experiments are {\em persistent}.
+In Figure~\ref{fig:exp-posteriors} we show the posterior distributions for these experiments; in Table~\ref{tab:exp-posterior} we report the {\em total variational distance} between the true posterior and the ABC posteriors using the first $10K$ samples and after $50K$ samples (averaged over $5$ chains).
+Of note is the poor approximation of SG-Thermostats when the seeds are not persistent.
+By adding persistent seeds, SG-Thermostats gives similar posteriors to the other methods.
+
+In Figure~\ref{fig:exp-theta-traces} we show the trace plots of the last $1000$ samples from a single chain for each algorithm.
+In the left column, traces for non-persistent random seeds are shown, and on the right, traces for persistent seeds.
+We can observe that persistent random seeds further reduces the random walk behavior of all three methods.
+We also observe small improvements in total variational distance for SL-MCMC and SGLD, while SGNHT improves significantly.
+%the random seeds were changed at each time-step In Figure~\ref{fig:exp-theta-traces} we show the posterior trace plots for ABC-MCMC, SGLD, SGHMC, and SGNHT versions of HABC using SL gradient estimates ($S=5$).
 The behavior of the trajectories is quite different, though they all produce the same correct posterior estimates (Figure~\ref{fig:exp-posteriors}).
-In Table~\ref{tab:exp-posteriors-nosticky} we report the {\em total variational distance} between the true posterior and the ABC posteriors after different numbers of samples (repeated 5 times).
+In Table~\ref{tab:exp-posterior} we report the {\em total variational distance} between the true posterior and the ABC posteriors after different numbers of samples (repeated 5 times).
 
@@
 \begin{table}[h]
-\caption{Posterior inference for Exponential problem}
+\caption{Average total variational distance (tvd) for the demonstration problem.
+{\em Non-persistent} used no persistent random seeds, whereas {\em Persistent} randomly proposes a new $\omega_s$ with $\gamma=0.1$.
+Each algorithms' parameters were optimized for minimal tvd after $10K$ samples.
+The results for SGHMC (not shown) and SGNHT are nearly identical.
+All algorithms tend to benefit from persistent seeds, in particular SGHMC and SGNHT, the methods with long Hamiltonian trajectories.}
 \label{tab:exp-posterior}
 \begin{center}
-\begin{tabular}{|l||c|c|c|c|c}
-  \hline 
-Algo & $1000$ & $10000$ & $50000$ \\ \hline \hline 
-SL-ABC & $0.099$ & $0.047$ & $0.045$ \\
-SGLD & $0.082$ & $0.049$ & $0.048$ \\
-SGNHT & $0.271$ & $0.162$ & $0.161$ \\
-SGNHT* & $0.144$ & $0.056$ & $0.075$ \\ \hline 
+\begin{tabular}{l|c|c||c|c|}
+  \cline{2-5}
+% Algo & $1000$ & $10000$ & $50000$ \\ \hline \hline
+% SL-ABC & $0.099$ & $0.047$ & $0.045$ \\
+% SGLD & $0.082$ & $0.049$ & $0.048$ \\
+% SGNHT & $0.271$ & $0.162$ & $0.161$ \\
+% SGNHT* & $0.144$ & $0.056$ & $0.075$ \\ \hline
+ & \multicolumn{2}{|c||}{Non-persistent} & \multicolumn{2}{c|}{Persistent} \\
+ \cline{1-5} 
+\multicolumn{1}{|l|}{Algo} & $10K$ & $50K$ & $10K$ & $50K$ \\ \hline \hline 
+\multicolumn{1}{|l|}{SL-ABC} & $0.047$ & $0.045$ & $0.045$ & $0.045$ \\
+\multicolumn{1}{|l|}{SGLD} & $0.049$ & $0.048$ & $0.048$ & $0.043$ \\
+%\multicolumn{1}{|l|}{SGHMC} & $0.170$ & $0.165$ & $0.052$ & $0.047$ \\
+\multicolumn{1}{|l|}{SGNHT} & $0.232$ & $0.239$ & $0.055$ & $0.051$ \\\hline 
 \end{tabular}
@@
 \end{table}
-
 % \begin{table}[h]
@@
 
-\subsection{Sticky Random Numbers}
-Finally we compare posterior estimates using sticky CRNs.
+% \subsection{Sticky Random Numbers}
+% Finally we compare posterior estimates using sticky CRNs.
 Using a flip rate of $1\%$, i.e.
@@
 
+\begin{figure}[t]
+\vskip 0.2in
+\begin{center}
+\includegraphics[width=1.1\columnwidth]{./images/blowfly/bf_theta_thermo_timeseries.png}
+%\includegraphics[width=0.45\columnwidth]{./images/exp-ABC-MCMC-posterior_hist.pdf}
+%\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
+%\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Langevin-posterior_hist.pdf}
+%\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
+\caption{\small{}}
+\label{fig:exp-posteriors}
+\end{center}
+\vskip -0.2in
+\end{figure}
+
 \subsection{Bayesian Autoencoders}\label{sec:auto}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{CONCLUSION} \label{sec:conclusion}
+\section{DISCUSSION AND CONCLUSION} \label{sec:conclusion}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 We found that using gradients based on the synthetic-likelihood greatly reduced their variance, but introduced a small bias, because of its Gaussian assumption.
+The second issue is setting algorithm parameters, in particular the step-sizes $\eta$, the injected noise $C$ (for SGHMC/SGNHT), and the number of SPSA repetitions $R$.
+All of these parameters are highly interactive.
+Can we use statistical tests during the MCMC run to determine $R$?
+Are should $\eta$ and $C$ be set differently in the ABC setting?
+One final issue is monitoring or determining whether the correct amount of noise is being injected to ensure proper sampling.
+In SGLD, for example, we can always turn down $\eta$ so that the injected noise term dominates, but when our goal is efficient exploration of the posterior, this is a not very satisfying solution.
 

### 2015-03-03 scripts_for_figures/exp_problem_tables.py

@@ -3,16 +3,17 @@ import pylab as pp
 import scipy as sp
 
 results_dir = "../../abc_sgld/code/exponential_problem/images/"
+results_dir2 = "../../abc_sgld_exponential/"
 
 algonames = ["SL-ABC", "SGLD", "SGHMC", "SGNHT"]
 
 times = np.loadtxt( results_dir+"exp-SL-MCMC-tvd-times-omega-rate-0p01.txt" )
 times = times.astype(int)
 sticky_files = []
-sticky_files.append( results_dir+"exp-SL-MCMC-tvd-omega-rate-0p01.txt" )
-sticky_files.append( results_dir+"exp-SG-Langevin-tvd-omega-rate-0p5.txt" )
-sticky_files.append( results_dir+"exp-SG-HMC-tvd-omega-rate-0p01.txt" )
-sticky_files.append( results_dir+"exp-SG-Thermostats-tvd-omega-rate-0p01.txt" )
+sticky_files.append( results_dir2+"exp2-SL-MCMC-tvd-omega-rate-0p1.txt" )
+sticky_files.append( results_dir2+"exp2-SG-Langevin-tvd-omega-rate-0p1.txt" )
+sticky_files.append( results_dir2+"exp2-SG-HMC-tvd-omega-rate-0p1.txt" )
+sticky_files.append( results_dir2+"exp2-SG-Thermostats-tvd-omega-rate-0p1.txt" )
 
 
 non_sticky_files = []
@@ -22,7 +23,7 @@ non_sticky_files.append( results_dir+"exp-SG-HMC-tvd-omega-rate-100p0.txt" )
 non_sticky_files.append( results_dir+"exp-SG-Thermostats-tvd-omega-rate-100p0.txt" )
 
 print times
-time_ids = [2,4,8]
+time_ids = [4,8]
 
 for algo, fn in zip( algonames, sticky_files ):
   tdv = np.loadtxt( fn )
@@ -39,28 +40,62 @@ for algo, fn in zip( algonames, non_sticky_files ):
   for idx in time_ids:
     print "%s at %08d has %3.4f tdv"%(algo,times[idx],mn[idx])
 
-algonames = ["SL-ABC", "SGLD", "SGNHT", "SGNHT*"]
+algonames = ["SL-ABC", "SGLD", "SGHMC", "SGNHT"]
 files2use = []
 files2use.append( results_dir+"exp-SL-MCMC-tvd-omega-rate-100p0.txt" )
[…]

### 2015-03-03 draft.tex
major edits and additions

@@
 In ABC, the vector of $J$ observations are typically informative statistics of the raw observations.
-It can be shown that if the statistics used in the likelihood function are sufficient, then these algorithms sample correctly from an approximation to the true posterior \cite{abc-sources}.
+It can be shown that if the statistics used in the likelihood function are sufficient, then these algorithms sample correctly from an approximation to the true posterior.
   The simulator is treated as generator of random pseudo-observations, i.e.
@@
 As part of a Metropolis-Hastings transition operator, our idea is to randomly propose {\em flipping} each seed $\omega_s$ at time $t$ with some probability $\gamma$.
-This implies that our simulation $\x^(s)$ changes automatically (as it is deterministic).
+
+Consider a Metropolis-Hastings transition where we condition on $\thetav$ and propose changing the random seeds.
+It is straightforward to show that this leaves the target distribution invariant.
+Let $\pi(\omega_s)$ be a uniform prior over random seeds.
+We propose $\omega_s^{'} \sim q(\omega_s^{'} | \omega_s) = \pi(\omega_s)$ (independent of the current seed).
+Pseudo-data is simulated deterministically $\x^{'(s)} = f( \thetav, \omega_s^{'})$.
+Our proposal distribution has probability $\pi(x^{'(s)} | \thetav, \omega_s^{'}) = \delta( x^{'(s)} - f( \thetav, \omega_s^{'}))$.
+The acceptance probability ratio reduces to a ratio of likelihoods [TODO]:
+% \begin{equation}
+%   \frac{\pi(\omega^{'})\pi(\y,\x^{'})\pi(\x^{'} | \thetav, \omega_{'})}{\pi(\omega)\pi(\y,\x)\pi(\x | \thetav, \omega)}  \rp
+% % \min\lp 1, \frac{\pi(\y,\x^{'})}{\pi(\y,\x)}  \rp
+% \end{equation}
+
+[TODO: TIE THIS IN] This implies that our simulation $\x^(s)$ changes automatically (as it is deterministic).
 We argue that this is a valid transition operator using the similar arguments that can be made for pseudo-marginal ABC-MCMC which using equation~\ref{eq:pm-proposal} proposes $\thetavp$ and $\{ \x^{'(s)}\}$ jointly.
@@
 We have explicitly shown the smoothness of the simulator by generating data along trajectories of fixed seeds $\omega_s$; i.e.
-for several $\omega_s$ we vary $\theta$ (dashed lines are function $f(\theta, \omega_s)$) and randomly reveal data (blue circles).
+for several $\omega_s$ we vary $\theta$ (dashed lines are function $f(\theta, \omega_s)$) and randomly reveal simulation data (blue circles).
 The horizontal line with shading indicates $y \pm 2 \eps$, where $\eps = 0.37$ is used throughout the demonstration.
@@
 \subsection{Bias and Variance of $\nabla \hat{U}(\thetav)$}
-To test our assumption that the SL likelihood is better suited for HABC, we ran FDSA at the true $\theta_{\text{MAP}}$.
+To test our assumption that the synthetic-likelihood model is better suited for HABC, we ran FDSA at the true $\theta_{\text{MAP}}$.
 Using $S=5$ and $S=50$ and fixing $\eps = 0.37$, we gather $10K$ gradients samples using kernel-$\epsvec$ and SL likelihoods.
@@
 We can analytically estimate this bias as $S \rightarrow \infty$; for this example it is $-7.8$ which is what SL estimates are centered around ($-9.3$ for $S=5$ and $7.3$ for $S=50$).
-The kernel-$\epsvec$ likelihood has very low bias at $S=50$ (but large at $S=5$, $-12$).
+The kernel-$\epsvec$ likelihood, on the other hand, exhibits, very low bias at $S=50$.
 However, the second important result is the variances.
@@
 \begin{center}  
-\includegraphics[width=0.95\columnwidth]{./images/exp-SL-MCMC-theta-timeseries-omega-rate-100p0-chain0.pdf}
-\includegraphics[width=0.95\columnwidth]{./images/exponential/exp2-SL-MCMC-theta-timeseries-omega-rate-0p1-chain3.pdf}
-%\includegraphics[width=0.95\columnwidth]{./images/exp-SG-HMC-theta-timeseries-omega-rate-100p0-chain0.pdf}
-\includegraphics[width=0.95\columnwidth]{./images/exp-SG-Langevin-theta-timeseries-omega-rate-100p0-chain0.pdf}
-\includegraphics[width=0.95\columnwidth]{./images/exponential/exp2-SG-Langevin-theta-timeseries-omega-rate-0p1-chain3.pdf}
-\includegraphics[width=0.95\columnwidth]{./images/exponential/exp3-SG-Thermostats-theta-timeseries-omega-rate-100p0-chain0.pdf}
-\includegraphics[width=0.95\columnwidth]{./images/exponential/exp2-SG-Thermostats-theta-timeseries-omega-rate-0p1-chain3.pdf}
-\caption{\small{Last $1000$ $\thetav$ trajectories for the demonstration problem.
-{\bf Left column:} Non-persistent random seeds.
-{\bf Right column:} Persistent random seeds with $\gamma = 0.1$.
+\includegraphics[width=0.65\columnwidth]{./images/exp-SL-MCMC-theta-timeseries-omega-rate-100p0-chain0.pdf}
+\includegraphics[width=0.65\columnwidth]{./images/exp-SG-Langevin-theta-timeseries-omega-rate-100p0-chain0.pdf}
+\includegraphics[width=0.65\columnwidth]{./images/exponential/exp3-SG-Thermostats-theta-timeseries-omega-rate-100p0-chain0.pdf}
+\includegraphics[width=0.65\columnwidth]{./images/exponential/exp2-SL-MCMC-theta-timeseries-omega-rate-0p1-chain3.pdf}
+\includegraphics[width=0.65\columnwidth]{./images/exponential/exp2-SG-Langevin-theta-timeseries-omega-rate-0p1-chain3.pdf}
+\includegraphics[width=0.65\columnwidth]{./images/exponential/exp2-SG-Thermostats-theta-timeseries-omega-rate-0p1-chain3.pdf}
+\caption{\small{Trajectories of the last $1000$ $\thetav$ samples for the demonstration problem.
+{\bf Top row:} Non-persistent random seeds.
+{\bf Bottom row:} Persistent random seeds with $\gamma = 0.1$.
 Each algorithm's parameters were optimized to minimize the total variational distance.
 With persistent seeds, each algorithm's random walk behavior is suppressed.
-Without persistent seeds, SG-Thermostats chooses a small step-size $\eta$, resulting in an under-dispersed estimate of the posterior; when the seeds are persistent, the gradients are more consistent, allowing for a larger $\eta$ and therefore more injected noise.
+Without persistent seeds, the optimal step-size $\eta$ for SG-Thermostats is small, resulting in an under-dispersed estimate of the posterior; when the seeds are persistent, the gradients are more consistent, and the optimal step-size is larger and therefore there is larger injected noise.
 The resulting posteriors are shown in Figure~\ref{fig:exp-posteriors}.
@@
 We also observe small improvements in total variational distance for SL-MCMC and SGLD, while SGNHT improves significantly.
+We find this a compelling mystery.
+Is it because of the interaction between hyperparameters and stochastic gradients, or is this an artifact of this simple model?
 %the random seeds were changed at each time-step In Figure~\ref{fig:exp-theta-traces} we show the posterior trace plots for ABC-MCMC, SGLD, SGHMC, and SGNHT versions of HABC using SL gradient estimates ($S=5$).
@@
 We present experimental results comparing HABC with standard ABC-MCMC for two challenging simulators.
-The first is low-dimensional ($D=6$), but exhibits chaotic behavior.
-For the second problem we apply the techniques of HABC to a Bayesian image classifier and autoencoder.
+The first is the {\em blowfly} model which uses stochastic differential equations to model possibly chaotic population dynamics \cite{wood2010statistical}.
+Although it is a low-dimensional problem, the noise and chaotic behavior of the model make it challenging for gradient-based sampling.
+Our second experiment applies HABC to a Bayesian logistic regression model.
+Although we only use $2$ classes ($0$'s versus $1$'s), the dimensionality is very high ($D=1568$).
+We show that HABC can work well despite using SPSA its gradients.
+ %
+ % The first is low-dimensional ($D=6$), but exhibits chaotic behavior.
+Though low-dimensional,  For the second problem we apply the techniques of HABC to a Bayesian image classifier and autoencoder.
 The benefit of this problem is we can compare directly with the SGHD algorithms using the analytic gradients.
-This problem is has very high-dimension ($D>15000$), yet we are able to successfully apply SPSA-based gradients to HABC.
+This problem is has very high-dimension ($D=1568$), yet we are able to successfully apply SPSA-based gradients to HABC.
 
@@
 We use $\epsvec = \{ 1/2,1/2,1/2,1/2, 1/4,1/4,1/4,1/4,3/4,3/4 \}$ (slightly different $\epsvec$ from \cite{Meeds2014GpsUai}) and $S=10$ for all experiments.
-We use SPSA with $R=1$ using SL log-likelihoods for all HABC gradient estimates.
-Note that this requires $2S$ forward simulations per time-step, the same as marginal ABC and twice pseudo-marginal ABC.
+We use SPSA with $R=2$ using SL log-likelihoods for all HABC gradient estimates.
+Without persistent seeds, the number of simulations per time-step is $2 S R$ (about double marginal ABC-MCMC) and with it is $2SR+2S\gamma$.
 
-Figure~\ref{fig:bf-posterior-histograms} show the posterior distributions for $\log P, \log \delta$ and the posterior predictive distributions of [FILL IN].
-Traces of $\thetav$ are shown for SGHMC and SGNHT in Figure~\ref{}.
+Figure~\ref{fig:bf-results} show the posterior distributions for three parameters for SL-MCMC, SGLD, and SG-Thermostats using non-persistent seeds (persistent seeds, not shown, produced very similar posteriors).
+In the second row we show the trajectories of two parameters, clearly showing the suppressed random walk behavior of SGLD and SG-Thermostats relative to ABC-MCMC.
+In Figure~\ref{fig:bf-two-d-theta} the scatter plots of trajectories are shown for two parameters.
+
+  % Traces of $\thetav$ are shown for SGHMC and SGNHT in Figure~\ref{}.
 Finally, we compare the convergence to $\y$ using the online posterior predictive distribution in Figure~\ref{}.
 By using stick random numbers, the convergence is [FILL IN].
+
+  \begin{figure*}[ht!]
+  %\vskip 0.2in
+  \setlength{\linewidth}{\textwidth}
+  \setlength{\hsize}{\textwidth}
+  \begin{center}
+    \includegraphics[width=0.65\columnwidth]{./images/blowfly/non-sticky-theta-hist-SL-MCMC.pdf}
+    \includegraphics[width=0.65\columnwidth]{./images/blowfly/non-sticky-theta-hist-SG-Langevin.pdf}
+    \includegraphics[width=0.65\columnwidth]{./images/blowfly/non-sticky-theta-hist-SG-Thermostats.pdf}
+    \includegraphics[width=0.65\columnwidth]{./images/blowfly/non-sticky-theta-1d-SL-MCMC.pdf}
+    \includegraphics[width=0.65\columnwidth]{./images/blowfly/non-sticky-theta-1d-SG-Langevin.pdf}
+    \includegraphics[width=0.65\columnwidth]{./images/blowfly/non-sticky-theta-1d-SG-Thermo.pdf}
+   \caption{\small{Blowfly posterior distributions (non-persistent seeds).
+{\bf Top row}:  Posteriors for three parameters for SL-MCMC (left set of three), SGLD (middle), and SG-Thermostats (right).
+{\bf Bottom row:} Last trajectories of the last 1000 samples for two parameters for the same algorithms.
+  }}
+  \label{fig:bf-results}
+  \end{center}
+  \vskip -0.2in
+  \end{figure*}
+
+% \begin{figure}[t]
+% \vskip 0.2in
+% \begin{center}
+% \includegraphics[width=0.9\columnwidth]{./images/blowfly/non-sticky-theta-hist-SL-MCMC.pdf}
+% \includegraphics[width=0.9\columnwidth]{./images/blowfly/non-sticky-theta-hist-SG-Langevin.pdf}
+% \includegraphics[width=0.9\columnwidth]{./images/blowfly/non-sticky-theta-hist-SG-Thermostats.pdf}
+% \caption{\small{}}
+% \label{fig:bf-posteriors-no-sticky}
+% \end{center}
+% \vskip -0.2in
+% \end{figure}
+
+% \begin{figure}[t]
+% \vskip 0.2in
+% \begin{center}
+% \includegraphics[width=0.9\columnwidth]{./images/blowfly/sticky-theta-hist-SL-MCMC.pdf}
+% \includegraphics[width=0.9\columnwidth]{./images/blowfly/sticky-theta-hist-SG-Langevin.pdf}
+% \includegraphics[width=0.9\columnwidth]{./images/blowfly/sticky-theta-hist-SG-Thermostats.pdf}
+% \caption{\small{}}
+% \label{fig:bf-posteriors-sticky}
+% \end{center}
+% \vskip -0.2in
+% \end{figure}
+%
+% \begin{figure}[t]
+% \vskip 0.2in
+% \begin{center}
+% \includegraphics[width=0.9\columnwidth]{./images/blowfly/non-sticky-theta-1d-SL-MCMC.pdf}
+% \includegraphics[width=0.9\columnwidth]{./images/blowfly/non-sticky-theta-1d-SG-Langevin.pdf}
+% %\includegraphics[width=0.9\columnwidth]{./images/blowfly/sticky-theta-1d-SG-Langevin.pdf}
+% \includegraphics[width=0.9\columnwidth]{./images/blowfly/non-sticky-theta-1d-SG-Thermostats.pdf}
+% %\includegraphics[width=0.9\columnwidth]{./images/blowfly/sticky-theta-1d-SG-Thermostats.pdf}
+% \caption{\small{}}
+% \label{fig:bf-one-d-theta}
+% \end{center}
+% \vskip -0.2in
+% \end{figure}
 
@@
 \begin{center}
-\includegraphics[width=1.1\columnwidth]{./images/blowfly/bf_theta_thermo_timeseries.png}
-%\includegraphics[width=0.45\columnwidth]{./images/exp-ABC-MCMC-posterior_hist.pdf}
-%\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
-%\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Langevin-posterior_hist.pdf}
-%\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
-\caption{\small{}}
-\label{fig:exp-posteriors}
+%\includegraphics[width=0.30\columnwidth]{./images/blowfly/non-sticky-theta-2d-SL-MCMC.pdf}
+\includegraphics[width=0.75\columnwidth]{./images/blowfly/non-sticky-theta-2d-SG-Langevin.pdf}
+%\includegraphics[width=0.35\columnwidth]{./images/blowfly/sticky-theta-2d-SG-Langevin.pdf}
+\includegraphics[width=0.75\columnwidth]{./images/blowfly/non-sticky-theta-2d-SG-Thermo.pdf}
+%\includegraphics[width=0.35\columnwidth]{./images/blowfly/sticky-theta-2d-SG-Thermo.pdf}
+\caption{\small{Trajectories of two parameters over the last 1000 time-steps.
+{\bf Top:} SGLD and {\bf Bottom}: SG-Thermostats.
+Relative to SL-MCMC (not shown), the Hamiltonian dynamics clearly show persistent $\thetav$ trajectories.}}
+\label{fig:bf-two-d-theta}
 \end{center}
@@
 Can we use statistical tests during the MCMC run to determine $R$?
-Are should $\eta$ and $C$ be set differently in the ABC setting?
+Should $\eta$ and $C$ be set differently in the ABC setting?
 One final issue is monitoring or determining whether the correct amount of noise is being injected to ensure proper sampling.

### 2015-03-03 scripts_for_figures/bf_gather_theta.py

@@ -1,0 +1,50 @@
+import numpy as np
+import pylab as pp
+import scipy as sp
+import cPickle
+
+def extract_from_pkl( pkl_file, algo_key, save_name ):
+  print "loading %s"%(pkl_file)
+  r = cPickle.load( open( pkl_file ))
+  
+  nbr_chains = len(r[algo_key]["results"])
+  
+  np.savetxt("%s-%s-mean-errors.txt"%(save_name, algo_key), r[algo_key]['mean_errors'])
+  for chain_id in range(nbr_chains):
+    theta = r[algo_key]["results"][chain_id]["THETA"]
+    
+    np.save( "%s-%s-theta-chain%d.npy"%(save_name, algo_key,chain_id), theta )
+    
+    print save_name, algo_key, chain_id, theta.shape
+  
+
+results_dir = "../../abc_sgld/code/exponential_problem/images/"
+results_dir2 = "../../abc_sgld_blowfly/"
+
+algonames = ["SL-ABC", "SGLD", "SGHMC", "SGNHT"]
+
+SGLD_no_sticky   = "bf-no-sticky-SG-Langevin-results.pkl"
+SGLD_sticky   = "bf-sticky-SG-Langevin-results.pkl"
+
+SGNHT_no_sticky  = "bf-no-sticky-SG-Thermostats-results.pkl"
+SGNHT_sticky2  = "bf-sticky2-SG-Thermostats-results.pkl"
+SGNHT_no_sticky2 = "bf-no-sticky2-SG-Thermostats-results.pkl"
+SGNHT_sticky5  = "bf-sticky5-SG-Thermostats-results.pkl"
+
+SLMCMC_no_sticky = "bf-no-sticky-SL-MCMC-results.pkl"
+SLMCMC_sticky = "bf-sticky-SL-MCMC-results.pkl"
+
+times = np.loadtxt( results_dir+"exp-SL-MCMC-tvd-times-omega-rate-0p01.txt" )
+
+if __name__ == "__main__":
+  
+  
+  #r = extract_from_pkl( results_dir2 + SLMCMC_no_sticky, "SL-MCMC", "bf-no-sticky" )
+  #r = extract_from_pkl( results_dir2 + SLMCMC_sticky, "SL-MCMC", "bf-sticky" )
+  #r = extract_from_pkl( results_dir2 + SGNHT_no_sticky, "SG-Thermostats", "bf-no-sticky" )
[…]

### 2015-03-03 scripts_for_figures/bf_histogram_thetas.py

@@ -1,0 +1,137 @@
+import numpy as np
+import pylab as pp
+import scipy as sp
+from scipy import stats as spstats
+from abc_sgld.code.blowfly_problem.blowfly_wrapper import *
+
+def view_theta_hist( problem, theta_range, samples, algoname, burnin ):
+  figsize = (10,12)
+  alpha=0.75
+  pp.rc('text', usetex=True)
+  pp.rc('font', family='times')
+  f = pp.figure( figsize=figsize )
+  sp=pp.subplot(1,1,1)
+      
+  #theta0=theta_range[0]
+  #thetaN=theta_range[-1]
+  theta_names = [r'\text{log} $P$', r'\text{log} ${\delta}$', r'\text{log} ${N_0}$', r'\text{log} ${\sigma_d}$', r'\text{log} ${\sigma_p}$', r'$\tau$']
+  #pp.hist( samples, problem.p.nbins_coarse, range=problem.p.range,normed = True, alpha = alpha )
+  for i in range(6):
+    sp=pp.subplot(3,2,i+1)
+    pp.hist( samples[burnin:,i], 20, normed=True, alpha=0.75)
+    #pp.plot( samples[-howmany:,i], 'ko')
+    pp.xlabel( theta_names[i], rotation='horizontal')
+    # set_tick_fonsize( sp, 16 )
+    # set_title_fonsize( sp, 32 )
+    # set_label_fonsize( sp, 24 )
+  
+  #pp.ylim(theta0,thetaN)
+  #pp.xlabel( "theta")
+  #pp.ylabel( "x")
+  
+  #pp.xlabel( r'time' )
+  #pp.ylabel( r'${\bm \theta}$', rotation='horizontal' )
+  
+  pp.legend( [algoname], loc=0,fancybox=True,prop={'size':32} )
+  
+def view_theta_hist_short( problem, theta_range, samples, algoname, burnin, ids, color, note = "" ):
+  figsize = (10,4)
+  alpha=0.75
+  pp.rc('text', usetex=True)
+  pp.rc('font', family='times')
+  f = pp.figure( figsize=figsize )
+  sp=pp.subplot(1,1,1)
+      
+  #theta0=theta_range[0]
+  #thetaN=theta_range[-1]
[…]

### 2015-03-03 scripts_for_figures/bf_trajectory_theta.py

@@ -1,0 +1,217 @@
+import numpy as np
+import pylab as pp
+import scipy as sp
+from scipy import stats as spstats
+from abc_sgld.code.blowfly_problem.blowfly_wrapper import *
+
+def view_theta_timeseries_twod( problem, theta, algoname, nbr, ids, colors, note="" ):
+  figsize = (12,8)
+  alpha=0.75
+  pp.rc('text', usetex=True)
+  pp.rc('font', family='times')
+  f = pp.figure( figsize=figsize )
+  sp=pp.subplot(1,1,1)
+      
+  #theta0=theta_range[0]
+  #thetaN=theta_range[-1]
+  theta_names = [r'\text{log} $P$', r'{log} ${\delta}$', r'\text{log} ${N_0}$', r'{log} ${\sigma_d}$', r'\text{log} ${\sigma_p}$', r'$\tau$']
+  #pp.hist( samples, problem.p.nbins_coarse, range=problem.p.range,normed = True, alpha = alpha )
+  lgnames = []
+  
+  pp.plot( theta[-nbr:,ids[0]], theta[-nbr:,ids[1]], 'k-', ms=5,lw=6, alpha=0.75)
+  pp.plot( theta[-nbr:,ids[0]], theta[-nbr:,ids[1]], 'ro', ms=15, alpha=0.5)
+  pp.ylabel( r'%s'%(theta_names[ids[1]]) , rotation='vertical' )
+  pp.xlabel( r'%s'%(theta_names[ids[0]]) )
+  
+  #lgnames.append( r"%s  %s"%(algoname,theta_names[idx]))
+  pp.grid('on')
+  #pp.xlim( rg[0],rg[1])
+  set_tick_fonsize( sp, 32 )
+  set_title_fonsize( sp, 32 )
+  set_label_fonsize( sp, 32 )
+  
+  pp.legend( [algoname], loc='best',fancybox=True,prop={'size':32} )
+  
+  
+def view_theta_timeseries_short( problem, theta, algoname, nbr, ids, colors, note="" ):
+  figsize = (16,6)
+  alpha=0.75
+  pp.rc('text', usetex=True)
+  pp.rc('font', family='times')
+  f = pp.figure( figsize=figsize )
+  sp=pp.subplot(1,1,1)
+      
+  #theta0=theta_range[0]
+  #thetaN=theta_range[-1]
[…]

### 2015-03-03 draft.tex
edits

@@
 \subsection{Common and Sticky Random Numbers}
-The usefulness of using common random numbers (CRNS) for SPSA has been previously demonstrated \cite{kleinman1999simulation}.
-In that work, the same random numbers are used to simulate both sides of the optimization function.
-This makes sense intuitively, as we would generally assume that the expected simulation function changes smoothly with $d \thetav$;  by using CRNs, the smoothness can be captured by the gradient more easily (see Figure~\ref{fig:exp-crns}).
-This is analogous to using the same mini-batch for both sides of the gradient computation, if we were to apply SPSA to Bayesian learning.
-
-In addition to using CRNs in simulations for each gradient computation, we have found that using persistent or {\em sticky} random numbers helps the HABC explore the parameter landscape more easily.
-Intuitively, for a gradient-based sampling algorithm, it means a particle can slide along smooth Hamiltonian landscape because    the additive noise is suppressed.
-This is very similar to using dependent random streams to drive MCMC \cite{Murray2012,Neal2012}, the main difference we believe is that we are using the Hamiltonian dynamics to drive proposals for $\thetav$ and using $\omegav$ suppress simulation noise.
+The usefulness of applying common random numbers (CRNS) in SPSA has been previously demonstrated \cite{kleinman1999simulation}.
+In that work, the same random numbers are used to simulate both sides of the optimization function within the SPSA gradient.
+This makes sense intuitively, as we would generally assume that the expected simulation function varies smoothly in $d \thetav$;  by using CRNs, this smoothness is easily exploited (see Figure~\ref{fig:exp-crns}).
+If we were to apply SPSA to Bayesian learning, then using CRNs in the gradient step would be analogous to using the same mini-batch for both sides of the computation.
+
+In addition to using CRNs in simulations for each gradient computation, we have found that using {\em persistent} random {\em seeds} helps  HABC explore the parameter landscape more easily for some algorithms and problems.
+Intuitively, for a gradient-based sampling algorithm, it means a particle can slide along a smooth Hamiltonian landscape because    the additive noise is suppressed.
+This is very similar to using dependent random streams to drive MCMC \cite{Murray2012,Neal2012}, the main difference we believe is that we are using the Hamiltonian dynamics to drive proposals for $\thetav$ and using persistent seeds $\omegav$ to suppress simulation noise.
 
@@
 Using random seeds (versus, say, a set of random numbers) allows us to treat the simulator as a black-box, setting the random seed of its RNG without knowing the internal mechanisms it uses to generate random numbers.
-We propose including persistent random seeds $\omegav$ in the state of our Markov chain.
-As part of a Metropolis-Hastings transition operator, our idea is to randomly propose {\em flipping} each seed $\omega_s$ at time $t$ with some probability $\gamma$.
-
-Consider a Metropolis-Hastings transition where we condition on $\thetav$ and propose changing the random seeds.
+In light of our arguments above, we propose including persistent random seeds $\omegav$ in the state of our Markov chain.
+We will now describe a simple  Metropolis-Hastings transition operator that randomly proposes {\em flipping} each seed $\omega_s$ at time $t$ with some probability $\gamma$.
+
+This Metropolis-Hastings transition conditions of the current parameter location $\thetav$ and proposes changing a single random seed $\omega$ (it easily generalizes to $S$ seeds).
+The procedure is as follows: 1) propose a new seed $\omega^{'} \sim q(\omega^{'} | \omega) = \pi(\omega)$ (independent of the current seed and from its uniform prior); 2) simulate {\em deterministically} $\x^{'} = f( \thetav, \omega^{'})$; 3) compute the acceptance ratio (which reduces to the ratio of $\pi(\y | \x^{'})/\pi(\y | \x)$).
 It is straightforward to show that this leaves the target distribution invariant.
-Let $\pi(\omega_s)$ be a uniform prior over random seeds.
+The probability of the proposal is $q(x^{'}, \omega^{'} | \thetav, \omega) = \pi(\omega^{'})\delta( \x^{'} - f( \thetav, \omega^{'}))$, where $\delta(a)$ is a delta function at $a=0$.
+Because the $q$ has this form, acceptance ratio simplifies:
+\begin{equation}
+   \frac{\pi_{\epsvec}(\y | \x^{'})\pi(\omega^{'}) \pi(\x^{'} | \thetav, \omega_{'}) }{\pi_{\epsvec}(\y|\x)\pi(\omega)\pi(\x | \thetav, \omega)} \frac{\pi(\omega)\delta( \x - f( \thetav, \omega))}{\pi(\omega^{'})\delta( \x^{'} - f( \thetav, \omega^{'}))} =  \frac{\pi_{\epsvec}(\y | \x^{'})}{\pi_{\epsvec}(\y | \x)}
+% % \min\lp 1, \frac{\pi(\y,\x^{'})}{\pi(\y,\x)}  \rp
+\end{equation}
+In pseudo-marginal ABC-MCMC one could propose $q(\x^{'(s)} | \thetav)$ (fixing $\thetav$) and still sample correctly from the distribution of simulations with high likelihood at $\thetav$.
+What we propose is slightly different.
+By instead keeping the random seeds fixed, we can sample $\thetav$ using HABC and use $\omegav$ as CRNs within the gradient computation step and suppress gradient noise over time.
+In this way, random seeds carry over the same additive noise from one step to the next.
+% Let $\pi(\omega_s)$ be a uniform prior over random seeds.
 We propose $\omega_s^{'} \sim q(\omega_s^{'} | \omega_s) = \pi(\omega_s)$ (independent of the current seed).
@@
 The acceptance probability ratio reduces to a ratio of likelihoods [TODO]:
-% \begin{equation}
-%   \frac{\pi(\omega^{'})\pi(\y,\x^{'})\pi(\x^{'} | \thetav, \omega_{'})}{\pi(\omega)\pi(\y,\x)\pi(\x | \thetav, \omega)}  \rp
-% % \min\lp 1, \frac{\pi(\y,\x^{'})}{\pi(\y,\x)}  \rp
-% \end{equation}
-
-[TODO: TIE THIS IN] This implies that our simulation $\x^(s)$ changes automatically (as it is deterministic).
+% % \begin{equation}
+% %   \frac{\pi(\omega^{'})\pi(\y,\x^{'})\pi(\x^{'} | \thetav, \omega_{'})}{\pi(\omega)\pi(\y,\x)\pi(\x | \thetav, \omega)}  \rp
+% % % \min\lp 1, \frac{\pi(\y,\x^{'})}{\pi(\y,\x)}  \rp
+% % \end{equation}
+%
+% [TODO: TIE THIS IN] This implies that our simulation $\x^(s)$ changes automatically (as it is deterministic).
 We argue that this is a valid transition operator using the similar arguments that can be made for pseudo-marginal ABC-MCMC which using equation~\ref{eq:pm-proposal} proposes $\thetavp$ and $\{ \x^{'(s)}\}$ jointly.
-This is equivalent to flipping all $\omegav$ {\em and} proposing $\thetavp$.
+% This is equivalent to flipping all $\omegav$ {\em and} proposing $\thetavp$.
 In ABC-MCMC, one could instead propose $q(\x^{'(s)} | \thetav)$ (fixing $\thetav$) and still sample correctly (this is like rejection sampling $\x | \y, \thetav$ based on $\epsvec$).
@@
 The Metropolis-Hastings ratio reduces to the ratio of likelihoods for observations $\y$ given the new deterministic simulation values (just like it would for ABC-MCMC).
-By keeping the random seeds fixed, we can sample $\thetav$ using HABC and use $\omegav$ as CRNs within the gradient computation step.
 
@@
 We can analytically estimate this bias as $S \rightarrow \infty$; for this example it is $-7.8$ which is what SL estimates are centered around ($-9.3$ for $S=5$ and $7.3$ for $S=50$).
-The kernel-$\epsvec$ likelihood, on the other hand, exhibits, very low bias at $S=50$.
+The kernel-$\epsvec$ likelihood, on the other hand, exhibits low bias at $S=50$.
 However, the second important result is the variances.
-SL variances decrease quickly with $S$: $\sigma^2 = 43^2 \rightarrow 4.9^2$, whereas kernel-$\epsvec$ starts very high and remains much higher: $\sigma^2 = 147^2 \rightarrow 19^2$.
+SL variances decrease quickly with $S$: $\sigma^2 = 43^2 \rightarrow 4.9^2$, whereas kernel-$\epsvec$ starts very high and remains high: $\sigma^2 = 147^2 \rightarrow 19^2$.
 It is for this reason that we have chosen to use SL likelihoods for our gradient estimates, despite their small bias.
@@
 Although we only use $2$ classes ($0$'s versus $1$'s), the dimensionality is very high ($D=1568$).
-We show that HABC can work well despite using SPSA its gradients.
+We show that HABC can work well despite using SPSA  gradients.
  %
@@
 In Figure~\ref{fig:bf-two-d-theta} the scatter plots of trajectories are shown for two parameters.
+Though not shown due to space limitations, we have found that persistent seeds can improve convergence of the posterior predictive distribution.
+Further experiments with persistent seeds needs to be carried out to understand the extent to which the help and how to determine when they are necessary, if at all.
 
@@
 %\includegraphics[width=0.35\columnwidth]{./images/blowfly/sticky-theta-2d-SG-Thermo.pdf}
-\caption{\small{Trajectories of two parameters over the last 1000 time-steps.
+\caption{\small{Blowfly trajectories of two parameters over the last 1000 time-steps.
 {\bf Top:} SGLD and {\bf Bottom}: SG-Thermostats.
@@
 
-\subsection{Bayesian Autoencoders}\label{sec:auto}
-We perform Bayesian inference on an autoencoder neural network using MNIST images as observations.
+\subsection{Bayesian Logistic Regression}\label{sec:auto}
+We perform Bayesian inference on Bayesian inference on a logistic regression model using the digits $0$ and $1$ from MNIST.
+Despite its simplicity, the model still represents a high-dimensional problem for HABC ($D=1568$).
+We first run stochastic gradient descent to find $\thetav_{\text{MAP}}$ using the true gradient.
+We then run HABC SGLD and SG-Thermostats  starting  $\thetav_{\text{MAP}}$ to discover how well the algorithms explore the posterior.
+We compare with SGLD and SG-Thermostats using the {\em true} gradients.
+
+We use $n=100$ mini-batches and $R=10$ for SPSA.  Figure~\ref{fig:blr}
+\begin{figure}[t]
+%\vskip 0.2in
+\begin{center}
+\includegraphics[width=0.75\columnwidth]{./images/blowfly/non-sticky-theta-2d-SG-Thermo.pdf}
+\caption{\small{Bayesian logistic regression.}}
+\label{fig:blr}
+\end{center}
+\vskip -0.2in
+\end{figure}
+
+%  neural network using MNIST images as observations.
 We apply two inference algorithms: SGNHT and HABC.
@@
 Use same batches at each iteration for SGNHT and HABC?
-
-results: convergence of training error (batches), convergence in test error using parameters collected (record nbr of forward or backward passes required), show mean and variance of filters (how?), compare with using the last parameter sample only (stochastic optimization).
+%
+% results: convergence of training error (batches), convergence in test error using parameters collected (record nbr of forward or backward passes required), show mean and variance of filters (how?), compare with using the last parameter sample only (stochastic optimization).
 
@@
 Hamiltonian ABC proposes a new set of algorithms for Bayesian inference of likelihood-free models.
-HABC builds  upon the connections between Hamilton Monte Carlo with stochastic gradients and well-established gradient approximations based on a minimal number of forward simulations, even for high-dimensional parameter spaces.
-By showing the feasibility of running ABC in high-dimensions, we hope that the door has been opened for exploration of larger simulation-based models.
-
-Another innovation we introduce is the use of sticky or persistent random seeds to suppress the simulator noise and therefore smooth the simulation landscape over a local region of parameter space.
-By doing this, all our methods, but especially SGHMC and SGNHT, with their longer Hamiltonian trajectories, benefit from this smoothness that changes slowly over time.
-We feel that new classes of ABC algorithms could develop from using persistent random seeds, not solely gradient-based samplers.
+HABC builds  upon the connections between Hamilton Monte Carlo with stochastic gradients and well-established gradient approximations based on a minimal number of forward simulations, even in high-dimensions.
+By showing the feasibility of running ABC on both small and large problems, we hope that the door has been opened for exploration of larger simulation-based models.
+
+Another innovation we introduce is the use of persistent random seeds to suppress the simulator noise and therefore smooth the simulation landscape over a local region of parameter space.
+For some algorithms run on certain models, improved performance has been observed.
+This is most likely to be the case for simulators with large additive noise and algorithms that benefit from long Hamiltonian trajectories (i.e.
+SGHMC and SG-Thermostats).
+We feel that new classes of ABC algorithms could develop from using persistent random seeds, not just gradient-based samplers but traditional ABC-MCMC.
 
@@
 One final issue is monitoring or determining whether the correct amount of noise is being injected to ensure proper sampling.
-In SGLD, for example, we can always turn down $\eta$ so that the injected noise term dominates, but when our goal is efficient exploration of the posterior, this is a not very satisfying solution.
+In SGLD \cite{welling2011bayesian}, for example, we can always turn down $\eta$ so that the injected noise term dominates, but when our goal is efficient exploration of the posterior, this is  not a very satisfying solution.
 
@@
 However, previous work in Bayesian inference has shown the usefulness of HMC-based proposals based on Gaussian process of log-likelihood surfaces \cite{rasmussen:2003}.
-We could similarly use HABC with recent ABC surrogate models \cite{Meeds2014GpsUai,wilkinson:2014} to avoid simulations whenever possible, yet benefit from Hamiltonian dynamics.
+We could similarly use HABC with recent ABC surrogate models \cite{Meeds2014GpsUai,wilkinson:2014} to avoid simulations whenever possible, yet still  benefit from Hamiltonian dynamics.
 %We address this scenario in Section~\ref{future}.  

### 2015-03-03 draft.tex
penultimate

@@
 \begin{center}
-\includegraphics[width=0.95\columnwidth]{./images/exp_crn_figure.pdf}
+\includegraphics[width=0.75\columnwidth]{./images/exp_crn_figure.pdf}
+\vspace{-0.1in}
 \caption{\small{A view of a simulator in terms of common random numbers.
@@
 \end{center}
-\vskip -0.2in
+\vspace{-0.2in}
 \end{figure}
@@
 \begin{center}
-\includegraphics[width=0.95\columnwidth]{./images/exp_varg_figure.pdf}
+\includegraphics[width=0.75\columnwidth]{./images/exp_varg_figure.pdf}
+\vspace{-0.1in}
 \caption{\small{Variance of gradient estimation using kernel-$\eps$ and SL for different values of $S\in\{5,50\}$ and fixed $\eps = 0.37$ (the same used in the other results).  
@@
 \begin{center}
-\includegraphics[width=0.32\columnwidth]{./images/exp-SL-MCMC-posterior-hist-omega-rate-100p0-chain0.pdf}
-\includegraphics[width=0.32\columnwidth]{./images/exp-SG-Langevin-posterior-hist-omega-rate-100p0-chain0.pdf}
-\includegraphics[width=0.32\columnwidth]{./images/exponential/exp3-SG-Thermostats-posterior-hist-omega-rate-100p0-chain0.pdf}
-
-\includegraphics[width=0.32\columnwidth]{./images/exponential/exp2-SL-MCMC-posterior-hist-omega-rate-0p1-chain1.pdf}
-\includegraphics[width=0.32\columnwidth]{./images/exponential/exp2-SG-Langevin-posterior-hist-omega-rate-0p1-chain0.pdf}
-\includegraphics[width=0.32\columnwidth]{./images/exponential/exp2-SG-Thermostats-posterior-hist-omega-rate-0p1-chain1.pdf}
-
+\includegraphics[width=0.27\columnwidth]{./images/exp-SL-MCMC-posterior-hist-omega-rate-100p0-chain0.pdf}
+\includegraphics[width=0.27\columnwidth]{./images/exp-SG-Langevin-posterior-hist-omega-rate-100p0-chain0.pdf}
+\includegraphics[width=0.27\columnwidth]{./images/exponential/exp3-SG-Thermostats-posterior-hist-omega-rate-100p0-chain0.pdf}
+
+\includegraphics[width=0.27\columnwidth]{./images/exponential/exp2-SL-MCMC-posterior-hist-omega-rate-0p1-chain1.pdf}
+\includegraphics[width=0.27\columnwidth]{./images/exponential/exp2-SG-Langevin-posterior-hist-omega-rate-0p1-chain0.pdf}
+\includegraphics[width=0.27\columnwidth]{./images/exponential/exp2-SG-Thermostats-posterior-hist-omega-rate-0p1-chain1.pdf}
+\vspace{-0.1in}
 %\includegraphics[width=0.45\columnwidth]{./images/exp-ABC-MCMC-posterior_hist.pdf}
@@
 \end{center}
-\vskip -0.2in
+\vspace{-0.1in}
 \end{figure} 
@@
 The horizontal line with shading indicates $y \pm 2 \eps$, where $\eps = 0.37$ is used throughout the demonstration.
-
 % \begin{figure*}[t]
@@
 % \end{figure*}
+\vspace{-0.2in}
 
 \subsection{Bias and Variance of $\nabla \hat{U}(\thetav)$}
+\vspace{-0.1in}
 To test our assumption that the synthetic-likelihood model is better suited for HABC, we ran FDSA at the true $\theta_{\text{MAP}}$.
@@
 \includegraphics[width=0.65\columnwidth]{./images/exponential/exp2-SG-Thermostats-theta-timeseries-omega-rate-0p1-chain3.pdf}
+\vspace{-0.15in}
 \caption{\small{Trajectories of the last $1000$ $\thetav$ samples for the demonstration problem.
@@
 \end{center}
-\vskip -0.2in
+\vspace{-0.1in}
 \end{figure*}
@@
 \begin{table}[h]
+  \vspace{-0.1in}
 \caption{Average total variational distance (tvd) for the demonstration problem.
@@
 The results for SGHMC (not shown) and SGNHT are nearly identical.
-All algorithms tend to benefit from persistent seeds, in particular SGHMC and SGNHT, the methods with long Hamiltonian trajectories.}
+%All algorithms tend to benefit from persistent seeds, in particular SGHMC and SGNHT, the methods with long Hamiltonian trajectories.
+}
 \label{tab:exp-posterior}
@@
 \end{table}
+\vspace{-0.2in}
 % \begin{table}[h]
@@
     \includegraphics[width=0.65\columnwidth]{./images/blowfly/non-sticky-theta-1d-SG-Thermo.pdf}
+   \vspace{-0.1in}
    \caption{\small{Blowfly posterior distributions (non-persistent seeds).
@@
 %\includegraphics[width=0.35\columnwidth]{./images/blowfly/sticky-theta-2d-SG-Thermo.pdf}
+\vspace{-0.15in}
 \caption{\small{Blowfly trajectories of two parameters over the last 1000 time-steps.
@@
 \end{center}
-\vskip -0.2in
+\vspace{-0.1in}
 \end{figure}
 
+\vspace{-0.1in}
 \subsection{Bayesian Logistic Regression}\label{sec:auto}
@@
 
-We use $n=100$ mini-batches and $R=10$ for SPSA.  Figure~\ref{fig:blr}
+We use $n=100$ mini-batches and $R=10$ for SPSA.
+Figure~\ref{fig:blr} shows samples projected onto 2 dimensions using random projections.
+We can clearly see that the trajectories using SPSA exhibit very similar behavior to Bayesian learning with the true gradients.
 \begin{figure}[t]
@@
 \begin{center}
-\includegraphics[width=0.75\columnwidth]{./images/blowfly/non-sticky-theta-2d-SG-Thermo.pdf}
-\caption{\small{Bayesian logistic regression.}}
+\includegraphics[width=0.95\columnwidth]{./images/lr/random_proj.png}
+\vspace{-0.15in}
+\caption{\small{Bayesian logistic regression sampling trajectories randomly projected.
+{\bf White circle:} $\thetav_{\text{MAP}}$}; {\bf Dark blue:} true SGLD; {\bf Red:} SPSA SGLD; {\bf Green:} true SG-Thermostats; {\bf Cyan:} SPSA SG-Thermostats.}
 \label{fig:blr}
@@
 HABC builds  upon the connections between Hamilton Monte Carlo with stochastic gradients and well-established gradient approximations based on a minimal number of forward simulations, even in high-dimensions.
-By showing the feasibility of running ABC on both small and large problems, we hope that the door has been opened for exploration of larger simulation-based models.
+We have performed some preliminary experiments showing the feasibility of running ABC on both small and large problems, and we hope that the door has been opened for exploration of larger simulation-based models using HABC.
 

### 2015-03-03 draft.tex
penpenultimate

@@
 We compare with SGLD and SG-Thermostats using the {\em true} gradients.
-
+%
 We use $n=100$ mini-batches and $R=10$ for SPSA.
@@
 We can clearly see that the trajectories using SPSA exhibit very similar behavior to Bayesian learning with the true gradients.
+This is very positive result that indicates HABC can successfully exploit the noisy and less informative gradients of SPSA.
 \begin{figure}[t]
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\vspace{-0.1in}
 Hamiltonian ABC proposes a new set of algorithms for Bayesian inference of likelihood-free models.
@@
 However, previous work in Bayesian inference has shown the usefulness of HMC-based proposals based on Gaussian process of log-likelihood surfaces \cite{rasmussen:2003}.
-We could similarly use HABC with recent ABC surrogate models \cite{Meeds2014GpsUai,wilkinson:2014} to avoid simulations whenever possible, yet still  benefit from Hamiltonian dynamics.
+We could similarly use HABC with  ABC surrogate models \cite{Meeds2014GpsUai,wilkinson:2014} to minimize simulation calls, yet still  benefit from Hamiltonian dynamics.
 %We address this scenario in Section~\ref{future}.  

### 2015-03-03 draft.tex
penpenpenultimate

@@
 We also describe a new simple yet general approach of incorporating random seeds into the state of the Markov chain, further reducing the random walk behavior of HABC.
-We demonstrate HABC on several typical ABC problems, and show that HABC performs comparably to regular Bayesian inference on a high-dimensional problem from machine learning.
+We demonstrate HABC on several typical ABC problems, and show that HABC samples comparably to regular Bayesian inference using true gradients on a high-dimensional problem from machine learning.
   %
@@
 We use $n=100$ mini-batches and $R=10$ for SPSA.
-Figure~\ref{fig:blr} shows samples projected onto 2 dimensions using random projections.
+Figure~\ref{fig:blr} shows samples randomly projected onto 2 dimensions ($1000$ evenly sub-sampled from $10K$).
 We can clearly see that the trajectories using SPSA exhibit very similar behavior to Bayesian learning with the true gradients.
@@
 \begin{center}
-\includegraphics[width=0.95\columnwidth]{./images/lr/random_proj.png}
+%\includegraphics[width=0.95\columnwidth]{./images/lr/random_proj.png}
+\includegraphics[width=0.9\columnwidth]{./images/lr/lr_trajectories.pdf}
 \vspace{-0.15in}
-\caption{\small{Bayesian logistic regression sampling trajectories randomly projected.
-{\bf White circle:} $\thetav_{\text{MAP}}$}; {\bf Dark blue:} true SGLD; {\bf Red:} SPSA SGLD; {\bf Green:} true SG-Thermostats; {\bf Cyan:} SPSA SG-Thermostats.}
+\caption{\small{Bayesian logistic regression sampling trajectories randomly projected.  The yellow circle is the projected $\thetav_{\text{MAP}}$.}  
+%{\bf White circle:} $\thetav_{\text{MAP}}$}; {\bf Dark blue:} true SGLD; {\bf Red:} SPSA SGLD; {\bf %Green:} true SG-Thermostats; {\bf Cyan:} SPSA SG-Thermostats.
+}
 \label{fig:blr}

### 2015-06-08 supp.tex

\documentclass[]{article}
\usepackage{proceed2e}

% Set the typeface to Times Roman
\usepackage{times}
\usepackage{bm}
\usepackage{amsmath}
\usepackage[numbers]{natbib}
%\usepackage{array}
%\usepackage{nips_style} 
%\usepackage{sidecap}
%\usepackage[export]{adjustbox}
%\include{cdefs}

% For algorithms
\usepackage{algorithm}
\usepackage{algorithmic}
\include{cdefs}
\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{STOCHASTIC-GRADIENT LD} \label{sgld}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{eqnarray}
  h & = & \frac{N}{n} \sum_{i=1}^n \nabla \log p(\bx_i | \thetav)- \sum_{i=1}^N \nabla \log p(\bx_i | \thetav ) \\
    & = & \frac{N}{n} \sum_{i=1}^n s_i \\
  s_i & = & \nabla \log p(\bx_i | \thetav) + \frac{1}{N} \nabla p(\thetav) \\
  \text{V}(h) & = & \frac{N^2}{n^2} \sum_{i=1}^n \text{Var}(s_i) \\
                & = & \frac{N^2}{n^2} \sum_{i=1}^n \frac{1}{n} \lp \sum_{j=1} (s_i - \bar{s})^2 \rp \\
                & = & \frac{N^2}{n^2} \sum_{i=1}^n V_s \\
                & = & \frac{N^2}{n^2} n V_s \\
                & = & \frac{N^2}{n} V_s 
  %\text{Var}(\thetav) & = & \frac{\epsilon^2}{4}\text{Var}(h) + \epsilon \\
  %       \frac{\epsilon^2}{4}\text{Var}(h) & << & \epsilon \\
  %       \frac{\epsilon N^2}{4 n} V_s & << & 1       
\end{eqnarray}
\begin{eqnarray}
  \text{Var}(\thetav) & = & \frac{\epsilon^2}{4}\text{Var}(h) + \epsilon \\
         \frac{\epsilon^2}{4}\text{Var}(h) & << & \epsilon \\
         \frac{\epsilon N^2}{4 n} V_s & << & 1       
\end{eqnarray}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{SGLD-ABC} \label{sgld}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{ABC Cost Function}

For optimization purposes we can search for the maximum a posteriori (MAP) estimator $\hat{\thetav}$ using the log-joint distribution $\hat{\thetav} = \argmax_{\thetav} \mathcal{MAP}$, where $\mathcal{MAP}$ is
\begin{eqnarray}
  \mathcal{LJ} &    =    & \log \pi(\thetav) + \log \pi\lp \y | \thetav \rp \\
                & \approx &\log \pi(\thetav) + \log \sum_{s=1}^S \pi_{\epsvec}\lp \y | f\lp \thetav, \omega_s \rp \rp - \log S
\end{eqnarray} 
we can also maximize the lower-bound of $\mathcal{LJ} \geq \mathcal{LB}$, where 
\begin{eqnarray}
  \mathcal{LB} & \approx &\log \pi(\thetav) + \sum_{s=1}^S \log  \pi_{\epsvec}\lp \y | f\lp \thetav, \omega_s \rp \rp
\end{eqnarray} 
which will be more convenient for optimization for typical kernels (i.e. $\pi_{\epsilon}$ is Gaussian).

The simplest optimization is to set the objective function  $C$ to be $-\mathcal{LJ}$ or $-\mathcal{LB}$ and compute stochastic gradients using 2SPSA with $q$ repeats (negative log-likelihood because we will assume minimization):
\begin{eqnarray}
  \frac{\partial C}{\partial \thetav} & = & \frac{1}{q} \sum_{j=1}^{q} \frac{C\lp \thetav + c \Delta_j\rp - C{\lp\thetav - c \Delta_j\rp}}{2 c \Delta_j}
\end{eqnarray}
or 1SPSA
\begin{eqnarray}
  \frac{\partial C}{\partial \thetav} & = & \frac{1}{q} \sum_{j=1}^{q}  \frac{C\lp \thetav + c \Delta_j\rp - C(\thetav)}{c \Delta_j}
\end{eqnarray}
which is less accurate but is less computation if $C(\thetav)$ is computed at each step (TODO: possible algorithm).  TODO: rewrite $\frac{\partial C}{\partial \thetav}$ as $\hat{g}$ or similar.

\subsection{Special Case: Gaussian Kernels}
When the kernel is known we can use the chain rule to compute the gradients with the first part from the kernel and the second from the variation of the simulator pseudo-statistics, whose gradient can be estimated using SPSA.  Expanding the likelihood over $J$ statistics, the lower bound on the loglikelihood becomes:
\begin{eqnarray}
  \hat{L} & = & \sum_s \log \pi_{\epsilon}\lp \y | f(\thetav, \omega_s)\rp \\
          & = & \sum_s \log \prod_{j=1}^J \pi_{\epsilon_j}\lp y_j | f_j(\thetav, \omega_s)\rp \\
          & = & \sum_s \sum_j \log  \pi_{\epsilon_j}\lp y_j | f_j(\thetav, \omega_s)\rp \\
          & = & \sum_s \sum_j -\frac{1}{2} \frac{\lp y_j - f_j(\thetav, \omega_s)\rp^2}{\epsilon_j^2} - Z_j \\
\end{eqnarray}

The gradient:
\begin{eqnarray} 
  \nabla \hat{L} & = & \sum_s \sum_j \frac{\lp y_j - f_j(\thetav, \omega_s)\rp}{\epsilon_j^2} \cdot \frac{\partial f_j(\thetav, \omega_s)}{\partial \thetav} \\
                 & = & \sum_s \sum_j e_{js} \nabla_{js}
\end{eqnarray}
The first part $e_{js}$ is analytic, the second part $\nabla_{js}$ requires estimation by 2SPSA:
\begin{eqnarray} 
  \hat{ \nabla }_{js}  & = & \frac{1}{q} \sum_{r=1}^q \frac{f_j(\thetav + c \Delta_r , \omega_s) - f_j(\thetav - c \Delta_r , \omega_s)}{2 c \Delta_r}
\end{eqnarray}
or 1SPSA
\begin{eqnarray} 
  \hat{ \nabla }_{js} & = & \frac{1}{q} \sum_{r=1}^q \frac{f_j(\thetav + c \Delta_r , \omega_s) - f_j(\thetav, \omega_s)}{c \Delta_r}
\end{eqnarray}
the benefit of 1SPSA in this case is that the first gradients  $e_{js}$ can be computed from $f_j(\thetav, \omega_s)$.

\section{Applying SGLD to ABC}

There are two main sources of randomness in the ABC gradient computation: the random perturbation mask and the random seed of the simulator model, which controls the simulator noise.  This noise can be eliminated by using common random numbers \cite{spall-crn}.  On the other hand, it is not clear that this is the best strategy, as the noise from the masks at a fixed seed may be higher than

\begin{eqnarray}
  V_{\gradestimaterepeat{t}{\thetav_t}{j}} & = & \frac{1}{q-1} \sum_{j=1}^q (\gradestimaterepeat{t}{\thetav_t}{j} - \gradestimate{t}{\thetav_t})^2 \\
  V_{\gradestimate{t}{\thetav_t}} & = & \frac{1}{q-1} V_{\gradestimaterepeat{t}{\thetav_t}{j}} \\
  \frac{\epsilon^2}{4 }V_{\gradestimate{t}{\thetav_t}} & = & \frac{\epsilon^2}{4 q }V_{\gradestimaterepeat{t}{\thetav_t}{j}} \\
                                                       & << & \epsilon \\
                    \frac{\epsilon}{4 (q-1) }V_{\gradestimaterepeat{t}{\thetav_t}{j}}  & << & 1                                  
\end{eqnarray}

Check for $g(\thetav) + h(\thetav) + \eta$ that $h(\thetav) + \eta \approx \eta$

Also $\theta_{t+1} = \theta_t + \frac{\epsilon}{2}\gradestimate{t}{\thetav_t} + \sqrt{\epsilon}\mathcal{N}(0,1)$

\subsection{Statistical Tests for Gradients}

From Byrd:
\begin{eqnarray}
\frac{||V_{\gradestimaterepeat{t}{\thetav_t}{j}} ||_1}{q} & \leq r^2 || \hat{g}_t ||_2^2 \\
\hat{q} & = & \frac{||V_{\gradestimaterepeat{t}{\thetav_t}{j}} ||_1}{r^2 || \hat{g}_t ||_2^2}
\end{eqnarray}

TODO:
\begin{itemize}
  \item Check that $q$ gradients are normal.  Especially as $q$ increases.
  \item Merge Byrd's test for $\hat{q}$ and condition on SGLD-ABC.
  \item Variance check for increasing $q$ can also be used to stop optimization since the variance is less than the noise in the gradients.
\end{itemize}

\subsection{SGD, AdaGrad, RMSProp, AdaM}

\begin{itemize}
  \item  Arguably the two most useful procedures in simulation-based science are optimization and Bayesian inference.
  \item  Optimization can take the form of simple grid-search to sophisticated techniques like factorial design \cite{factorialdesign}, Bayesian experiment design \cite{bayesexpdesign}, (mention others).
  \item Recently, Bayesian optimization techniques have shown success in optimizing very expensive black-box simulators \cite{bosuccess}.
  \item Though gradients are not directly computable for simulators, they can be computed analytically by using finite differences (see \cite{kiefer1952}).  This quickly becomes infeasible for large $p$ problems.  There is however an alternative stochastic approximation algorithm by Spall \cite{spall1999} that requires only 2 simulation calls independent of $p$.
  \item By using this approximation, gradient-based algorithms can be adopted for simulators: for optimization, using analogous stochastic gradient descent algorithms and for Bayesian inference using Langevin dynamics \cite{langevin}.
\end{itemize}

With our stochastic gradient in hand, we can make use of robust stochastic gradient rules, including recent advances from machine learning.

\subsubsection{SGD}
The basic update rule is at time $t$ for minimization is
\begin{eqnarray}
  \Delta\thetav & = & - \alpha_t \hat{g_t}
\end{eqnarray}
with conditions $\sum_{t=1}^{\infty} \alpha_t = \infty$ and $\sum_{t=1}^{\infty} \alpha_t^2 < \infty$ required for convergence.
An additional momentum parameter $\beta_1$ can be added to improve convergence
\begin{eqnarray}
  \gmom & = & \beta_1 \gmom + (1-\beta_1)*\hat{g_t} \\
  \Delta\thetav & = & - \alpha_t \gmom
\end{eqnarray}

\subsubsection{AdaGrad}
AdaGrad uses the total cumulative element-wise square of the gradients to automatically both decrease $\alpha_t$ and scale the learning rates for each dimension:
\begin{eqnarray}
  \gmom    & = & \beta_1 \gmom + (1-\beta_1)*\hat{g_t} \\
  \gsquare & = & \sum_{t'=1}^t \hat{g_t}^2 \\
  \Delta\thetav & = & - \alpha_t \gmom / \sqrt{\gsquare}
\end{eqnarray}

\subsubsection{RMSProp}
\begin{eqnarray}
  \gmom    & = & \beta_1 \gmom + (1-\beta_1)*\hat{g_t} \\
  \gsquare & = & \beta_2 \gsquare + (1-\beta_2)*\hat{g_t}^2 \\
  \Delta\thetav & = & - \alpha_t \gmom / \sqrt{\gsquare}
\end{eqnarray}

\subsubsection{AdaM}
\begin{eqnarray}
  \gmom    & = & \beta_1 \gmom + (1-\beta_1)*\hat{g_t} \\
  \gsquare & = & \beta_2 \gsquare + (1-\beta_2)*\hat{g_t}^2 \\
  \gamma & = & \frac{ \sqrt{1-(1-\beta_2)^t}}{(1-(1-\beta_1)^t)} \\
  \Delta\thetav & = & - \alpha_t \gamma \gmom / \sqrt{\gsquare}
\end{eqnarray}

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \section{EXPERIMENTS} \label{experiments}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \subsection{Logistic Regression with Stochastic Gradients}
%
% \subsection{Optimization of Blowfly Dynamics}
%
% \subsubsection{Using AdaM}
%
% \begin{figure}[h]
% \vspace{1in}
% \includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_post_series.png}
% \includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_post_series_sgld5k.png}
% \includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_post.png}
% \includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_post_sgld5k.png}
% \includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_postpred_series.png}
% \includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_postpred_series_sgld5k.png}
% \includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_postpred.png}
% \includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_postpred_sgld5k.png}
% \includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_noise.png}
% \includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_noise_sgld5k.png}
% \caption{Optimizing Blowfly using $q=5$, $\epsilon = 0.01$, $\beta_1=0.75$,$\beta_2=0.9$,$\alpha=0.1$ and $c=1$.  Left: Optimization using AdaM, Right: ``SGLD'' with estimated noise. }
% \end{figure}

% \subsubsection{Using AdaM with ``SGLD" injected noise}

% \subsection{Likelihood-free Inference of Blowfly Dynamics}

% \subsection{Other Useful ABC problem}

\end{document}

### 2015-06-08 uai-cr.tex
added refs

--- 2015-03-03 draft.tex
@@
 In ABC, the vector of $J$ observations are typically informative statistics of the raw observations.
-It can be shown that if the statistics used in the likelihood function are sufficient, then these algorithms sample correctly from an approximation to the true posterior.
+It can be shown that if the statistics used in the likelihood function are sufficient, then these algorithms sample correctly from an approximation to the true posterior \cite{marin:2012}.
   The simulator is treated as generator of random pseudo-observations, i.e.
@@
 
-Of particular concern to this paper is the behavior of the log-likelihoods for different values of $\epsilon$.
-In the $\eps$-kernel case, the log-likelihood is very sensitive to small values of $\eps$:
+Of particular concern to this paper is the behavior of the log-likelihoods for different values of $\epsvec$.
+In the $\epsvec$-kernel case, the log-likelihood is very sensitive to small values of $\epsvec$:
 \begin{eqnarray}
@@
 where $m$ is the simulation that is closest to $\y$, $H$ is a sum over terms close to $0$.
-We can see that the log-likelihood can be set arbitrarily small by decreasing $\eps$.
+We can see that the log-likelihood can be set arbitrarily small by decreasing $\epsvec$.
 On the other hand, by using a model of the simulation at $\thetav$
@@
  For the SL model, $\epsvec$ acts as a smoothing term and can be set to small values with little change to the log-likelihood, as long as the SL estimators are fit appropriately.
-This insensitivity to $\eps$ will be used in Section~\ref{sec:habc} for estimating gradients of the ABC likelihood.
+This insensitivity to $\epsvec$ will be used in Section~\ref{sec:habc} for estimating gradients of the ABC likelihood.
 Before describing HABC in full detail however, we now explain how scaling Hamiltonian dynamics in Bayesian learning can be accomplished using stochastic gradients from batched data.
@@
 
-On the other hand, the synthetic-likelihood is stable; using a deterministic representation, 
+On the other hand, the synthetic-likelihood is stable; using a deterministic representation, we have 
 \begin{equation}
@@
 Hamiltonian ABC proposes a new set of algorithms for Bayesian inference of likelihood-free models.
-HABC builds  upon the connections between Hamilton Monte Carlo with stochastic gradients and well-established gradient approximations based on a minimal number of forward simulations, even in high-dimensions.
+HABC builds  upon the connections between Hamiltonian Monte Carlo with stochastic gradients and well-established gradient approximations based on a minimal number of forward simulations, even in high-dimensions.
 We have performed some preliminary experiments showing the feasibility of running ABC on both small and large problems, and we hope that the door has been opened for exploration of larger simulation-based models using HABC.

### 2015-06-08 draft.tex
Added larger legend overlay

@@
 \usepackage{graphicx} % more modern
+\usepackage{tikz}
 %\usepackage{epsfig} % less modern
@@
 \date{}
-
+\pagenumbering{gobble}
 \pagestyle{myheadings}
@@
 \begin{center}
-\includegraphics[width=0.65\columnwidth]{./images/exp-SL-MCMC-theta-timeseries-omega-rate-100p0-chain0.pdf}
-\includegraphics[width=0.65\columnwidth]{./images/exp-SG-Langevin-theta-timeseries-omega-rate-100p0-chain0.pdf}
-\includegraphics[width=0.65\columnwidth]{./images/exponential/exp3-SG-Thermostats-theta-timeseries-omega-rate-100p0-chain0.pdf}
-\includegraphics[width=0.65\columnwidth]{./images/exponential/exp2-SL-MCMC-theta-timeseries-omega-rate-0p1-chain3.pdf}
-\includegraphics[width=0.65\columnwidth]{./images/exponential/exp2-SG-Langevin-theta-timeseries-omega-rate-0p1-chain3.pdf}
-\includegraphics[width=0.65\columnwidth]{./images/exponential/exp2-SG-Thermostats-theta-timeseries-omega-rate-0p1-chain3.pdf}
+\begin{tikzpicture}
+    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=0.65\columnwidth]{./images/exp-SL-MCMC-theta-timeseries-omega-rate-100p0-chain0.pdf}};
+    \node[draw, fill=white] at (4.37, 1.95) {\small SL-MCMC};
+\end{tikzpicture}
+\begin{tikzpicture}
+    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=0.65\columnwidth]{./images/exp-SG-Langevin-theta-timeseries-omega-rate-100p0-chain0.pdf}};
+    \node[draw, fill=white] at (4.25, 1.93) {\small SG-Langevin};
+\end{tikzpicture}
+\begin{tikzpicture}
+    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=0.65\columnwidth]{./images/exponential/exp3-SG-Thermostats-theta-timeseries-omega-rate-100p0-chain0.pdf}};
+    \node[draw, fill=white] at (4.06, 1.95) {\small SG-Thermostats};
+\end{tikzpicture}
+\begin{tikzpicture}
+    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=0.65\columnwidth]{./images/exponential/exp2-SL-MCMC-theta-timeseries-omega-rate-0p1-chain3.pdf}};
+    \node[draw, fill=white] at (4.37, 1.95) {\small SL-MCMC};
+\end{tikzpicture}
+\begin{tikzpicture}
+    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=0.65\columnwidth]{./images/exponential/exp2-SG-Langevin-theta-timeseries-omega-rate-0p1-chain3.pdf}};
+    \node[draw, fill=white] at (4.25, 1.93) {\small SG-Langevin};
+\end{tikzpicture}
+\begin{tikzpicture}
+    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=0.65\columnwidth]{./images/exponential/exp2-SG-Thermostats-theta-timeseries-omega-rate-0p1-chain3.pdf}};
+    \node[draw, fill=white] at (4.06, 1.95) {\small SG-Thermostats};
+\end{tikzpicture}
+%\includegraphics[width=0.65\columnwidth]{./images/exp-SG-Langevin-theta-timeseries-omega-rate-100p0-chain0.pdf}
+%\includegraphics[width=0.65\columnwidth]{./images/exponential/exp3-SG-Thermostats-theta-timeseries-omega-rate-100p0-chain0.pdf}
+%\includegraphics[width=0.65\columnwidth]{./images/exponential/exp2-SL-MCMC-theta-timeseries-omega-rate-0p1-chain3.pdf}
+%\includegraphics[width=0.65\columnwidth]{./images/exponential/exp2-SG-Langevin-theta-timeseries-omega-rate-0p1-chain3.pdf}
+%\includegraphics[width=0.65\columnwidth]{./images/exponential/exp2-SG-Thermostats-theta-timeseries-omega-rate-0p1-chain3.pdf}
 \vspace{-0.15in}

### 2015-06-08 uai-cr.tex
typos persistent

@@
 
-\author{} % LEAVE BLANK FOR ORIGINAL SUBMISSION.
+%\author{} % LEAVE BLANK FOR ORIGINAL SUBMISSION.
           % UAI  reviewing is double-blind.
+% \author{ {\bf Edward Meeds } \\
+% Computer Science Dept. \\
+% Cranberry University\\
+% Pittsburgh, PA 15213 \\
+% \And
+% {\bf Coauthor}  \\
+% Affiliation          \\
+% Address \\
+% \And
+% {\bf Coauthor}   \\
+% Affiliation \\
+% Address    \\
+% (if needed)\\
+% }
 
+\author{ {\bf Edward Meeds}  \\
+Informatics Institute\\
+University of Amsterdam\\
+\texttt{tmeeds@gmail.com} \\
+\And
+{\bf Robert Leenders } \\
+Informatics Institute \\
+University of Amsterdam \\
+\texttt{leenders.robert@gmail.com} \\
+\And
+{\bf Max Welling }\\
+Informatics Institute\\
+University of Amsterdam\\
+\texttt{welling.max@gmail.com}
+}
+%
 % \author{
@@
 Second, they can maintain reasonable approximations to the Hamiltonian dynamics and therefore avoid a Metropolis-Hastings correction step involving the full data set.
-Different strategies are employed to do this: small step-sizes combined with  Langevin dynamics \cite{welling2011bayesian}, using friction to prevent accumulation of errors in the Hamiltonian \cite{chen2014stochastic}, and using a thermostat to control the temperature of the Hamiltonian \cite{ding2014bayesian}.
+Different strategies are employed to do this: small step-sizes combined with  Langevin dynamics \cite{welling2011bayesian} (stochastic gradient Langevin dynamics---SGLD), using friction to prevent accumulation of errors in the Hamiltonian \cite{chen2014stochastic} (stochastic gradient HMC---SGHMC), and using a thermostat to control the temperature of the Hamiltonian \cite{ding2014bayesian} (stochastic gradient Nose-Hoover thermostats---SGNHT).
 Each of these strategies can be used by HABC.
@@
 The second difficulty is that for high-dimensional parameter spaces, computing the gradients naively (i.e.
-by finite differences \cite{kiefer1952stochastic}) can squash the gains brought by the Hamiltonian dynamics.
+by finite differences (FD) \cite{kiefer1952stochastic}) can squash the gains brought by the Hamiltonian dynamics.
 Fortunately, we can use existing stochastic approximation algorithms \cite{spall1992multivariate,spall2000adaptive} that can be used to compute unbiased estimators of the gradient with a small number of forward simulations that is {\em independent} of the parameter dimension.
@@
 
-%Approximate Bayesian Computation (ABC) is a set of inference procedures for so-called {\em likelihood-free} models \cite{abccitations}.
-In contrast to typical Bayesian modeling, where the parameters of a probabilistic model of data are the focus, the primary interest in ABC are the parameters of a simulation model.
-The simulator can generate pseudo or auxiliary data by running the simulator with a set of parameter values, but it cannot compute a likelihood directly.
-Inference involves running simulations, comparing pseudo data with observations, and using a distance between the two as an approximate likelihood.
-Markov chain Monte Carlo (MCMC) or Sequential Monte Carlo (SMC) are the most common inference algorithms.
-It can be shown that if the statistics used in the likelihood function are sufficient, then these algorithms sample correctly from an approximation to the true posterior.
-
-%The approximation to the true posterior can be made arbitrarily small, in principle, by controlling the kernel width parameter $\epsilon$.
-As $\epsilon \rightarrow 0$, however, simulated data must be closer and closer to the observations (or receive zero likelihood).
-The acceptance rate of a Markov chain for small $\epsilon$ can therefore be low.
-This inefficiency of ABC sampling is exacerbated by the random walk proposals.
-As the dimensionality of the parameter space grows,  $\epsilon$ must grow to prevent low acceptance rates.
-This has perhaps prevented application of ABC to high-dimensional simulators.
-Our paper addresses this directly: can we use the sampling efficiency of Hamiltonian dynamics (i.e.
-high acceptance rates) in high-dimensional simulators?
-
-%In this paper we adopt the developments in scaling Bayesian inference to ABC.
-To do this we use forward simulations to approximate the likelihood-free gradient.
-The key difference with SG methods mentioned above is that the stochasticity of the gradient does not come from approximating the full data gradient with a batch gradient, but by the stochasticity of the simulator.
-It is therefore not the expense of the simulator (though this could very well be the case for many interesting simulation-based models, and we address this in Future Work) that requires an approximation to the gradient, but the likelihood-free nature of the problem.
-For high-dimensional parameter spaces, computing the gradients naively (i.e.
-by finite differences) can squash the gains brought by the Hamiltonian dynamics; to address this further approximations to the gradient can be used \cite{spall}.
-
-%The approximate gradients used by HABC can be noisy and difficult to manage is care is not taken.
-To address this we show how a local Gaussian approximation of the simulator output can improve the stability of the gradient estimates over a direct estimate using the ABC kernel function.
-
-A further innovation of this paper is the use of common random numbers (CRN) to improve the efficiency of the Hamiltonian dynamics.
-The idea behind CRNs is to use the same set of random seeds for estimating a gradient by FD or SPSA, i.e.
+A further innovation of this paper is the use of persistent random numbers (PRNs) to improve the efficiency of the Hamiltonian dynamics.
+The idea behind PRNs is to use the same set of random seeds for estimating a gradient by FD or SPSA, i.e.
 when simulating $\pi(\x| \theta + d\theta)$ and $\pi(\x| \theta - d\theta)$ use the same random seeds.
@@
 in simulation $dt = \eta$.
-% Actual simulation of the Hamiltonian requires taking discrete steps, and therefore introduces some error.
-The error can be corrected using a MH step, as shown in Algorithm 1 (showing leap-frog algorithm).
-By only taking a single leap-frog step, HMC reduces to Langevin dynamics which can be represented as a gradient descent algorithm with injected noise.
-By using small step-sizes the acceptance rate approaches 1, so that no MH correction steps are necessary and the algorithm can be run to generate a stream of (possibly highly correlated) samples from the posterior.
-
-% Two problems must be addressed.
-The first is loglikelihood computation.
-For correctness, Bayesian inference algorithms require frequent computation of the (deterministic) loglikelihood over the entire observation set.
-Contrast this to optimization, where stochastic gradient techniques are necessary for learning parameters.
-The second problem is that massive parameter sets is synonymous with massive datasets.
-As datasets grow, so too do our models.
-This is particularly true in the area of deep learning, where parameter vectors start in the hundred thousands and for some models are in the billions.
-Hamiltonian Monte Carlo addresses this for Bayesian inference: by using gradients of the log-likelihood (log-prior) it is possible to avoid random walk behavior and more efficiently explore the posterior.
-
-% Recent advances in scaling Bayesian inference to big data has shown that both batches and gradients can be used to correctly sample from the posterior distribution.
-These use Hamiltonian dynamics based on stochastic gradients (where the stochasticity comes from approximating the true gradient with one based on a batch of data) with appropriate step-sizes (and variations, see below).
-We now describe these algorithms and then show how we can adopt them for ABC.
 
@@
 If the log-likelihood over the full data set is replaced with a mini-batch estimate, as is done for the following {\em stochastic gradient Hamiltonian dynamics} (SGHDs) algorithms, then the error in simulating the Hamiltonian dynamics comes not only from the discretization, but from the variance of the stochastic gradient.
-As long as this error is controlled, either by using small steps $\eta$ (SGLD), or adding friction terms $A$ (SGHMC), or using a thermostat $\xi$ (SGNHT), the expensive MH correction step can be avoided and values of $\thetav$ from the Hamiltonian dynamics can be used as approximate samples from the posterior.
+As long as this error is controlled, either by using small steps $\eta$ (SGLD), or adding friction terms $B$ (SGHMC), or using a thermostat $\xi$ (SGNHT), the expensive MH correction step can be avoided and values of $\thetav$ from the Hamiltonian dynamics can be used as approximate samples from the posterior.
 
@@
 \subsection{Deterministic Representations of Simulations}
-Implicit in each simulation run $\x \simsim \pi(\x | \thetav)$ is a sequence if internally generated random numbers that are used to produce random draws from $\pi(\x | \thetav)$.
+Implicit in each simulation run $\x \simsim \pi(\x | \thetav)$ is a sequence of internally generated random numbers that are used to produce random draws from $\pi(\x | \thetav)$.
 These random numbers are important to HABC because we wish to control the stochasticity of the simulator when computing its gradient.
@@
 \subsection{From Finite Differences to Simultaneous Perturbations}
-\input{algo_fdsa_abc}
+%\input{algo_fdsa_abc}
 \input{algo_spsa_abc}
-Algorithm~\ref{algo:fdsa} shows the {\em finite difference stochastic approximation} (FDSA) \cite{kiefer1952stochastic} to $\nabla U(\thetav)$ as a function of random seeds $\omegav$.
+[TODO: no longer algo1] Algorithm~\ref{algo:fdsa} shows the {\em finite difference stochastic approximation} (FDSA) \cite{kiefer1952stochastic} to $\nabla U(\thetav)$ as a function of random seeds $\omegav$.
 Note we have deliberately shown the deterministic simulations ($f$) outside of $\lleps$ to emphasize its dependence on $\x$.
@@
 \vspace{-0.1in}
-\caption{\small{A view of a simulator in terms of common random numbers.
+\caption{\small{A view of a simulator using persistent random numbers.
 The horizontal line represents $\y$ and red shading $\pm 2\eps$.
@@
 
-\subsection{Common and Sticky Random Numbers}
-The usefulness of applying common random numbers (CRNS) in SPSA has been previously demonstrated \cite{kleinman1999simulation}.
+\subsection{Persistent Random Numbers}
+The usefulness of applying {\em persistent} random numbers (PRNs) in SPSA has been previously demonstrated \cite{kleinman1999simulation}.
 In that work, the same random numbers are used to simulate both sides of the optimization function within the SPSA gradient.
-This makes sense intuitively, as we would generally assume that the expected simulation function varies smoothly in $d \thetav$;  by using CRNs, this smoothness is easily exploited (see Figure~\ref{fig:exp-crns}).
-If we were to apply SPSA to Bayesian learning, then using CRNs in the gradient step would be analogous to using the same mini-batch for both sides of the computation.
+This makes sense intuitively, as we would generally assume that the expected simulation function varies smoothly in $d \thetav$;  by using PRNs, this smoothness is easily exploited (see Figure~\ref{fig:exp-crns}).
+If we were to apply SPSA to Bayesian learning, then using PRNs in the gradient step would be analogous to using the same mini-batch for both sides of the computation.
+In the case where the number of random numbers in unknown or is itself random, we can simply consider seeds of the random number generator instead of vectors of random numbers.
 
-In addition to using CRNs in simulations for each gradient computation, we have found that using {\em persistent} random {\em seeds} helps  HABC explore the parameter landscape more easily for some algorithms and problems.
+In addition to using PRNs in simulations for each gradient computation,  we have found that using PRNs helps  HABC explore the parameter landscape more easily for some algorithms and problems.
 Intuitively, for a gradient-based sampling algorithm, it means a particle can slide along a smooth Hamiltonian landscape because    the additive noise is suppressed.
@@
 %
-%
-% [REWRITE] At $\theta_{\text{MAP}}$ we estimated the gradient $\nabla \hat{U}(\theta)$ using finite differences for both the kernel-$\epsilon$ and synthetic-likelihood representation of $\pi_{\eps}(y | x )$.
-For $S=5$ and $S=50$, $10K$ repetitions of computing the gradients was performed; there histograms are shown in Figure~\ref{fig:exp-varg}.
-At $S=5$ both methods reveal a small bias (values X and Y for kernel-$\eps$ and SL, respectively).
-At $S=50$, the bias for kernel-$\eps$ is much smaller, but its variance remains high; for SL, the bias remains unchanged but the variance has decreased significantly (values X pm S and Y pm T).
-The bias in the SL representation comes from the Gaussian approximation $\pi(y|\theta)$ which we can compute analytically using the CLT as $\text{N}( 1/\theta, 1/N \theta^2 + \eps^2)$.
-% (We believe this bias is due to the normal approximation assuming a symmetric mode as the bias goes to zero at the mean of the true posterior).
-Note that this bias is directional signal only appears nears the mode.
-For the $S=5$ setting in our experiment both approaches contain similar bias.
-[REWRITE END]
 
@@
 We ran chains of length $50K$ for SL-MCMC, SGLD, SGHMC, and SGNHT versions of HABC using SL gradient estimates ($S=5$).
-A pseudo-marginal version of SL-MCMC was used.
+SL-MCMC refers to pseudo-marginal ABC-MCMC.
 We note that SGHMC gave results nearly identical to SGNHT, so are not shown do to space limitations.
-In one set of experiments, common random seeds were used for gradient computations only, and did not persist over time steps; these experiments are called {\em non-persistent}.
+In one set of experiments, the same random seeds were used for gradient computations but did not persist over time steps; these experiments are called {\em non-persistent}.
 In another set of runs, we resampled $\omega_s$ at each time step with probability $\gamma = 0.1$; these experiments are {\em persistent}.
@@
   \cline{2-5}
-% Algo & $1000$ & $10000$ & $50000$ \\ \hline \hline
-% SL-ABC & $0.099$ & $0.047$ & $0.045$ \\
-% SGLD & $0.082$ & $0.049$ & $0.048$ \\
-% SGNHT & $0.271$ & $0.162$ & $0.161$ \\
-% SGNHT* & $0.144$ & $0.056$ & $0.075$ \\ \hline
  & \multicolumn{2}{|c||}{Non-persistent} & \multicolumn{2}{c|}{Persistent} \\
@@
 \multicolumn{1}{|l|}{Algo} & $10K$ & $50K$ & $10K$ & $50K$ \\ \hline \hline 
-\multicolumn{1}{|l|}{SL-ABC} & $0.047$ & $0.045$ & $0.045$ & $0.045$ \\
+\multicolumn{1}{|l|}{SL-MCMC} & $0.047$ & $0.045$ & $0.045$ & $0.045$ \\
 \multicolumn{1}{|l|}{SGLD} & $0.049$ & $0.048$ & $0.048$ & $0.043$ \\
-%\multicolumn{1}{|l|}{SGHMC} & $0.170$ & $0.165$ & $0.052$ & $0.047$ \\
 \multicolumn{1}{|l|}{SGNHT} & $0.232$ & $0.239$ & $0.055$ & $0.051$ \\\hline 
@@
 \vspace{-0.2in}
-% \begin{table}[h]
-% \caption{Posterior inference for Exponential problem}
-% \label{tab:exp-prob-no-sticky}
-% \begin{center}
-% \begin{tabular}{l||c|c|c|c|c}
-% {\bf Algo}  & 100 & 1000 & 10K & 50K \\
-% \hline \\
-% SL-MCMC &  $0.23$ & $0.12$ & $0.053$ & $0.044$ \\
-% SGLD    &  $0.26$ & $0.099$ & $0.047$ & $0.045$ \\
-% SGHMC   &  $0.26$ & $0.099$ & $0.047$ & $0.045$ \\
-% SGNHT   &  $0.26$ & $0.099$ & $0.047$ & $0.045$ \\
-% \end{tabular}
-% \end{center}
-% \end{table}
-
-% DIR: exp-eta-0p005-dtheta-0p01-C10-S5-omega-0p01
-% SL-MCMC omega_rate 100%
-% 
-% times 10,   100,  1000,  5000, 10000, 20000, 30000, 40000, 50000
-% mean 0.6052864 ,  0.26126289,  0.09895101,  0.05295502,  0.04721529,0.04513989,  0.04622511,  0.04491134,  0.04539522
-% std 0.09348374,  0.046628  ,  0.01530586,  0.00861296,  0.00604954, 0.00239858,  0.00327312,  0.00303499,  0.00242383
-% 
-% omega_Rate 1%
-% mean 0.63282705,  0.23141512,  0.11508726,  0.07034678,  0.05338616, 0.05246243,  0.0538948 ,  0.0507813 ,  0.04435618
-% std 0.11183616,  0.05139663,  0.05390608,  0.03376141,  0.01785321, 0.0125916 ,  0.00662542,  0.01255384,  0.00603819
-
-% SGLD omega_rate 100%
-% array([ 0.66893503,  0.41631787,  0.22451057,  0.15184296,  0.14112572, 0.15204372,  0.14857288,  0.15116372,  0.15076312])
-% std(0)array([ 0.08087815,  0.11567097,  0.04555416,  0.03197215,  0.01855117, 0.01417793,  0.01341749,  0.01312386,  0.00855595])
-% SGLD omega_rate 1%
-% mean 0.75547928,  0.35519016,  0.17736532,  0.09150572,  0.06159863,0.0553024 ,  0.06191501,  0.05637656,  0.05472899
-% std 0.11641948,  0.09003194,  0.08414296,  0.03832745,  0.01937652, 0.02211211,  0.01263801,  0.00796597,  0.00836038
-% SGHMC omega 100%
-% array([ 0.59033585,  0.25581098,  0.24731947,  0.22616621,  0.22842994,
-%         0.22780039,  0.22743227,  0.2256994 ,  0.22623605])
-%
-% In [25]: tvd.std(0)
-% Out[25]:
-% array([ 0.1870709 ,  0.05443133,  0.04280439,  0.01782881,  0.01695731,
-%         0.01044903,  0.00970157,  0.0081758 ,  0.00759141])
-% omega 1%
-% array([ 0.51193673,  0.36801527,  0.16197407,  0.09853979,  0.0929902 ,
-%         0.0976414 ,  0.09743771,  0.09113688,  0.08687801])
-%
-% In [28]: tvd.std(0)
-% Out[28]:
-% array([ 0.17030828,  0.10124135,  0.0653025 ,  0.0180961 ,  0.01414473,
-%         0.01492984,  0.0169915 ,  0.01351012,  0.009596  ])
-% 
-% SGTHEMO
-% omega 100%
-% In [33]: tvd.mean(0)
-% Out[33]:
-% array([ 0.61253343,  0.2419202 ,  0.15161436,  0.16610309,  0.18284641,
-%         0.18875997,  0.19055003,  0.1928232 ,  0.19340706])
-%
-% In [34]: tvd.std(0)
-% Out[34]:
-% array([ 0.17654396,  0.04334123,  0.04052037,  0.01846459,  0.0212887 ,
-%         0.009026  ,  0.00640309,  0.00575303,  0.00415439])
-%
-
-% omega 1%
-% In [30]: tvd.std(0)
-% Out[30]:
-% array([ 0.17343684,  0.10098802,  0.06547166,  0.02232569,  0.01339252,
-%         0.01902682,  0.0181153 ,  0.01225494,  0.00760486])
-%
-% In [31]: tvd.mean(0)
-% Out[31]:
-% array([ 0.47453793,  0.33112199,  0.14890592,  0.07600464,  0.06685565,
-%         0.06371391,  0.06363715,  0.05882387,  0.05590603])
-
-% \subsection{Sticky Random Numbers}
-% Finally we compare posterior estimates using sticky CRNs.
-Using a flip rate of $1\%$, i.e.
-each $\omega_s$ has $1\%$ chance of changing randomly at each time-step.
-In Figure~\ref{exp-traces-x-sticky} we show the traces plots of $<x^{(s)>}$ for the different methods.
-While the Hamiltonian dynamics provide persistence in $\thetav$-space, the sticky random numbers do the same for the ABC statistics.
-The Table~\ref{tab:exp-tvd-sticky} we report the same {\em tvd} error using sticky numbers.
-[DISCUSS difference with no-sticky].
-The potential benefit of using persistent random seeds is shown for all methods, including ABC-MCMC.
-
-%\subsection{Visualizing Noise from Gradient versus Injected Noise}
 
@@
 We show that HABC can work well despite using SPSA  gradients.
- %
- % The first is low-dimensional ($D=6$), but exhibits chaotic behavior.
-Though low-dimensional,  For the second problem we apply the techniques of HABC to a Bayesian image classifier and autoencoder.
-The benefit of this problem is we can compare directly with the SGHD algorithms using the analytic gradients.
-This problem is has very high-dimension ($D=1568$), yet we are able to successfully apply SPSA-based gradients to HABC.
 
@@
 \subsection{Bayesian Logistic Regression}\label{sec:auto}
-We perform Bayesian inference on Bayesian inference on a logistic regression model using the digits $0$ and $1$ from MNIST.
+We perform Bayesian inference on a logistic regression model using the digits $0$ and $1$ from MNIST.
 Despite its simplicity, the model still represents a high-dimensional problem for HABC ($D=1568$).
-We first run stochastic gradient descent to find $\thetav_{\text{MAP}}$ using the true gradient.
+We first ran stochastic gradient descent to determine $\thetav_{\text{MAP}}$ using the true gradient.
 We then run HABC SGLD and SG-Thermostats starting  $\thetav_{\text{MAP}}$ to discover how well the algorithms explore the posterior.
@@
 %
-We use $n=100$ mini-batches and $R=10$ for SPSA.
+We use $n=100$ sized mini-batches and $R=10$ perturbations for SPSA.
 Figure~\ref{fig:blr} shows samples randomly projected onto 2 dimensions ($1000$ evenly sub-sampled from $10K$).
@@
 These could be matched by occasional gradient estimates (Osborne).
+\subsubsection*{Acknowledgements}
+We thank the anonymous reviewers for the many useful comments that improved this manuscript.
 

### 2015-06-09 uai-cr.tex
post camera ready

@@
 % For algorithms
-%\usepackage{algorithm}
-%\usepackage{algorithmic}
+\usepackage{algorithm}
+\usepackage{algorithmic}
 
@@
 % this with the following command.
-%\newcommand{\theHalgorithm}{\arabic{algorithm}}
+\newcommand{\theHalgorithm}{\arabic{algorithm}}
 
@@
 % For algorithms
-\usepackage{algorithm}
-\usepackage{algorithmic}
+%\usepackage{algorithm}
+%\usepackage{algorithmic}
 
+\pagenumbering{gobble}
 % As of 2011, we use the hyperref package to produce hyperlinks in the
@@
 \And
-{\bf Max Welling }\\
+{\bf Max Welling }\thanks{Donald Bren School of Information and Computer Sciences
+University of California, Irvine, and Canadian Institute for Advanced Research.} \\
 Informatics Institute\\
@@
 These are called {\em likelihood-free} models because, in contrast to probabilistic models, their likelihoods are either intractable to compute or must be approximated by simulations.
-To perform inference in likelihood-free models, a broad class of algorithms called Approximate Bayesian Computation \cite{beaumont2002approximate,marjoram2003markov,sisson2007sequential,sisson:2010,marin:2012,fan:2013} are employed.
+To perform inference in likelihood-free models, a broad class of algorithms called Approximate Bayesian Computation \cite{beaumont2002approximate,marjoram2003markov,sisson:2010,marin:2012} are employed.
 
@@
 We call these {\em stochastic gradient Hamiltonian dynamics} (SGHD) algorithms.
-SGHD are computationally efficient for two reasons.
+SGHD algorithms are computationally efficient for two reasons.
 First, they avoid computing the gradient of the log-likelihood over the entire data set, instead approximating it using small batches of data, i.e.
@@
 The first is due to the form of the ABC log-likelihood.
-As we show in Section~\ref{sec:abc}, using a conditional model for $\pi( \x | \thetav )$ provides an estimate of the ABC likelihood that is less sensitive to $\eps$ and therefore is more conducive to stochastic gradient computations.
+As we show in Section~\ref{sec:abc}, using a conditional model for $\pi( \x | \thetav )$ provides an estimate of the ABC likelihood that is less sensitive to $\epsvec$ and therefore is more conducive to stochastic gradient computations.
 The second difficulty is that for high-dimensional parameter spaces, computing the gradients naively (i.e.
@@
 In Section~\ref{sec:scaling} we review three approaches to stochastic gradient inference using Hamiltonian dynamics: SGLD, SGHMC, and SGNHT.
-We then introduce Hamiltonian ABC in Section~\ref{sec:habc}, where  we will show how to improve the stability of the gradient estimates by using CRNs and local density estimators of the simulator.
+We then introduce Hamiltonian ABC in Section~\ref{sec:habc}, where  we will show how to improve the stability of the gradient estimates by using PRNs and local density estimators of the simulator.
 Extensions to high-dimensional parameter spaces are also discussed.
@@
 It can be shown that if the statistics used in the likelihood function are sufficient, then these algorithms sample correctly from an approximation to the true posterior \cite{marin:2012}.
-  The simulator is treated as generator of random pseudo-observations, i.e.
+  The simulator is treated as a generator of random pseudo-observations, i.e.
 $\x \simsim \pi( \x | \thetav )$ is a draw from the simulator.
-Discrepancies between the simulator outputs $\x$ and the observations $\y$ are scaled by a closeness parameter $\eps$ and treated as likelihoods.
-This is the equivalent to putting an $\eps$-kernel around the observations, and using a Monte Carlo estimate of the likelihood using $S$ draws of $\x$:
+Discrepancies between the simulator outputs $\x$ and the observations $\y$ are scaled by a closeness parameter $\epsvec$ and treated as likelihoods.
+This is the equivalent to putting an $\epsvec$-kernel around the observations, and using a Monte Carlo estimate of the likelihood using $S$ draws of $\x$:
 \begin{equation}
@@
 
-In ABC Markov chain Monte Carlo (MCMC) \cite{marjoram2003markov,Wilkinson2013,sisson:2010} the Metropolis-Hastings (MH) proposal distribution is composed of the product of the proposal for the parameters $\thetav$ and the proposal for the simulator outputs:
+In ABC Markov chain Monte Carlo (MCMC) \cite{marjoram2003markov,Wilkinson2013} the Metropolis-Hastings (MH) proposal distribution is composed of the product of the proposal for the parameters $\thetav$ and the proposal for the simulator outputs:
 \begin{equation}
@@
 \begin{equation}
-\alpha = \min \lp 1, \frac{\pi\lp\thetavp\rp \sum_{s=1}^S \pi_{\epsvec}(\y | \xps )  q( \thetav | \thetavp )}{\pi\lp\thetav\rp \sum_{s=1}^S \pi_{\epsvec}(\y | \xs ) q( \thetavp | \thetav )} \rp \label{eq:abc_mh_acceptance_with_s}
+\alpha = \min \lp 1, \frac{\pi\lp\thetavp\rp \sum_{s=1}^S \pi_{\epsvec}(\y | \x^{(s)'} )  q( \thetav | \thetavp )}{\pi\lp\thetav\rp \sum_{s=1}^S \pi_{\epsvec}(\y | \x^{(s)} ) q( \thetavp | \thetav )} \rp \label{eq:abc_mh_acceptance_with_s}
 \end{equation}
@@
 Most importantly, these methods have combined the ability of HMC to explore high-dimensional parameter spaces with the computational efficiency of using stochastic gradients based on small mini-batches of the full dataset.
-After an overview of HMC, we will briefly describe stochastic gradient Hamiltonian dynamics (SGHDs), starting with using  Langevin dynamics \cite{welling2011bayesian}, then HMC with friction \cite{chen2014stochastic}, and finally HMC with thermostats \cite{ding2014bayesian}.
-We will then make the connection between SGHDs and HABC in Section~\ref{sec:habc}.
+After an overview of HMC, we will briefly describe stochastic gradient Hamiltonian dynamics (SGHD), starting with using  Langevin dynamics \cite{welling2011bayesian}, then HMC with friction \cite{chen2014stochastic}, and finally HMC with thermostats \cite{ding2014bayesian}.
+We will then make the connection between SGHD and HABC in Section~\ref{sec:habc}.
 
@@
 
-More formally, the Hamiltonian $H\lp\thetav, \rhov \rp = U(\thetav) + K(\rhov)$ is a function of the current potential energy $U(\thetav)$ and kinetic energy $K(\rhov) = \rhov^T M^{-1}\rhov/2$ ($M$ is a diagonal matrix of masses which for presentation are set to $1$).
+More formally, the Hamiltonian $H\lp\thetav, \rhov \rp = U(\thetav) + K(\rhov)$ is a function of the current potential energy $U(\thetav)$ and kinetic energy $K(\rhov) = \rhov^T \Mv^{-1}\rhov/2$ ($\Mv$ is a diagonal matrix of masses which for presentation are set to $1$).
 The potential energy is defined by the negative log joint density of the data and prior:
@@
 \subsection{Stochastic Gradient Hamiltonian Dynamics}\label{se:sghd}
-If the log-likelihood over the full data set is replaced with a mini-batch estimate, as is done for the following {\em stochastic gradient Hamiltonian dynamics} (SGHDs) algorithms, then the error in simulating the Hamiltonian dynamics comes not only from the discretization, but from the variance of the stochastic gradient.
+If the log-likelihood over the full data set is replaced with a mini-batch estimate, as is done for the following {\em stochastic gradient Hamiltonian dynamics} (SGHD) algorithms, then the error in simulating the Hamiltonian dynamics comes not only from the discretization, but from the variance of the stochastic gradient.
 As long as this error is controlled, either by using small steps $\eta$ (SGLD), or adding friction terms $B$ (SGHMC), or using a thermostat $\xi$ (SGNHT), the expensive MH correction step can be avoided and values of $\thetav$ from the Hamiltonian dynamics can be used as approximate samples from the posterior.
+SGHD algorithms belong to a larger class of {\em noisy Monte Carlo} methods  that target intractable likelihoods; see \cite{alquier2014noisy} for an extensive overview of noisy Monte Carlo.
+% For a general overview of speeding of MCMC for both large-scale data and for likelihood-free models, see .
 
-SGHDs replace the full potential energy and its gradient with a mini-batch approximation:
+We develop SGHD from the large-scale data case, where the intractability is due to computing the full potential energy and its gradient; it is approximated using mini-batches:
 \begin{eqnarray}
@@
 \end{eqnarray} 
-where $n$ is the mini-batch size, and $h_i$ are indices chosen randomly without replacement from $[1,N]$ (i.e. it defined a random mini-batch).  
+where $n$ is the mini-batch size, and $h_i$ are indices chosen randomly without replacement from $[1,N]$ (i.e.
+it defined a random mini-batch).
+In likelihood-free settings, the stochasticity of the potential energy due to the mini-batches is instead caused by simulation noise; further likelihood assumptions, such as a Gaussian model, add another layer of approximation to our posterior.
+Below we describe three SGHD algorithms, originally developed for large-scale data applications, but for which we will apply directly to likelihood-free inference using gradient approximations in Section~\ref{sec:habc}.
 %The main motivation for these is to avoid the expensive MH correction step while maximizing the benefit of Hamiltonian dynamics.
@@
 
-{\bf Stochastic gradient Langevin dynamics} \cite{welling2011bayesian} performs one full leap-frog step of HMC.   Starting with a half step for the momentum, the update for $\thetav$ is 
-\begin{eqnarray}
-  \rhov_t & \sim & \mathcal{N}(0,\eye_p) \\
-  \rhov_{t+\frac{1}{2}} & = & \rhov_t - \eta \nabla \hat{U}(\thetav_t) /2 \\
-  \thetav_{t+1} & = & \thetav_t + \eta \rhov_{t+\frac{1}{2}}
-\end{eqnarray}
-It is not necessary to include $\rhov$ in the updates since there is only one step:
+{\bf Stochastic gradient Langevin dynamics} (SGLD) \cite{welling2011bayesian} performs one full leap-frog step of HMC.
+In doing so, SGLD avoids explicitly computing updates for momenta $\rhov$; the update for $\thetav$ is
+% \begin{eqnarray}
+%   \rhov_t & \sim & \mathcal{N}(0,\eye_p) \\
+%   \rhov_{t+\frac{1}{2}} & = & \rhov_t - \eta \nabla \hat{U}(\thetav_t) /2 \\
+%   \thetav_{t+1} & = & \thetav_t + \eta \rhov_{t+\frac{1}{2}}
+% \end{eqnarray}
+%It is not necessary to include $\rhov$ in the updates since there is only one step:
 \begin{eqnarray}
@@
 \end{eqnarray}
-One of the potential drawbacks of SGLD is that the momentum term is {\em refreshed} for every update of the parameters, and since this means the parameter update only uses the current gradient approximation, it limits the benefits of using Hamiltonian dynamics.
+One of the potential drawbacks of SGLD is that the momentum term is {\em refreshed} (implicitly) for every update of the $\thetav$, and since this means the parameter update only uses the current gradient approximation, it limits the benefits of using Hamiltonian dynamics.
 On the other hand, this also prevents SGLD from accumulating errors in the Hamiltonian dynamics.
+SGLD has been applied to another intractable likelihood model, Gibbs random fields \cite{alquier2014noisy}, which closely resembles how SGLD is applied in this paper.
 
 {\bf Stochastic Gradient HMC} (SGHMC) \cite{chen2014stochastic}  avoids $\rhov$ refreshment altogether.
-By applying HMC directly using the stochastic approximation $\hat{U}$ and $\nabla \hat{U}$, which the authors call {\em naive SGHMC}, the variance of the gradient will introduce errors that left unaddressed will result in sampling from the incorrect target distribution.
-Under the assumption that $\nabla \hat{U}(\thetav) = \nabla U(\thetav) + \mathcal{N}\lp \zerov, \varv_{\thetav}\rp$, where $\varv_{\thetav}$ is the covariance of the gradient approximation, and  updates $\rhov_{t+1} = \rhov_{t} + \Delta \rhov_{t}$ and $\thetav_{t+1} = \thetav_t + \eta \rhov_{t+1}$, the change in momenta $\Delta \rhov$ from one full step is
- \begin{equation}
-   - \eta \lp \nabla U(\thetav) + \mathcal{N}\lp \zerov, \varv_{\thetav}\rp\rp  =  - \eta \nabla U(\thetav) + \mathcal{N}\lp \zerov, \eta^2 \varv_{\thetav}\rp  
- \end{equation}
-By adding a friction term $\Bv$ to $\Delta \rhov$ proportional to $\varv_{\thetav}$, the correction step can be avoided
- \begin{eqnarray}
-   \Delta \rhov & = & - \eta \Bv \rhov_t - \eta \nabla U(\thetav_t) + \mathcal{N}\lp \zerov, 2 \eta \Bv\rp 
- \end{eqnarray}
- where  $\Bv=\frac{1}{2}\eta \varv_{\thetav_t}$.
+SGHMC makes the assumption $\nabla \hat{U}(\thetav) = \nabla U(\thetav) + \mathcal{N}\lp \zerov, \varv_{\thetav}\rp$, where $\varv_{\thetav}$ is the covariance of the gradient approximation.
+To avoid a MH correction step at the end of a trajectory, a friction term $\Bv$  proportional to $\varv_{\thetav}$ is added to $\Delta \rhov$.
+In practice, since we can only approximate $\Bv$, a user defined friction term $\Cv$ is used.  In our experiments we compute an online estimate $\hat{\varv}$ and set $\Cv = c \eye_p + \hat{\varv}$.
+
+{\bf Stochastic Gradient thermostats} (SGNHT) \cite{ding2014bayesian} addresses the difficulty of estimating $\Bv$ by introducing a scalar variable $\xi$ who's addition to the Hamiltonian dynamics maintains the temperature of the system constant, i.e.
+it acts as a (Nose-Hoover) thermostat \cite{leimkuhler2009metropolis}.
+
+%In summary, the hyperparameters required for these algorithms are $\eta$ and $\Cv$ (for SGHMC and SGNHT only), and in practice, some way of estimating $\hat{\varv}$ for SGHMC.
+
+% {\bf Stochastic Gradient HMC} (SGHMC) \cite{chen2014stochastic} avoids $\rhov$ refreshment altogether.
+% % By applying HMC directly using the stochastic approximation $\hat{U}$ and $\nabla \hat{U}$, which the authors call {\em naive SGHMC}, the variance of the gradient will introduce errors that left unaddressed will result in sampling from the incorrect target distribution.
+% SGHMC assumes that $\nabla \hat{U}(\thetav) = \nabla U(\thetav) + \mathcal{N}\lp \zerov, \varv_{\thetav}\rp$, where $\varv_{\thetav}$ is the covariance of the gradient approximation.
+To avoid a MH correction step at the end of a trajectory, a friction term $\Bv$ to $\Delta \rhov$ proportional to $\varv_{\thetav}$ is added to $\Delta \rhov$.
+% In practice we can only approximate $\Bv$, a user defined friction term $\Cv$ is used (with $\Cv - \hat{\Bv}$ is semi-positive definite).
+Thus the updates used for $\Delta \rhov = - \eta \Cv \rhov_t - \eta \nabla \hat{U}(\thetav_t) + \mathcal{N}\lp \zerov, 2 \eta (\Cv-\hat{\Bv}) \rp $.
+%In our experiments we compute an online estimate $\hat{\varv}$ and set $\Cv = c \eye_p + \hat{\varv}$. 
+%
+%
+% , and  updates $\rhov_{t+1} = \rhov_{t} + \Delta \rhov_{t}$ and $\thetav_{t+1} = \thetav_t + \eta \rhov_{t+1}$, the change in momenta $\Delta \rhov$ from one full step is
+%  \begin{equation}
+%    - \eta \lp \nabla U(\thetav) + \mathcal{N}\lp \zerov, \varv_{\thetav}\rp\rp  =  - \eta \nabla U(\thetav) + \mathcal{N}\lp \zerov, \eta^2 \varv_{\thetav}\rp
+%  \end{equation}
+% By adding a friction term $\Bv$ to $\Delta \rhov$ proportional to $\varv_{\thetav}$, the correction step can be avoided
+%  \begin{eqnarray}
+%    \Delta \rhov & = & - \eta \Bv \rhov_t - \eta \nabla U(\thetav_t) + \mathcal{N}\lp \zerov, 2 \eta \Bv\rp
+%  \end{eqnarray}
+%  where  $\Bv=\frac{1}{2}\eta \varv_{\thetav_t}$.
 In practice, since we can only estimate $\Bv$ by some $\hat{\Bv}$ and can only compute $\hat{U}$, a user defined friction term $\Cv$ is used (with $\Cv - \hat{\Bv}$ is semi-positive definite).
 Thus the updates used for $\Delta \rhov$ for SGHMC:
- \begin{equation}
-   - \eta \Cv \rhov_t - \eta \nabla \hat{U}(\thetav_t) + \mathcal{N}\lp \zerov, 2 \eta (\Cv-\hat{\Bv}) \rp 
- \end{equation}
- In our experiments we compute an online estimate $\hat{\varv}$ and set $\Cv = c \eye_p + \hat{\varv}$.  
+%  \begin{equation}
+%    - \eta \Cv \rhov_t - \eta \nabla \hat{U}(\thetav_t) + \mathcal{N}\lp \zerov, 2 \eta (\Cv-\hat{\Bv}) \rp
+%  \end{equation}
+%  In our experiments we compute an online estimate $\hat{\varv}$ and set $\Cv = c \eye_p + \hat{\varv}$.
  %As mentioned above, the main benefit of using SGHMC over SGLD is the ability of avoiding refreshing $\rhov$ while not accumulating Hamiltonian errors.
-
- {\bf Stochastic Gradient thermostats} (SGNHT) \cite{ding2014bayesian} addresses the difficulty of estimating $\hat{\Bv}$ by introducing a scalar variable $\xi$ who's addition to the Hamiltonian dynamics maintains the temperature of the system constant, i.e.
+ %
+ % {\bf Stochastic Gradient thermostats} (SGNHT) \cite{ding2014bayesian} addresses the difficulty of estimating $\hat{\Bv}$ by introducing a scalar variable $\xi$ who's addition to the Hamiltonian dynamics maintains the temperature of the system constant, i.e.
 it acts as a (Nose-Hoover) thermostat \cite{leimkuhler2009metropolis}.
 The update equations remain simple: initialize $\xi = \Cv$ (or $c$), then for $t=1\ldots$
- \begin{eqnarray}
-   \rhov_{t+1} & = & \rhov_t - \eta \xi_t \rhov_t - \eta \nabla \hat{U}(\thetav_t) + \mathcal{N}\lp \zerov, 2 \eta_t \Cv \rp \\
-   \thetav_{t+1} & = & \thetav_t + \eta \rhov_{t+1} \\
-   \xi_{t+1} & = & \xi_t + \eta \lp  \rhov_{t+1}^T \rhov_{t+1} / D - 1 \rp
- \end{eqnarray}
- In summary, the hyperparameters required for these algorithms are $\eta$ and $\Cv$ (for SGHMC and SGNHT only), and in practice, some way of estimating $\hat{\varv}$ for SGHMC.
+ % \begin{eqnarray}
+ %   \rhov_{t+1} & = & \rhov_t - \eta \xi_t \rhov_t - \eta \nabla \hat{U}(\thetav_t) + \mathcal{N}\lp \zerov, 2 \eta_t \Cv \rp \\
+ %   \thetav_{t+1} & = & \thetav_t + \eta \rhov_{t+1} \\
+ %   \xi_{t+1} & = & \xi_t + \eta \lp  \rhov_{t+1}^T \rhov_{t+1} / D - 1 \rp
+ % \end{eqnarray}
 
@@
 With this in mind we leave the details of the Hamiltonian updates to previous work \cite{welling2011bayesian,chen2014stochastic,ding2014bayesian} and focus on the details of how stochastic gradients are computed in the likelihood-free setting.
+Note that in our implementation, we do not use a MH correction (except when switching seeds), though this can easily be added for any particular problem.
 
@@
 This is particularly important to HABC as our gradient approximations are proportional to differences in $\lleps(\thetav)$; if the variance of the stochastic gradients is too high, then we must choose a very small step-size $\eta$, eliminating the usefulness of HMC for ABC.
-Under the deterministic representation of $\x^{(s)}$, we can write the loglikelihood as
+Under the deterministic representation of $\x^{(s)}$, we can write the log-likelihood as
 \begin{eqnarray}
@@
 \begin{equation}
-\mu_{\thetav} = \frac{1}{S} \sum_s f( \thetav, \omega_s ) ~~~~~~~~~~~~~ \sigma^s_{\thetav} = \frac{1}{S-1} \sum_s ( \mu_{\thetav} - f( \thetav, \omega_s ) )^2
+\mu_{\thetav} = \frac{1}{S} \sum_s f( \thetav, \omega_s ) ~~~~~~ \sigma^s_{\thetav} = \frac{1}{S-1} \sum_s ( \mu_{\thetav} - f( \thetav, \omega_s ) )^2
 \end{equation}
@@
 \end{equation} 
-In Figure~\ref{fig:exp-varg}, as part of our demonstration of HABC, we compare the gradient approximations around the true $\thetav_{\text{MAP}}$ using SL versus kernel-$\epsvec$ Figure~\ref{fig:exp-varg} for a simple problem.
-We find that although, for this particular problem, SL has a small bias due to its Gaussian assumption, it has much smaller variance, an important property for HABC.
+In Figure~\ref{fig:exp-varg}, as part of our demonstration of HABC, we compare the gradient approximations around the true $\thetav_{\text{MAP}}$ using SL versus kernel-$\epsvec$ for a simple problem.
+Although there is a small bias using SL due to its Gaussian assumption, it has much smaller variance, convergence to this (biased) posterior should be stable.
+Further, \cite{roberts1996exponential} showed that convergence for SGHD type algorithms depends on the tails of the log-posterior, which suggests that despite its bias, the non-heavy tails of the Gaussian may allow SL to produce a more efficient Markov chain.
+
+% dependsit has been shown that convergence of Langevin-based samplers for heavy-tailed log-posteriors it much slower than non heavy-tailed distributions like the Gaussian to stay correctly on the  Hamiltonian of the biased target distribution, but convergence to Gaussian log-posteriors should be faster, compared to those with heavy tails .
+%
+% \cite{roberts1996exponential}
+% swithout because 1) using a approximation may speed convergence for Langevin-based samplers convergence to  using a Gaussian approximation improve convergence to the posterior, which for Langevin-based samplers, depends on the tails of the log-posterior , and 2) having a lower variance gradient may keep the Hamiltonian trajectory closer to the (biased) ABC posterior, which is fundamental issue for scaling SGHD samplers \cite{betancourt2015fundamental}.
 
@@
 \input{algo_spsa_abc}
-[TODO: no longer algo1] Algorithm~\ref{algo:fdsa} shows the {\em finite difference stochastic approximation} (FDSA) \cite{kiefer1952stochastic} to $\nabla U(\thetav)$ as a function of random seeds $\omegav$.
-Note we have deliberately shown the deterministic simulations ($f$) outside of $\lleps$ to emphasize its dependence on $\x$.
+If the dimension of $\thetav$ is small, then {\em finite difference stochastic approximation} (FDSA) \cite{kiefer1952stochastic} can be applied to $\nabla U(\thetav)$ (conditioned on random seeds $\omegav$).
+% Note we have deliberately shown the deterministic simulations ($f$) outside of $\lleps$ to emphasize its dependence on $\x$.  
 The number of simulations required for FDSA is $2 S D$, which may be acceptable for some small ABC problems.
-Our goal is to scale ABC to high-dimensions and for that we need an alternative stochastic approximation of $\nabla U(\thetav)$.
+Our main goal is to scale ABC to high-dimensions and for that we need an alternative stochastic approximation to $\nabla U(\thetav)$.
 
@@
 Let $L$ be the gradient-free function we wish to optimize.
-Each approximation randomly generates a {\em perturbation mask} (our name) $\Deltav$ of dimension $D=|\thetav|$ where entry $\Deltavd \sim 2 \text{Bernouilli}(1/2) - 1$.
+Each approximation randomly generates a {\em perturbation mask} (our name) $\Deltav$ of dimension $D$ where entry $\Deltavd \sim 2 \text{Bernouilli}(1/2) - 1$ (i.e.
+all entries randomly set to $\pm 1$).
 Then $L$ is evaluated at $\thetav+\dtheta \Deltav$ and $\thetav-\dtheta \Deltav$, giving the gradient approximation $\gradvest(\thetav) \approx \partial L(\thetav) / \partial \thetav$:
@@
 Variations of SPSA include {\em one-sided} SPSA \cite{spall2000adaptive} (we use what Spall calls 2SPSA) and an algorithm for estimating the Hessian based on the same principle as SPSA \cite{spall2005monte}.
-The one-sided version is attractive computationally, but for HABC, the updates for $\thetav$ require simulating two-sides anyway (once at $\thetav$, after an step, and once for the one-sided gradient), so using 2SPSA makes more sense.
+The one-sided version is attractive computationally, but for HABC, the updates for $\thetav$ require simulating two-sides anyway (once at $\thetav$, after a step is taken, and once for the one-sided gradient).
 SPSA has also been used within a procedure for maximum-likelihood estimation for hidden Markov models using ABC \cite{Ehrlich2013}.
@@
 \vspace{-0.1in}
-\caption{\small{A view of a simulator using persistent random numbers.
+\caption{\small{A view of a simulator using persistent random numbers; in other contexts, these are called common random numbers \cite{kleinman1999simulation}.
 The horizontal line represents $\y$ and red shading $\pm 2\eps$.
@@
 
-\begin{figure}[t]
-\vskip 0.2in
-\begin{center}
-\includegraphics[width=0.75\columnwidth]{./images/exp_varg_figure.pdf}
-\vspace{-0.1in}
-\caption{\small{Variance of gradient estimation using kernel-$\eps$ and SL for different values of $S\in\{5,50\}$ and fixed $\eps = 0.37$ (the same used in the other results).  
-%
-When $S=5$, the empirical estimates of  $\nabla \hat{U}(\thetav_{\text{MAP}})$ are $-12 \pm 147$ (kernel-$\eps$) and $-9.3 \pm 43$ (SL).
-When $S=50$ they are $-0.80 \pm 19$ (kernel-$\eps$) and $-7.3 \pm 4.9$ (SL).
-Note the large discrepancy in variance.
-Note the limit of $S \rightarrow \infty$,   $\nabla \hat{U}(\thetav_{\text{MAP}}) = -7.8$.
-The bias if SL gradients is due to its Gaussian approximation (smoothed by $\eps$) of $\pi(\x|\thetav)$, which is a heavy-tailed Gamma distribution (the sum of $N$ exponentials).}}
-\label{fig:exp-varg}
-\end{center}
-\vspace{-0.2in}
-\end{figure}
-
 \subsection{Persistent Random Numbers}
@@
 If we were to apply SPSA to Bayesian learning, then using PRNs in the gradient step would be analogous to using the same mini-batch for both sides of the computation.
-In the case where the number of random numbers in unknown or is itself random, we can simply consider seeds of the random number generator instead of vectors of random numbers.
+In the case where the number of random numbers is unknown or is itself random, we can simply consider seeds of the random number generator instead of vectors of random numbers.
 
@@
 This is very similar to using dependent random streams to drive MCMC \cite{Murray2012,Neal2012}, the main difference we believe is that we are using the Hamiltonian dynamics to drive proposals for $\thetav$ and using persistent seeds $\omegav$ to suppress simulation noise.
+The full benefits of suppressing the noise may be limited, however.
+Recent work has shown that scaling HMC for large data applications may be fundamentally limited \cite{betancourt2015fundamental}: noise from mini-batches causes biases in trajectories, which require either increasing mini-batch sizes (in our case, running more simulations) or decreasing the step size.
 
@@
 The probability of the proposal is $q(x^{'}, \omega^{'} | \thetav, \omega) = \pi(\omega^{'})\delta( \x^{'} - f( \thetav, \omega^{'}))$, where $\delta(a)$ is a delta function at $a=0$.
-Because the $q$ has this form, acceptance ratio simplifies:
+Because the $q$ has this form, the acceptance ratio simplifies:
+% \begin{equation}
+%    \frac{\pi_{\epsvec}(\y | \x^{'})\pi(\omega^{'}) \pi(\x^{'} | \thetav, \omega_{'}) }{\pi_{\epsvec}(\y|\x)\pi(\omega)\pi(\x | \thetav, \omega)} \frac{\pi(\omega)\delta( \x - f( \thetav, \omega))}{\pi(\omega^{'})\delta( \x^{'} - f( \thetav, \omega^{'}))} =  \frac{\pi_{\epsvec}(\y | \x^{'})}{\pi_{\epsvec}(\y | \x)}
+% % % \min\lp 1, \frac{\pi(\y,\x^{'})}{\pi(\y,\x)}  \rp
+% \end{equation}
 \begin{equation}
-   \frac{\pi_{\epsvec}(\y | \x^{'})\pi(\omega^{'}) \pi(\x^{'} | \thetav, \omega_{'}) }{\pi_{\epsvec}(\y|\x)\pi(\omega)\pi(\x | \thetav, \omega)} \frac{\pi(\omega)\delta( \x - f( \thetav, \omega))}{\pi(\omega^{'})\delta( \x^{'} - f( \thetav, \omega^{'}))} =  \frac{\pi_{\epsvec}(\y | \x^{'})}{\pi_{\epsvec}(\y | \x)}
+   \frac{\pi_{\epsvec}(\y | \x^{'})\pi(\omega^{'}) \pi(\x^{'} | \thetav, \omega_{'})\pi(\omega)\delta( \x - f( \thetav, \omega)) }{\pi_{\epsvec}(\y|\x)\pi(\omega)\pi(\x | \thetav, \omega)\pi(\omega^{'})\delta( \x^{'} - f( \thetav, \omega^{'}))}=  \frac{\pi_{\epsvec}(\y | \x^{'})}{\pi_{\epsvec}(\y | \x)}
 % % \min\lp 1, \frac{\pi(\y,\x^{'})}{\pi(\y,\x)}  \rp
@@
 What we propose is slightly different.
-By instead keeping the random seeds fixed, we can sample $\thetav$ using HABC and use $\omegav$ as CRNs within the gradient computation step and suppress gradient noise over time.
+By instead keeping the random seeds fixed, we can sample $\thetav$ using HABC and use $\omegav$ as PRNs within the gradient computation step and suppress gradient noise over time.
 In this way, random seeds carry over the same additive noise from one step to the next.
-% Let $\pi(\omega_s)$ be a uniform prior over random seeds.
-We propose $\omega_s^{'} \sim q(\omega_s^{'} | \omega_s) = \pi(\omega_s)$ (independent of the current seed).
-Pseudo-data is simulated deterministically $\x^{'(s)} = f( \thetav, \omega_s^{'})$.
-Our proposal distribution has probability $\pi(x^{'(s)} | \thetav, \omega_s^{'}) = \delta( x^{'(s)} - f( \thetav, \omega_s^{'}))$.
-The acceptance probability ratio reduces to a ratio of likelihoods [TODO]:
-% % \begin{equation}
-% %   \frac{\pi(\omega^{'})\pi(\y,\x^{'})\pi(\x^{'} | \thetav, \omega_{'})}{\pi(\omega)\pi(\y,\x)\pi(\x | \thetav, \omega)}  \rp
-% % % \min\lp 1, \frac{\pi(\y,\x^{'})}{\pi(\y,\x)}  \rp
-% % \end{equation}
+
 %
-% [TODO: TIE THIS IN] This implies that our simulation $\x^(s)$ changes automatically (as it is deterministic).
-We argue that this is a valid transition operator using the similar arguments that can be made for pseudo-marginal ABC-MCMC which using equation~\ref{eq:pm-proposal} proposes $\thetavp$ and $\{ \x^{'(s)}\}$ jointly.
-% This is equivalent to flipping all $\omegav$ {\em and} proposing $\thetavp$.
-In ABC-MCMC, one could instead propose $q(\x^{'(s)} | \thetav)$ (fixing $\thetav$) and still sample correctly (this is like rejection sampling $\x | \y, \thetav$ based on $\epsvec$).
-What we propose is slightly different.
-Again keeping $\thetav$ fixed, we propose new random seeds with some probability $\gamma$ from some uniform prior distribution over seeds.
-The Metropolis-Hastings ratio reduces to the ratio of likelihoods for observations $\y$ given the new deterministic simulation values (just like it would for ABC-MCMC).
-
-% Our updates for $\omegav$ are simple.
-At each time step $t$, each seed is refreshed to a new random value with probability $\gamma$, i.e.
-$\omega_s \leftarrow \omega_s$ with probability $\gamma$, and $\omega_s \leftarrow \mathcal{U}(1,W)$, where $W=10^8$ in our algorithms.
-By observing the simulation at different $\thetav$ but using same seeds, we can eliminate much of the simulator noise from the Markov chain.
-By treating $\omegav$ as part of the state of the Markov chain, we can smoothly move along noiseless outputs for several iterations and then propose new seeds.
-Though proposing new seeds produces a valid Metropolis-Hastings algorithm (keeping $\thetav$ fixed and changing $\omegav$), we found that using {\em sticky} random seeds works well.
-These seeds can be precomputed before the chain is run and can be seen as apart from the state of the Markov chain.
-[CONFUSING]
-
-%[We may want anti-thetic instead of CRNs, ie overrelax versus underrelax, etc.]
-\begin{figure}[t]
-\vskip 0.2in
-\begin{center}
-\includegraphics[width=0.27\columnwidth]{./images/exp-SL-MCMC-posterior-hist-omega-rate-100p0-chain0.pdf}
-\includegraphics[width=0.27\columnwidth]{./images/exp-SG-Langevin-posterior-hist-omega-rate-100p0-chain0.pdf}
-\includegraphics[width=0.27\columnwidth]{./images/exponential/exp3-SG-Thermostats-posterior-hist-omega-rate-100p0-chain0.pdf}
-
-\includegraphics[width=0.27\columnwidth]{./images/exponential/exp2-SL-MCMC-posterior-hist-omega-rate-0p1-chain1.pdf}
-\includegraphics[width=0.27\columnwidth]{./images/exponential/exp2-SG-Langevin-posterior-hist-omega-rate-0p1-chain0.pdf}
-\includegraphics[width=0.27\columnwidth]{./images/exponential/exp2-SG-Thermostats-posterior-hist-omega-rate-0p1-chain1.pdf}
-\vspace{-0.1in}
-%\includegraphics[width=0.45\columnwidth]{./images/exp-ABC-MCMC-posterior_hist.pdf}
-%\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
-%\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Langevin-posterior_hist.pdf}
-%\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
-\caption{\small{Posterior distributions for the demonstration problem.
-{\bf Top row:} No persistent seeds.
-{\bf Bottom row:} Persistent seeds with $\gamma=0.1$.
-Histograms of the posterior estimates are overlaid with the true posterior (dashed line).
-All algorithms (except for SG-Thermostats for non-persistent $\omegav$) give roughly the same posterior estimate.
-By adding persistent $\omegav$ SG-Thermostats achieved similar posteriors to the other algorithms.}}
-\label{fig:exp-posteriors}
-\end{center}
-\vspace{-0.1in}
-\end{figure} 
-%
-% \begin{figure}[t]
-% \vskip 0.2in
-% \begin{center}
-% \includegraphics[width=0.45\columnwidth]{./images/exp-SL-MCMC-posterior-hist-omega-rate-0p01-chain0.pdf}
-% \includegraphics[width=0.45\columnwidth]{./images/exp-SG-HMC-posterior-hist-omega-rate-0p01-chain0.pdf}
-% \includegraphics[width=0.45\columnwidth]{./images/exp-SG-Langevin-posterior-hist-omega-rate-0p01-chain0.pdf}
-% \includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior-hist-omega-rate-0p01-chain0.pdf}
-% %\includegraphics[width=0.45\columnwidth]{./images/exp-ABC-MCMC-posterior_hist.pdf}
-% %\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
-% %\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Langevin-posterior_hist.pdf}
-% %\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
-% \caption{\small{Using sticky numbers at rate $1\%$}}
-% \label{fig:exp-posteriors}
-% \end{center}
-% \vskip -0.2in
-% \end{figure}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\vspace{-0.1in}
 We use a simple $D=1$ problem to demonstrate HABC.
@@
 % \end{figure*}
-\vspace{-0.2in}
+% \vspace{-0.2in}
 
 \subsection{Bias and Variance of $\nabla \hat{U}(\thetav)$}
-\vspace{-0.1in}
+%\vspace{-0.1in}
 To test our assumption that the synthetic-likelihood model is better suited for HABC, we ran FDSA at the true $\theta_{\text{MAP}}$.
@@
 It is for this reason that we have chosen to use SL likelihoods for our gradient estimates, despite their small bias.
-As mentioned in Section~\ref{sec:habc-grads} it is possible that other likelihood models, such as KDE, might provide low bias and low variance gradient estimates.
+As mentioned in Section~\ref{sec:habc-grads} it is possible that other likelihood models, such as a kernel density estimate, might provide low bias and low variance gradient estimates.
 We leave this for future work.
 %
+\begin{figure}[t]
+\vskip 0.2in
+\begin{center}
+\includegraphics[width=0.75\columnwidth]{./images/exp_varg_figure.pdf}
+\vspace{-0.1in}
+\caption{\small{Variance of gradient estimation using kernel-$\eps$ and SL for different values of $S\in\{5,50\}$ and fixed $\eps = 0.37$ (the same used in the other results).  
+%
+When $S=5$, the empirical estimates of  $\nabla \hat{U}(\thetav_{\text{MAP}})$ are $-12 \pm 147$ (kernel-$\eps$) and $-9.3 \pm 43$ (SL).
+When $S=50$ they are $-0.80 \pm 19$ (kernel-$\eps$) and $-7.3 \pm 4.9$ (SL).
+Note the large discrepancy in variance.
+Note the limit of $S \rightarrow \infty$,   $\nabla \hat{U}(\thetav_{\text{MAP}}) = -7.8$.
+The bias if SL gradients is due to its Gaussian approximation (smoothed by $\eps$) of $\pi(\x|\thetav)$, which is a heavy-tailed Gamma distribution (the sum of $N$ exponentials).}}
+\label{fig:exp-varg}
+\end{center}
+\vspace{-0.2in}
+\end{figure}
+
+%[We may want anti-thetic instead of PRNs, ie overrelax versus underrelax, etc.]
+\begin{figure}[t]
+\vskip 0.2in
+\begin{center}
+\includegraphics[width=0.27\columnwidth]{./images/exp-SL-MCMC-posterior-hist-omega-rate-100p0-chain0.pdf}
+\includegraphics[width=0.27\columnwidth]{./images/exp-SG-Langevin-posterior-hist-omega-rate-100p0-chain0.pdf}
+\includegraphics[width=0.27\columnwidth]{./images/exponential/exp3-SG-Thermostats-posterior-hist-omega-rate-100p0-chain0.pdf}
+
+\includegraphics[width=0.27\columnwidth]{./images/exponential/exp2-SL-MCMC-posterior-hist-omega-rate-0p1-chain1.pdf}
+\includegraphics[width=0.27\columnwidth]{./images/exponential/exp2-SG-Langevin-posterior-hist-omega-rate-0p1-chain0.pdf}
+\includegraphics[width=0.27\columnwidth]{./images/exponential/exp2-SG-Thermostats-posterior-hist-omega-rate-0p1-chain1.pdf}
+\vspace{-0.1in}
+%\includegraphics[width=0.45\columnwidth]{./images/exp-ABC-MCMC-posterior_hist.pdf}
+%\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
+%\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Langevin-posterior_hist.pdf}
+%\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
+\caption{\small{Posterior distributions for the demonstration problem; columns left to right: SL-MCMC, SGLD (SG-Langevin), SGNHT (SG-Thermostats).
+{\bf Top row:} No persistent seeds.
+{\bf Bottom row:} Persistent seeds with $\gamma=0.1$.
+Histograms of the posterior estimates are overlaid with the true posterior (dashed line).
+All algorithms (except for SGNHT for non-persistent $\omegav$) give roughly the same posterior estimate.
+By adding persistent $\omegav$ SGNHT achieved similar posteriors to the other algorithms.}}
+\label{fig:exp-posteriors}
+\end{center}
+\vspace{-0.1in}
+\end{figure}
+
+%
+% \begin{figure*}[t]
+% %\vskip 0.2in
+% \setlength{\linewidth}{\textwidth}
+% \setlength{\hsize}{\textwidth}
+% \begin{center}
+% \includegraphics[width=0.65\columnwidth]{./images/exp-SL-MCMC-theta-timeseries-omega-rate-100p0-chain0.pdf}
+% \includegraphics[width=0.65\columnwidth]{./images/exp-SG-Langevin-theta-timeseries-omega-rate-100p0-chain0.pdf}
+% \includegraphics[width=0.65\columnwidth]{./images/exponential/exp3-SG-Thermostats-theta-timeseries-omega-rate-100p0-chain0.pdf}
+% \includegraphics[width=0.65\columnwidth]{./images/exponential/exp2-SL-MCMC-theta-timeseries-omega-rate-0p1-chain3.pdf}
+% \includegraphics[width=0.65\columnwidth]{./images/exponential/exp2-SG-Langevin-theta-timeseries-omega-rate-0p1-chain3.pdf}
+% \includegraphics[width=0.65\columnwidth]{./images/exponential/exp2-SG-Thermostats-theta-timeseries-omega-rate-0p1-chain3.pdf}
+% \vspace{-0.15in}
+% \caption{\small{Trajectories of the last $1000$ $\thetav$ samples for the demonstration problem.
+{\bf Top row:} Non-persistent random seeds.
+{\bf Bottom row:} Persistent random seeds with $\gamma = 0.1$.
+Each algorithm's parameters were optimized to minimize the total variational distance.
+With persistent seeds, each algorithm's random walk behavior is suppressed.
+Without persistent seeds, the optimal step-size $\eta$ for SGNHT (SG-Thermostats) is small, resulting in an under-dispersed estimate of the posterior; when the seeds are persistent, the gradients are more consistent, and the optimal step-size is larger and therefore there is larger injected noise.
+The resulting posteriors are shown in Figure~\ref{fig:exp-posteriors}.
+% %For SG-Thermostats, persistent $\omegav$ allows for larger $\eta$.Sticky  Explanation thermostats: with no %sticky, need to lower eta, so it shows smoother trajectory.
+With sticky, the gradients are more consistent, %allowing for a larger eta and therefore more noise injection.
+% }}
+% \label{fig:exp-theta-traces}
+% \end{center}
+% \vspace{-0.1in}
+% \end{figure*}
+
+\subsection{Posterior Inference using HABC}
+We ran chains of length $50K$ for SL-MCMC, SGLD, SGHMC, and SGNHT versions of HABC using SL gradient estimates ($S=5$).
+SL-MCMC refers to pseudo-marginal ABC-MCMC.
+We note that SGHMC gave results nearly identical to SGNHT, so are not shown due to space limitations.
+In one set of experiments, the same random seeds were used for gradient computations but did not persist over time steps; these experiments are called {\em non-persistent}.
+In another set of runs, we resampled $\omega_s$ at each time step with probability $\gamma = 0.1$; these experiments are {\em persistent}.
+In Figure~\ref{fig:exp-posteriors} we show the posterior distributions for these experiments; in Table~\ref{tab:exp-posterior} we report the {\em total variational distance} between the true posterior and the ABC posteriors using the first $10K$ samples and after $50K$ samples (averaged over $5$ chains).
+Of note is the poor approximation of SG-Thermostats when the seeds are not persistent.
+By adding persistent seeds, SGNHT gives similar posteriors to the other methods.
 
@@
 With persistent seeds, each algorithm's random walk behavior is suppressed.
-Without persistent seeds, the optimal step-size $\eta$ for SG-Thermostats is small, resulting in an under-dispersed estimate of the posterior; when the seeds are persistent, the gradients are more consistent, and the optimal step-size is larger and therefore there is larger injected noise.
+Without persistent seeds, the optimal step-size $\eta$ for SGNHT is small, resulting in an under-dispersed estimate of the posterior; when the seeds are persistent, the gradients are more consistent, and the optimal step-size is larger and therefore there is larger injected noise.
 The resulting posteriors are shown in Figure~\ref{fig:exp-posteriors}.
@@
 
-\subsection{Posterior Inference using HABC}
-We ran chains of length $50K$ for SL-MCMC, SGLD, SGHMC, and SGNHT versions of HABC using SL gradient estimates ($S=5$).
-SL-MCMC refers to pseudo-marginal ABC-MCMC.
-We note that SGHMC gave results nearly identical to SGNHT, so are not shown do to space limitations.
-In one set of experiments, the same random seeds were used for gradient computations but did not persist over time steps; these experiments are called {\em non-persistent}.
-In another set of runs, we resampled $\omega_s$ at each time step with probability $\gamma = 0.1$; these experiments are {\em persistent}.
-In Figure~\ref{fig:exp-posteriors} we show the posterior distributions for these experiments; in Table~\ref{tab:exp-posterior} we report the {\em total variational distance} between the true posterior and the ABC posteriors using the first $10K$ samples and after $50K$ samples (averaged over $5$ chains).
-Of note is the poor approximation of SG-Thermostats when the seeds are not persistent.
-By adding persistent seeds, SG-Thermostats gives similar posteriors to the other methods.
-
 In Figure~\ref{fig:exp-theta-traces} we show the trace plots of the last $1000$ samples from a single chain for each algorithm.
@@
 Is it because of the interaction between hyperparameters and stochastic gradients, or is this an artifact of this simple model?
-%the random seeds were changed at each time-step In Figure~\ref{fig:exp-theta-traces} we show the posterior trace plots for ABC-MCMC, SGLD, SGHMC, and SGNHT versions of HABC using SL gradient estimates ($S=5$).
-The behavior of the trajectories is quite different, though they all produce the same correct posterior estimates (Figure~\ref{fig:exp-posteriors}).
-In Table~\ref{tab:exp-posterior} we report the {\em total variational distance} between the true posterior and the ABC posteriors after different numbers of samples (repeated 5 times).
-
-% \begin{table}[h]
-% \caption{Posterior inference for Exponential problem}
-% \label{tab:exp-prob-no-sticky}
-% \begin{center}
-% \begin{tabular}{l||c|c|c|c|c}
-% {\bf Algo}  & 100 & 1000 & 10K & 50K \\
-% \hline \\
-% SL-MCMC &  $0.26$ & $0.099$ & $0.047$ & $0.045$ \\
-% SGLD    &  $0.42$ & $0.22$ & $0.14$ & $0.045$ \\
-% SGHMC   &  $0.26$ & $0.099$ & $0.047$ & $0.045$ \\
-% SGNHT   &  $0.26$ & $0.099$ & $0.047$ & $0.045$ \\
-% \end{tabular}
-% \end{center}
-% \end{table}
 
@@
 The results for SGHMC (not shown) and SGNHT are nearly identical.
-%All algorithms tend to benefit from persistent seeds, in particular SGHMC and SGNHT, the methods with long Hamiltonian trajectories.
 }
@@
 \end{table}
-\vspace{-0.2in}
 
@@
 We compare difference HABC algorithms with ABC-MCMC for the blowfly population problem.
-We use $\epsvec = \{ 1/2,1/2,1/2,1/2, 1/4,1/4,1/4,1/4,3/4,3/4 \}$ (slightly different $\epsvec$ from \cite{Meeds2014GpsUai}) and $S=10$ for all experiments.
+We use $\epsvec = \{ 1/2,1/2,1/2,1/2, 1/4,1/4,1/4,1/4,3/4,3/4 \}$ (slightly different $\epsvec$ from \cite{Meeds2014GpsUai}) and $S=10$ for all experiments (this means that there are $S$ random seeds).
 We use SPSA with $R=2$ using SL log-likelihoods for all HABC gradient estimates.
@@
 
-Figure~\ref{fig:bf-results} show the posterior distributions for three parameters for SL-MCMC, SGLD, and SG-Thermostats using non-persistent seeds (persistent seeds, not shown, produced very similar posteriors).
+Figure~\ref{fig:bf-results} show the posterior distributions for three parameters for SL-MCMC, SGLD, and SGNHT using non-persistent seeds (persistent seeds, not shown, produced very similar posteriors).
 In the second row we show the trajectories of two parameters, clearly showing the suppressed random walk behavior of SGLD and SG-Thermostats relative to ABC-MCMC.
@@
    \caption{\small{Blowfly posterior distributions (non-persistent seeds).
-{\bf Top row}:  Posteriors for three parameters for SL-MCMC (left set of three), SGLD (middle), and SG-Thermostats (right).
+{\bf Top row}:  Posteriors for three parameters for SL-MCMC (left set of three), SGLD (SG-Langevin) (middle), and SGNHT (SG-Thermostats) (right).
 {\bf Bottom row:} Last trajectories of the last 1000 samples for two parameters for the same algorithms.
@@
 
-% \begin{figure}[t]
-% \vskip 0.2in
-% \begin{center}
-% \includegraphics[width=0.9\columnwidth]{./images/blowfly/non-sticky-theta-hist-SL-MCMC.pdf}
-% \includegraphics[width=0.9\columnwidth]{./images/blowfly/non-sticky-theta-hist-SG-Langevin.pdf}
-% \includegraphics[width=0.9\columnwidth]{./images/blowfly/non-sticky-theta-hist-SG-Thermostats.pdf}
-% \caption{\small{}}
-% \label{fig:bf-posteriors-no-sticky}
-% \end{center}
-% \vskip -0.2in
-% \end{figure}
-
-% \begin{figure}[t]
-% \vskip 0.2in
-% \begin{center}
-% \includegraphics[width=0.9\columnwidth]{./images/blowfly/sticky-theta-hist-SL-MCMC.pdf}
-% \includegraphics[width=0.9\columnwidth]{./images/blowfly/sticky-theta-hist-SG-Langevin.pdf}
-% \includegraphics[width=0.9\columnwidth]{./images/blowfly/sticky-theta-hist-SG-Thermostats.pdf}
-% \caption{\small{}}
-% \label{fig:bf-posteriors-sticky}
-% \end{center}
-% \vskip -0.2in
-% \end{figure}
 %
-% \begin{figure}[t]
-% \vskip 0.2in
-% \begin{center}
-% \includegraphics[width=0.9\columnwidth]{./images/blowfly/non-sticky-theta-1d-SL-MCMC.pdf}
-% \includegraphics[width=0.9\columnwidth]{./images/blowfly/non-sticky-theta-1d-SG-Langevin.pdf}
-% %\includegraphics[width=0.9\columnwidth]{./images/blowfly/sticky-theta-1d-SG-Langevin.pdf}
-% \includegraphics[width=0.9\columnwidth]{./images/blowfly/non-sticky-theta-1d-SG-Thermostats.pdf}
-% %\includegraphics[width=0.9\columnwidth]{./images/blowfly/sticky-theta-1d-SG-Thermostats.pdf}
-% \caption{\small{}}
-% \label{fig:bf-one-d-theta}
-% \end{center}
-% \vskip -0.2in
-% \end{figure}
-
 \begin{figure}[t]
@@
 \caption{\small{Blowfly trajectories of two parameters over the last 1000 time-steps.
-{\bf Top:} SGLD and {\bf Bottom}: SG-Thermostats.
+{\bf Top:} SGLD and {\bf Bottom}: SGNHT (SG-Thermostats).
 Relative to SL-MCMC (not shown), the Hamiltonian dynamics clearly show persistent $\thetav$ trajectories.}}
@@
 
-\vspace{-0.1in}
+%\vspace{-0.1in}
 \subsection{Bayesian Logistic Regression}\label{sec:auto}
 We perform Bayesian inference on a logistic regression model using the digits $0$ and $1$ from MNIST.
-Despite its simplicity, the model still represents a high-dimensional problem for HABC ($D=1568$).
+Although not technically an ABC problem because we use the actual likelihoods, it still represents a high-dimensional problem ($D=1568$) and is therefore useful to evaluate the potential of SPSA-like gradients in actual HABC problems.
 We first ran stochastic gradient descent to determine $\thetav_{\text{MAP}}$ using the true gradient.
-We then run HABC SGLD and SG-Thermostats starting  $\thetav_{\text{MAP}}$ to discover how well the algorithms explore the posterior.
-We compare with SGLD and SG-Thermostats using the {\em true} gradients.
+We then ran SGLD and SGNHT starting  $\thetav_{\text{MAP}}$ to discover how well the algorithms explore the posterior.
+We examine how SGLD and SGNHT trajectories are affected by using SPSA instead of the {\em true} gradients.
 %
@@
 \begin{center}
-%\includegraphics[width=0.95\columnwidth]{./images/lr/random_proj.png}
-\includegraphics[width=0.9\columnwidth]{./images/lr/lr_trajectories.pdf}
+\includegraphics[width=0.9\columnwidth]{./images/lr/lr_trajectories.jpg}
 \vspace{-0.15in}
-\caption{\small{Bayesian logistic regression sampling trajectories randomly projected.  The yellow circle is the projected $\thetav_{\text{MAP}}$.}  
-%{\bf White circle:} $\thetav_{\text{MAP}}$}; {\bf Dark blue:} true SGLD; {\bf Red:} SPSA SGLD; {\bf %Green:} true SG-Thermostats; {\bf Cyan:} SPSA SG-Thermostats.
+\caption{\small{Bayesian logistic regression sampling trajectories randomly projected.  The yellow circle is the projected MAP of $\thetav$.}
 }
@@
 
-%  neural network using MNIST images as observations.
-We apply two inference algorithms: SGNHT and HABC.
-For SGNHT, we use the known gradient and for HABC we apply 2SPSA.
-After 1000 iterations we collect parameter vectors every 100 iterations for a total of 100 sets.
-(Do we use SGHMC to estimate Bhat in first 1000 iterations?).
-Use same batches at each iteration for SGNHT and HABC?
-%
-% results: convergence of training error (batches), convergence in test error using parameters collected (record nbr of forward or backward passes required), show mean and variance of filters (how?), compare with using the last parameter sample only (stochastic optimization).
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section{DISCUSSION} \label{sec:discussion}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% \begin{itemize}
-%   \item
-%   \item
-% \end{itemize}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\vspace{-0.1in}
+%\vspace{-0.1in}
 Hamiltonian ABC proposes a new set of algorithms for Bayesian inference of likelihood-free models.
 HABC builds  upon the connections between Hamiltonian Monte Carlo with stochastic gradients and well-established gradient approximations based on a minimal number of forward simulations, even in high-dimensions.
-We have performed some preliminary experiments showing the feasibility of running ABC on both small and large problems, and we hope that the door has been opened for exploration of larger simulation-based models using HABC.
+We have performed some preliminary experiments showing the feasibility of running HABC on both small and large problems, and we hope that the door has been opened for exploration of larger simulation-based models using HABC.
 
@@
 This is most likely to be the case for simulators with large additive noise and algorithms that benefit from long Hamiltonian trajectories (i.e.
-SGHMC and SG-Thermostats).
+SGHMC and SGNHT).
 We feel that new classes of ABC algorithms could develop from using persistent random seeds, not just gradient-based samplers but traditional ABC-MCMC.
@@
 
-% \begin{itemize}
-%   \item New set of ABC algorithms for efficient exploration of posterior distribution.
-%   \item Key: opens the door to high-dimensional ABC inference.
-%   \item [Test for more accurate gradients] For ABC, the gradients we compute are stochastic because we must approximate $\pi(\x | \thetav)$ with forward simulations.
-We can make the approximation arbitrarily good by running more simulations, but this is often unnecessary for sampling, because we can make use of the uncertainty in our current approximation to decide if our sample is correct or not (cite UAI2014?).
-%   \item Unbiased gradients with low noise.  SL bias might be intolerable; use KDE to get stable unbiased gradients?
-% \end{itemize}
-
 Expensive simulators are an important class of models that we do not address in this work.
@@
 We could similarly use HABC with  ABC surrogate models \cite{Meeds2014GpsUai,wilkinson:2014} to minimize simulation calls, yet still  benefit from Hamiltonian dynamics.
-%We address this scenario in Section~\ref{future}.  
-%Shown for relatively fast simulators.
-Extensions to surrogates with GPS (Meeds2014GpsUai, rasmussen:2003, wilkinson:2014) can use derivative of GP for likelihood as GP of gradient.
-These could be matched by occasional gradient estimates (Osborne).
+
 \subsubsection*{Acknowledgements}
-We thank the anonymous reviewers for the many useful comments that improved this manuscript.
+We thank the anonymous reviewers for the many useful comments that improved this manuscript.  MW acknowledges support from Facebook, Google, and Yahoo.
 

### 2015-06-29 uai-2015-poster.tex
poster template

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% From a template maintained at https://github.com/jamesrobertlloyd/cbl-tikz-poster
%
% Code near the top should be fairly standard and not need to be changed
%  - except for the document class
% Code lower down is more likely to be customised
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[portrait,a0b,final,a4resizeable]{include/a0poster}

\usepackage{multicol}
\usepackage{color}
\usepackage{morefloats}
\usepackage[pdftex]{graphicx}
\usepackage{rotating}
\usepackage{amsmath, amsthm, amssymb, bm}
\usepackage{array}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{hyperref}

\usepackage{include/picins}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric,arrows,chains,matrix,positioning,scopes,calc}
\tikzstyle{mybox} = [draw=white, rectangle]
\definecolor{darkblue}{rgb}{0,0.08,0.45}
\definecolor{blue}{rgb}{0,0,1}

\usepackage{dsfont}
\input{include/jlposter.tex}

%\def \figPath{../uai2014-GPSABC/}
\def \figPath{images/}

\input{cdefs}
\begin{document}
\begin{poster}

% Potentially add some space at the top of the poster
\vspace{0\baselineskip}

%%% Header
\begin{center}
\begin{pcolumn}{0.99}

\newcommand{\logowidth}{0.099\textwidth}  % width mauna decomp
%\newcommand{\logowidth}{0.099\textwidth}  % width mauna decomp

\pbox{0.99\textwidth}{}{linewidth=2mm,framearc=0.3,linecolor=camdarkblue,fillstyle=gradient,gradangle=0,gradbegin=white,gradend=white,gradmidpoint=1.0,framesep=1em}{
%
%%% Cambridge Logo
\begin{minipage}[c]{\logowidth}
  \begin{center}
    \includegraphics[width=14cm,clip=true,trim=0.5cm 1.5cm 0.5cm 1.5cm]{badges/University-of-Amsterdam-logo.png}
  \end{center}
\end{minipage}
%
%%% Title
\begin{minipage}[c][9cm][c]{0.76\textwidth}
  \begin{center}
    {\sffamily \VeryHuge \textbf{Hamiltonian ABC}}\\[10mm]
    {\huge\sffamily \Huge Edward Meeds and Robert Leenders and  Max Welling\\[7.5mm]
    %\texttt{\{ti242, dkd23, zoubin\}@cam.ac.uk}
    }
  \end{center}
\end{minipage}
%
%
% Harvard logo
\begin{minipage}[c]{\logowidth}
  \begin{flushright}
    \includegraphics[width=8cm]{badges/500px-University_of_Amsterdam_logo.png}
  \end{flushright}
\end{minipage}

}
\end{pcolumn}
\end{center}

%\vspace*{1.5cm}

\large

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Beginning of Document
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{multicols}{3}

\mysection{Introduction}

%\vspace{0.5in}

%\begin{minipage}[c]{0.17\columnwidth}
\begin{itemize}
	\item Scientists express hypotheses through complex and expensive simulators.
	\item Posterior distributions give insight into models for both understanding underlying phenomena and improving hypotheses.
	\item Approximate Bayesian Computation provides a Bayesian framework for posterior analysis, but is very inefficient.
  \item This work uses a surrogate of the simulator to speed-up ABC.
  \item Based on the \emph{Metropolis-Hastings Error}, our algorithms determine when a simulation is necessary and provides the user with a ``knob'' $\xi$ to control it.
\end{itemize}

%\vspace{0.5in}

\mysection{Approximate Bayesian Computation}
%\vspace{-2cm}
\begin{tabular}{cc}
%\begin{minipage}[c]{0.73\columnwidth}
\begin{minipage}[c]{0.57\columnwidth}
  % put description of blowfly here
  \begin{itemize}
    %\item Bayesian inference: $\p( \thetav | \y ) = \frac{\p(\thetav) \p( \y | \thetav)}{\int \p(\thetav) \p( \y | \thetav) d \thetav}$
  	\item ABC is a \emph{likelihood-free} method because $\p( \y | \thetav)$ is either not computable or very expensive: \item[] \begin{equation}
  	\p_\epsilon( \thetav | \y ) = \frac{\p(\thetav)}{\p(\y)} \int \pi_\epsilon(\y|\x) \p(\x|\thetav) d \x \nonumber
  \end{equation}
      %\end{itemize}

      \item Kernel functions $\pi_\epsilon(\y|\x)$ are proxies for the likelihood, based on draws $\x \overset{\simulator}{\sim} \pi( \x| \thetav )$ from simulator.
      \item Rejection sampling with $\epsilon$-tube kernel is very inefficient. 
      \item We use ABC MCMC, which approximates the likelihood by Monte Carlo approximation:
      \begin{equation}
       \p_\epsilon(\y|\thetapv) \approx \frac{1}{S} \sum_{s=1}^S \pi_{\epsilon}(\y | \x^{(s)}, \thetapv ) \nonumber
      \end{equation}
      We accept the proposed parameter $\thetapv$ with probability equal to $\alpha( \thetapv | \thetav )  = $
      \begin{equation}
      \min \lp 1,  \frac{\pi(\thetapv) \sum_s \pi_{\epsilon}(\y | \xps, \thetapv )q( \thetav | \thetapv) }{\pi(\thetav) \sum_s \pi_{\epsilon}(\y | \x^{(s)}, \thetav )q( \thetapv | \thetav) } \rp \nonumber
      \end{equation} 
  \end{itemize}
\end{minipage}
&
%\begin{minipage}[c]{0.39\columnwidth}
\begin{minipage}[c]{0.43\columnwidth}
  % put blowfly problem pic here
  \begin{centering}
  \begin{tabular}{c}
  \includegraphics[width=\columnwidth,clip=true,trim=2.0cm 1cm 1cm 2cm]{\figPath exponential_problem.eps} \\
  \begin{minipage}[c]{\columnwidth}
    \begin{tabular}{cc}
      \multicolumn{2}{c}{$\eps$ trades-off precision and mixing}  \\
      \includegraphics[width=0.45\columnwidth,clip=true,trim=1.5cm 1.0cm 1.5cm 1.0cm]{\figPath abc_mcmc_eps_0p5.png} &
      \includegraphics[width=0.45\columnwidth,clip=true,trim=1.5cm 1.0cm 1.5cm 1.0cm]{\figPath abc_mcmc_eps_0p01.png}\\
      $\eps=0.5$ & $\eps=0.01$
    \end{tabular}
  \end{minipage}\\

  %Convergence to observations: \\
  %\hspace{-0.0cm}\includegraphics[width=\columnwidth, clip, trim = 0cm 0cm 1cm 0.61cm]{\figPath blowfly_conv_per_sims.pdf} \\
  \end{tabular}
  \end{centering}
\end{minipage}

\end{tabular}

\mysection{The Synthetic Likelihood}
\begin{itemize}
  \item Introduced by Wood (2010), replace the Monte Carlo approximation with a Gaussian with estimators based on the  pseudo-data $\{\x_1,..,\x_S\}$ simulated at $\thetav$:
  \begin{align}
  	\muhattheta    &=  \frac{1}{S} \sum_s \x^{(s)} & 
  	\Sigmahattheta &=  \frac{1}{S-1} \sum_s \lp \x^{(s)} - \muhattheta \rp \lp \x^{(s)} - \muhattheta \rp^T \nonumber
  \end{align}
  \item With a Gaussian kernel:
  \begin{equation}
  \pi_{\eps}(\y|\x) = \Keps\lp \y, \x \rp = \frac{1}{(2\pi\eps)^{J/2}} e^{-\frac{1}{2\epsilon^2}\lp\x-\y\rp^T \lp\x-\y\rp }\nonumber
  \end{equation}
  \item The synthetic likelihood can then be computed analytically:
  \begin{align}
  	\pi(\y|\thetav) &= \int \Keps\lp \y, \x \rp \mathcal{N}\lp \muhattheta,\Sigmahattheta\rp d\x \nonumber 
  	= \mathcal{N}\lp \muhattheta,\Sigmahattheta + \epsilon^2\eye \rp \nonumber
  \end{align}
\end{itemize}

\newpage 
\mysection{MCMC with a Random Acceptance Probability}

\begin{tabular}{cc}
\begin{minipage}[c]{0.65\columnwidth}
  \begin{itemize}
    \item The randomness of $\muhattheta$ (by re-running $S$ simulations), induces a distribution over the acceptance probabilities.
    \item Approximate randomness in $\mutheta$ by drawing $M$ times $\muthetam \sim \mathcal{N}\lp \muhattheta,\Sigmahattheta/S\rp$
    \item Estimate $p(\alpha)$ using Monte Carlo approximation based on these $M$ samples: 
    % \begin{align}
    %   p(\muthetapm) &= \mathcal{N}\lp \muhattheta,\Sigmahattheta\rp \\
    %    \alpham  &=  \min \lp 1,  \frac{\pi(\thetapv)\mathcal{N}\lp \y | \muthetapm, \Sigmahatthetap + \epsilon^2\eye \rp q( \thetav | \thetapv) }{\pi(\thetav) \mathcal{N}\lp \y | \muthetam, \Sigmahattheta + \epsilon^2\eye \rp q( \thetapv | \thetav) } \rp
    % \end{align}
    \begin{equation}
    		\alpham  =  \min \lp 1,  \frac{\pi(\thetapv)\mathcal{N}\lp \y | \muthetapm, \Sigmahatthetap + \epsilon^2\eye \rp q( \thetav | \thetapv) }{\pi(\thetav) \mathcal{N}\lp \y | \muthetam, \Sigmahattheta + \epsilon^2\eye \rp q( \thetapv | \thetav) } \rp  \nonumber
    \end{equation}
  \end{itemize}
\end{minipage}
&
\begin{minipage}[c]{0.35\columnwidth}
\centering
\begin{tabular}{c}
  $p(\alpha )$ and $F(\alpha)$  \\ 
  \includegraphics[width=\columnwidth]{\figPath acceptance_distribution_with_fold.png}
\end{tabular}
\end{minipage}
\end{tabular}
\begin{itemize}
  \item Based on $p(\alpha)$, compute the probability of making a M-H Error.  This is the area under the folded CDF and is minimized at the median of $p(\alpha)$, denoted $\tau$. \\ \\
  \vspace{0.5cm}
  \begin{minipage}[c]{\columnwidth}
    \begin{tabular}{cc}
      {\bf Probability of acceptance error} & {\bf Probability of rejection error} \\
      $P( \alpha < u ) = \frac{1}{M} \sum_m \lb \alpham < u\rb$ & $P( \alpha > u ) = \frac{1}{M} \sum_m \lb \alpham \geq u\rb$  
    \end{tabular}
  \end{minipage}
   \vspace{0.5cm}
  \item Total {\em conditional} error: $\errorcond = \lb u \leq \tau\rb P( \alpha < u ) + \lb u > \tau\rb P( \alpha \geq u )$
  \item MHE: $\erroruncond = \int \errorcond \mathcal{U}(0,1) d u$ 
  \item The crux of the MH step is to run simulations while $\erroruncond > \xi$.
\end{itemize}

%\newcommand{\gpdrawbox}[1]{
%\setlength\fboxsep{0pt}
%\hspace{-0.36in} 
% \fbox{
% \includegraphics[width=0.23\columnwidth]{../figures/deep_draws/deep_gp_sample_layer_#1}
% }}

% As depth increases, there is usually only one direction we can move $\vx$ to change $\vy$.
%
% \vspace{0.5in}
% We also visualize a distribution warped by successive functions drawn from a \gpt{} prior:
% \vspace{0.5in}
%
% \centering
% \begin{tabular}{cccc}
% $p(\vx)$ & $p(\vf^{(1)}(\vx))$ & $p(\vf^{(1:4)}(\vx))$ &  $p(\vf^{(1:6)}(\vx))$ \\
% %\gpdrawbox{1} & \gpdrawbox{2} & \gpdrawbox{4} & \gpdrawbox{6} \\
% \end{tabular}
% As depth increases, the density concentrates along one-dimensional filaments.
% \vspace{0.3in}

\mysection{Gaussian Process Surrogate ABC}

\begin{itemize}
  \item Adaptive SL is wasteful for expensive simulations: all results are discarded.
  \item GPS-ABC follow directly from synthetic likelihood ABC with randomized acceptance.
  \item Gaussian processes provides uncertainty estimates of the marginal likelihood informing us of the need to conduct additional experiments in order to make confident accept/reject decisions.
  \item For each statistic $j$, the surrogate provides the following conditional predictive distribution of the expected value of statistic $j$ based on $N$ training points $\muthetaj \sim	\mathcal{N}\lp \muthetajbar, \sigmasqtheta \rp $
  \begin{equation}
    \muthetajbar 
  	=
    \kthetaN 
  	\lb
  	\KNN + \sigma^2_j \eye
  	\rb^{-1}
  	\X[:,j] \nonumber
  \end{equation}
  \begin{equation}
      \sigmasqtheta  =  
    \kthetatheta 
    -
    \kthetaN 
  	\lb
  	\KNN + \sigma^2_j \eye
  	\rb^{-1}
    \kthetaN \nonumber
  \end{equation}
	\item Adjusting $\xi$ affects precision and computation:
  % (yellow -- model uncertainty, red -- training points):
\end{itemize}
\begin{tabular}{ccc}
  \begin{minipage}[c]{0.3\columnwidth}
    \centering
    $\xi = 0.4$, $N=9$
    \\
    \includegraphics[clip, trim = 1cm 1cm 2cm 1cm, width=\columnwidth]{\figPath toy_gp_s0_5_e_0p40_T_10000_gp_zoom_n_9.png}
    \\
    {\small high model uncertainty.}
  \end{minipage}
  &
  \begin{minipage}[c]{0.3\columnwidth}
    \centering
    $\xi = 0.2$, $N=184$
    \\
    \includegraphics[clip, trim = 1cm 1cm 2cm 1cm, width=\columnwidth]{\figPath toy_gp_s0_20_e_0p20_T_10000_gp_zoom_n_184.png}
    \\
    {\small model uncertainty drops with $\xi$}
  \end{minipage}
  &
  \begin{minipage}[c]{0.3\columnwidth}
    \centering
    $\xi = 0.05$, $N=1297$
    \\
    \includegraphics[clip, trim = 1cm 1cm 2cm 1cm, width=\columnwidth]{\figPath toy_gp_s0_20_e_0p05_T_10000_gp_zoom_n_1297.png} 
    \\
    {\small high precision, many simulations}
  \end{minipage}
\end{tabular}
\begin{itemize}
  \item[]
  \item The key advantage of GPS-ABC is that with increasing frequency, we will not have to do any expensive simulations whatsoever during a MH step because the GP surrogate is sufficiently confident about the statistics' surface in that region of parameter space. 
\end{itemize}

\newpage
\mysection{Toy problem: Inferring parameters of an exponential distributions}
\begin{itemize}
	\item GPS-ABC learns the surface, eventually eliminating any new simulations:
\end{itemize}
\begin{tabular}{ccc}
  \begin{minipage}[c]{0.3\columnwidth}
    \centering
    \includegraphics[clip, trim = 0.75cm 0.5cm 1cm 0.9cm, width=\columnwidth]{\figPath exp_paper1berror_v_samples.pdf}
    \\
    Error per sample
  \end{minipage}
  &
  \begin{minipage}[c]{0.3\columnwidth}
    \centering
    \includegraphics[clip, trim = 0.75cm 0.5cm 1cm 0.9cm, width=\columnwidth]{\figPath exp_paper1berror_v_both.pdf}
    \\
    Error per simulation
  \end{minipage}
  &
  \begin{minipage}[c]{0.3\columnwidth}
    \centering
    \includegraphics[clip, trim = 0.75cm 0.5cm 1cm 0.9cm, width=\columnwidth]{\figPath exp_paper1btime_v_sims.pdf} 
    \\
    Simulations per sample
  \end{minipage}
\end{tabular}
%\newcommand{\spectrumpic}[1]{
%\hspace{-0.2in}
%\includegraphics[trim=5mm 0mm 4mm 6.4mm, clip, width=0.25\columnwidth]{../figures/spectrum/layer-#1}}

% \begin{tabular}{cc}
% \begin{minipage}[c]{0.73\columnwidth}
%
% %{\bf Theorem}
% %The Jacobian of a deep \gp{} with a product kernel is a product of independent Gaussian matrices, with each entry in each matrix being drawn independently.
%
% \begin{itemize}
% \item Jacobian of a deep GP is a product of independent Gaussian matrices.
% \item Singular values spectrum of Jacobian quantifies relative size of derivatives.
% \item As the net gets deeper, distribution of SVs becomes heavy-tailed, and the largest singular value dominates.
% %\item Eventually, there is only one direction we can move $\vx$, in order to change $\vy$.
% \end{itemize}
%
% %Normalized singular value spectrum of the Jacobian of a deep GP.    This implies that with high probability, there is only one effective degree of freedom in the representation being computed.  As depth increases, the distribution on singular values also becomes heavy-tailed.
%
% \end{minipage}
% &
% \begin{minipage}[c]{\columnwidth}
% \begin{centering}
% \begin{tabular}{ccc}
% 2 layer SV spectrum \\
% %\hspace{-0.16in} \spectrumpic{2} \\
% %4 layers &
% 6 layers SV spectrum \\
% %\hspace{-0.16in} \spectrumpic{6}
% %2 layers & 4 layers & 6 layers
% \end{tabular}
% \end{centering}
% \end{minipage}
% \end{tabular}

\mysection{Chaotic Ecological Systems}

\begin{tabular}{cc}
\begin{minipage}[c]{0.47\columnwidth}
%\begin{minipage}[c]{0.95\columnwidth}
  % put description of blowfly here
  \begin{itemize}
    \item Adult blowfly populations exhibit dynamic, sometimes chaotic, behavior for which several competing population models exist.  
    \item We use observational data and a simulation model from Wood (2010).
    %, based on their improvement upon previous population dynamics theory.
    \item Population dynamics are modeled using (discretized) differential equations that can produce chaotic behavior for some parameter settings.
    \item The population dynamics equation generates  $N_1, \ldots, N_T$:
    \begin{tabular}{rcl}
    	$N_{t+1}$    & $=$ & $P N_{t-\tau} \exp(-N_{t-\tau}/N_0) e_t$ \\
    	 &  & $ + N_t \exp(-\delta \epsilon_t)$
    \end{tabular}
    \item Injected noise: $e_t \sim  \mathcal{G}( 1/{\sigma_p^2},1/{\sigma_p^2})$,  $\epsilon_t  \sim  \mathcal{G}(1/{\sigma_d^2},1/{\sigma_d^2})$ 
    \item $\theta$: $\{ \log P, \log \delta, \log N_0, \log \sigma_d, \log \sigma_p, \tau\}$
    \item Statistics $\y$ ($J=10$): the log of the mean of all $25\%$ quantiles of $N/1000$, the mean of the $25\%$ quantiles of the first-order differences of $N/1000$, and the maximal peaks of smoothed $N$, with 2 different thresholds.
    \item Models: rejection sampling ($\eps=0.5$), SL ($S=10$), GPS-ABC ($\xi=0.3$) 
    \item Results: generated observations (top), posterior samples $\thetav$ (middle), convergence to $\ystar$ using posterior predictive $p(\y | \ystar )$ (bottom).
  \end{itemize}
\end{minipage}
&
%\begin{minipage}[c]{0.39\columnwidth}
\begin{minipage}[c]{0.47\columnwidth}
  % put blowfly problem pic here
  \begin{centering}
  \begin{tabular}{c}
  \hspace{-0.0cm}\includegraphics[width=\columnwidth, clip, trim = 0cm 0cm 0cm 0.1cm]{\figPath blowfly_problem.pdf} \\
  %Convergence to observations: \\
  \hspace{-0.0cm}\includegraphics[width=\columnwidth, clip, trim = 0cm 0cm 1cm 0.1cm]{\figPath blowfly_theta_given_y.pdf} \\
  \hspace{-0.0cm}\includegraphics[width=\columnwidth, clip, trim = 0cm 0cm 1cm 0.1cm]{\figPath blowfly_conv_per_sims.pdf} \\
  \end{tabular}
  \end{centering}
\end{minipage} %\\
%
% \begin{minipage}[c]{0.47\columnwidth}
%   % put blowfly problem pic here
%   \begin{centering}
%   \begin{tabular}{c}
%   \hspace{-0.0cm}\includegraphics[width=\columnwidth, clip, trim = 0cm 0cm 1cm 0.1cm]{\figPath blowfly_theta_given_y.pdf}
%   \end{tabular}
%   \end{centering}
% \end{minipage}
% &
% \begin{minipage}[c]{0.47\columnwidth}
%   % put blowfly problem pic here
%   \begin{centering}
%   \begin{tabular}{c}
%   \hspace{-0.0cm}\includegraphics[width=\columnwidth, clip, trim = 0cm 0cm 1cm 0.1cm]{\figPath blowfly_theta_given_y.pdf}
%   \end{tabular}
%   \end{centering}
%\end{minipage}

\end{tabular}

\mysection{Conclusions}
\begin{itemize}
  \item Adaptive ABC algorithms using Metropolis-Hastings Error controls the computational complexity of the inference
	\item GP surrogate models incorporate both model and pseudo-data uncertainty into MCMC
	\item Improvements: Hamiltonian dynamics on the GP surface for independent samples; covariance on the outputs; alternative surrogate models; acquisition functions.
\end{itemize}

% \raggedright
%
% \begin{tabular}{cc}
% \begin{minipage}[c]{0.8\columnwidth}
%
% Code at \texttt{github.com/duvenaud/deep-limits}
%
% Paper at \texttt{mlg.eng.cam.ac.uk/duvenaud/papers/verydeep.pdf}
%
% \end{minipage}
% &
% \begin{minipage}[c]{0.2\columnwidth}
% \begin{centering}
% %\includegraphics[width=\linewidth]{figures/qrcode-paper-link}
% \end{centering}
% \end{minipage}
% \end{tabular}
%
%
%
\end{multicols}
\end{poster}

\end{document}
